### 2021-11-14 multipoleSSB/Haah.tex
first pull from overleaf

\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry} 
\usepackage{mathtools}
\usepackage{xcolor}

\newcommand{\note}[1]{\textcolor{red}{#1}}
\renewcommand{\cal}{\mathcal}
\newcommand{\nn}{\nonumber\\}

\title{Multipole group for Haah's U(1) code}
\author{charlie}
\date{August 2021}

\begin{document}

\maketitle

\section{Maximal quadrupole group}

In 3d, the maximal quadrupole group is generated by the set
\begin{align}
\left\{T_i, R_{ij}, \cal{P}_0, \cal{P}_1^{I_1}, \cal{P}_2^{I_2},
\right\}
\end{align}
where $T_i$ and $R_{ij}$, with $i,j=1,2,3$, are the translation and rotations. $\cal{P}_0$ is the constant shift, $\cal{P}_1^{I_1}$ are dipole shifts with $I_1=1,2,3$, and $\cal{P}_2^{I_2}$ are quadrupole shifts with $I_2=1,\dots,6$.

\section{Haah group}

Now, the group consists of
\begin{align}
\left\{T_i, R_{(1,1,1)}, \cal{P}_0, \cal{P}_1^1, \cal{P}_1^2, \cal{P}_2^1, \cal{P}_2^2
\right\},
\end{align}
where $R_{(1,1,1)}$ is a rotation in the plane perpendicular to that vector. The included polynomial shift symmetries are 
\begin{align}
\cal{P}_1^1 &= x_1-x_2, \nn
\cal{P}_1^2 &= x_1+ x_2 - 2x_3, \nn
\cal{P}_2^1 &= (x_1 - x_2) (x_1 + x_2 - 2x_3)\nn
\cal{P}_2^2 &= (2x_1 - x_2 - x_3) (x_2 - x_3),
\end{align}
which is, of course, a subgroup of the maximal quadrupole group.

This \note{almost} contains the 2d maximal dipole group as a subgroup\dots

If we define new variables $x = (x_1 - x_2)/\sqrt{2}$, $y = (x_1 + x_2 - 2x_3)/\sqrt{6}$, and $z = (x_1 + x_2 + x_3)/\sqrt{3}$, then we can write the Haah group as 
\begin{align}
\cal{M_H} &= \left\{T_i, R_{z}, \cal{P}_0, \cal{P}_1^1, \cal{P}_1^2, \cal{P}_2^1, \cal{P}_2^2
\right\},\nn
\cal{P}_1^1 &= \sqrt{2} x , \nn
\cal{P}_1^2 &= \sqrt{6} y, \nn
\cal{P}_2^1 &= \sqrt{12} xy\nn
\cal{P}_2^2 &= (3\sqrt{2} x + \sqrt{6} y) (\sqrt{6} y - \sqrt{2} x)\nn
&= 6y^2 +2\sqrt{12} xy - 6x^2,
\end{align}
so that $y^2-x^2$ is included in the group, but $y^2+x^2$ is not.

% \bibliographystyle{unsrt}
% \bibliography{big}
\end{document}

### 2021-11-14 multipoleSSB/dipole1d.tex
first pull from overleaf

\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry} 
\usepackage{mathtools}
\usepackage{xcolor}
\usepackage{hyperref}

\newcommand{\note}[1]{\textcolor{red}{#1}}
\newcommand{\nn}{\nonumber\\}
\newcommand{\goesto}{\rightarrow}
\renewcommand{\int}{\text{int}}
\newcommand{\ex}[1]{\left\langle#1\right\rangle}

\title{1d Lattice with dipole symmetry}
\author{charlie}
\date{July 2021}

\begin{document}

\maketitle

\section{First Attempt (classical spins)}

\subsection{The model}

Consider a 1d lattice, with d.o.f. on sites and links. On the sites we have spins $\sigma$ and on the links we have spins $\tau$.
Let the Hamiltonian be
\begin{align}
H_0 &= H_\sigma + H_\tau \nn
H_\int &= H_\sigma + H_\tau + H_{\sigma\tau}\nn
H_\sigma &= \sum_x \left( m_\sigma^2 |\sigma_x|^2 + \lambda_\sigma|\sigma_x|^4 \right) + \sum_x g_\sigma  \left| \sigma_{x+1}\sigma_{x-1} - \sigma_x^2 \right|^2\nn
H_\tau &= \sum_x \left( m_\tau^2 |\tau_{x,x+1}|^2 + \lambda_\tau|\tau_{x,x+1}|^4 \right) + \sum_x g_\tau \left| \tau_{x+1,x} - \tau_{x,x+1}\right|^2 \nn
H_{\sigma\tau} &= \sum_x g_{\sigma\tau} \left| \sigma_x^*\sigma_{x+1} - \tau_{x,x+1} \right|^2.
\end{align}
This Hamiltonian is invariant under $\sigma_x \goesto e^{i(ax+b)}\sigma_x$, $\tau_{x,x+1} \goesto e^{ia}\tau_{x,x+1}$. Thus, $a$ parameterizes the dipole part of the multipole group and $b$ is the monopole part. Note that the interaction term is
\begin{align}
H_{\sigma\tau} &= \sum_x g_{\sigma\tau} \left| \sigma_x^*\sigma_{x+1} - \tau_{x,x+1} \right|^2\nn
&= \sum_x g_{\sigma\tau} \left(| \sigma_x^*\sigma_{x+1} |^2 - \sigma_x^*\sigma_{x+1}\tau_{x,x+1}^* - \sigma_x\sigma_{x+1}^*\tau_{x,x+1} + |\tau_{x,x+1} |^2 \right),
\end{align}
where the first and last terms are potential terms.

\subsection{So what?}

The $(m,\lambda)$ terms allow us to condense $\sigma$ or $\tau$ separately, so there are four phases. When $\sigma$ is condensed, write it as $\sigma_x = e^{i\theta_x}$, ignoring normalization. Similarly, write $\tau_x = e^{i\phi_x}$. In the uncondensed phase, there is an unbroken dipole symmetry. Now, assume all the $\lambda$ and $g$ are positive.

For $H_0$, there is a simple phase diagram, just depending on the signs of $m_\sigma$ and $m_\tau$. If $m_\sigma<0$, then the $\sigma$ condense, so that the $\theta$ are well-defined. This fully breaks the dipole symmetry, since $\theta_x \goesto \theta_x + ax+b$. The interesting phase is when $m_\tau<0<m_\sigma$. Then, $\phi_x$ is well defined byt $\theta_x$ is not, so that the ground state is invariant under $b$-type (monopole) transformations, but not $a$-type (dipole) transformation. \note{However, as you pointed out, this is somewhat trivial since the a-type transformations act as monopoles operators on the $\tau$ degrees of freedom}.

If we instead focus on $H_\int$, there are interactions between the $\sigma$ and the $\tau$. \note{Is this what you meant by the link dofs facilitating hopping of the site dofs?}

\section{Quantum rotors}

The intermediate case between classical spins and bosons is quantum rotors. These have the advantage that the system is exactly solvable. On sites, we have degrees of freedom with states $|\theta\rangle$, $0\le \theta<2\pi$. The operators are $\theta$ and $L = -i\frac{\partial}{\partial\theta}$. On links, let the operators be $\phi$ and $J = -i\frac{\partial}{\partial\phi}$. The symmetry operator is 
\begin{align}
\mathcal{O}(\xi) &= \prod_je^{i\xi(jL_j+J_{j,j+1})}
\end{align}
which acts by rotating all variables in the same direction.

Then
\begin{align}
H_b &= \sum_i \; A  L_i + B L_i^2 + C \cos \left( \theta_{i-1} - 2\theta_i+\theta_{i+1} \right)\nn
H_d &= \sum_i \; A' J_{i,i+1} + B' J_{i,i+1}^2 + C' \cos \left( \phi_{i-1,i}-\phi_{i,i+1} \right) \nn
H_{\int} &= g \sum_i \; \cos \left( \theta_{i}+\phi_{i,i+1} - \theta_{i+1} \right) \nn
&= \frac{g}{2}\sum_i\left( e^{i\left( \theta_{i}+\phi_{i,i+1} - \theta_{i+1} \right)} + e^{-i\left( \theta_{i}+\phi_{i,i+1} - \theta_{i+1} \right) } \right)
\end{align}
The ground states are either number eigenstates or angle eigenstates. The advantage here is that the Hamiltonian is exactly solvable. The disadvantage is that the physical picture of condensed link variables facilitating hopping is not present.

\subsection{Noninteracting phases}

The noninteracting phases ($g=0$) are somewhat uninteresting. This is because symmetry breaking for the site systems behaves like dipole symmetry breaking, while the link systems undergo monopole symmetry breaking. To see this, consider the four phases.

In the trivial phase, $C,C'<<A,B,A',B'$, all rotors are in number eigenstates. The state is gapped and the symmetry operator acts as an overall phase. If we let $C'>>A',B'$, then the link states will be angle eigenstates, with a dispersion quadratic in the first derivative of the angle. \note{Is this a monopole symmetry breaking phase or a dipole symmetry breaking phase?}

If we instead let $C>>A,B$, the site variables condense. Since we have the lattice second derivative in the cosine, the dispersion is now quadratic in the second derivative of the angle. However, nothing extra happens when we let both variable types condense. The takeaway is that the $g=0$ system should display two types of CMW-type theorems, but they should match either system independently.

\subsection{Interacting phases}

The fixed point of the fully unbroken phase has the Hamiltonian
\begin{align}
H_b &= \sum_i \; A  L_i + B L_i^2 \nn
H_d &= \sum_i \; A' J_{i,i+1} + B' J_{i,i+1}^2,
\end{align}
so that the ground state is $\left|\frac{A}{2B}\right\rangle$ on sites and $\left|\frac{A'}{2B'}\right\rangle$ on links, rounded to the nearest integer. This is a gapped phase, with the symmetry fully unbroken.

The interesting behavior is when the link variables are condensed. Then the effective Hamiltonian is 
\begin{align}
H_\text{eff} &= \sum_i\; AL_i+BL_i^2 + g'\cos \left( \theta_i-\theta_{i+1} \right) + C\cos \left( \theta_{i-1} -2\theta_i + \theta_{i+1} \right),
\end{align}
which appears to have a relevant first lattice derivative term. \note{Does this allow some sort of workaround for the CMW theorem for the site operators?}

\section{Try again with bosons}

\subsection{Dipole hopping}

\begin{align}
H_b &= \sum_i \; A  n_i + B n_i^2 + C \left[ b^{\dag}_{i-1} b_i^2 b^{\dag}_{i+1} + H.C. \right]\nn
H_d &= \sum_i \; A' n'_{i,i+1} + B' (n'_{i,i+1})^2 + C' \left[d^{\dag}_{i-1,i} d_{i,i+1} +H.C.\right]\nn
H_{\int} &= g \sum_i \; b^{\dag}_i b_{i+1} d^{\dag}_{i,i+1} + H.C. 
\end{align}

The symmetry operator is $U = \prod_je^{i(aj+b)n_j}e^{ian'_{j,j+1}}$. Analysis follows Ref.~\cite{Chen2018}. We can access the interesting transitions by leaving $A, B, A', B'$, and $g$ fixed, tuning $C$ and $C'$. For concreteness, let $\ex{b^\dag b}=\ex{d^\dag d}=1$. \note{What does this line look like in the phase diagram?}

The disordered phase just has well-defined occupation numbers for sites and links. The fully-ordered phase has coherent states on sites, so the symmetry is completely broken. In this phase it doesn't really matter what's happening on the links. \note{Is this correct?} The interesting phase would be the one where the link variables are condensed and the site variables are not. Then the effective Hamiltonian is
\begin{align}
H_\text{eff} &= \sum_i\; A  n_i + B n_i^2 + g' \left[b_i^\dag b_{i+1} +H.C.\right] + C \left[ b^{\dag}_{i-1} b_i^2 b^{\dag}_{i+1} + H.C. \right],
\end{align}
so clearly there is some sort of cool phase transition. \note{Is this Hamiltonian correct?}

In this intermediate phase, the monopole symmetry is preserved but the dipole symmetry is broken. This is almost trivially true in the link sector, but the statement is a little more interesting in the site sector.

\subsection{Ground states}

In the unbroken sector,
\begin{align}
|\Psi_0(n,m)\rangle &= \bigotimes_{i}|n_i\rangle |m_{ij} \rangle, \nn
U(a,b)|\Psi_0(n,m)\rangle &= e^{i(aj+b)n+iam} |\Psi_0(n,m) \rangle \note{?}
\end{align}
In the first intermediate sector
\begin{align}
|\Psi_1(\gamma,m) &= \bigotimes_{i}|\gamma_i\rangle |m_{ij} \rangle, \nn
&= 
\end{align}

\subsection{Correlation functions}

Let $C_{ij}^b = \ex{b^\dag_ib_j}$ and $C_{ij}^d = \ex{d^\dag_{i,i+1}d_{j,j+1}}$

In the unbroken sector
\begin{align}
C_{ij}^b &= 
\end{align}

\subsection{Dispersion relations}

\bibliographystyle{unsrt}
\bibliography{big}
\end{document}

### 2021-11-14 multipoleSSB/disperions.tex
first pull from overleaf

\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry} 
\usepackage{mathtools}
\usepackage{xcolor}

\newcommand{\note}[1]{\textcolor{red}{#1}}
\renewcommand{\cal}{\mathcal}
\newcommand{\nn}{\nonumber\\}

\title{Nontrivial dispersion relations}
\author{charlie}
\date{September 2021}

\begin{document}

\maketitle

\section{Symmetries and dispersions}

Monopole symmetry means the kinetic term cannot see constant shifts, so the invariant derivative has to be $D\phi = \partial_i \phi$. Then the kinetic term is $(\partial_i\phi)^2$, and the propagator is $k_ik_i = k^2$. In general, if there are no solutions to Eqns.~28-31 in Gromov's paper, then we have to use the maximal kinetic term $(\partial_{i_1}\dots \partial_{i_n})$, which is invariant under any transformation from $\mathcal{M}_\text{max}^{n-1}$. Then the kinetic term is $(\partial_{i_1}\dots \partial_{i_n} \phi)^2$, the propagator is $(k^2)^n$, and the (classical) critical dimension is $d=2n = 2(a+1)$, the result Gromov quotes from the Griffin paper.

Gromov points out that Eqns.~28-31 are overdetemined and it is not clear when they have a solution. However, they do sometimes have a solution. For example, consider the dipole group extended by transformations of the form $\delta \phi = \mu_{ij}x^ix^j$ where $\mu$ is traceless. Then the most-relevant kinetic term is $(\partial^2 \phi)^2$, so that the propagator is $(k^2)^2$. Clearly, the classical critical dimension is $d=4$, in contrast to Gromov's claim that it should be $d=6$ since we include some polynomials of degree $2$. Gromov's claim is incorrect because he is citing the Griffin paper, where they always assume the maximal multipole group.

Can we make any more general statements about the correct critical dimension? We can divide the cases into isotropic and anisotropic as follows.

\section{CMW theorems}

\subsection{Isotropic}

The non-maximal, isotropic dispersion relations with smallest degree are the two parts of the dipole dispersion,
\begin{align}
E_1 &\sim (k^2)^2, \qquad \qquad E_2 \sim (k_i k_j)^2, \quad i\ne j.
\end{align}
Note that $(k^2)^2 = (k_ik_j)^2$, where the latter is the dispersion for the maximal dipole group. Furthermore, $(d-1)E_1=dE_2$. However, the two dispersions are built from different invariant derivatives using Gromov's prescription (Eqn. 36) and the requirement that the kinetic term be rotationally invariant.
Although these are dipole dispersion relations, they each individually admit a reduced but still isotropic quadrupole symmetry. Since they are constant multiples of $(k^2)^2$, they both share their critical dimension with the maximal dipole dispersion.

\subsection{Anisotropic}

Now, what about the dispersion relation $E = k^4 + p^2$, where $k$ has $d-1$ components and $p$ is a scalar? The correlation is
\begin{align}
C &\sim \int \frac{d^{(d-1)}k \, dp}{k^4 + p^2} \nn
&\sim \int \frac{d^{(d-1)}k}{k^2},
\end{align}
which diverges at $d=3$, which does not match the CMW theorem for any maximal dispersion relation. We can explain this result by assuming the field varies linearly in the $p$ direction and quadratically in the $k$ directions. The energy density will be $\sim L_p^{-2}$ or $L_k^{-4}$, depending on which direction it is varying in. The total energy cost is
\begin{align}
E &\sim L_p^{-2} L_k^{(d-1)} L_p + L_k^{-4} L_k^{(d-1)} L_p\nn
&\sim L_k^{(d-1)} L_p^{-1} + L_k^{(d-5)} L_p
\end{align}
which is minimized when $L_p \sim L_k^2$. Then $E\sim L_k^{d-3}$, so there is no energy cost at $d=3$, agreeing with the correlation calculation.

We can generalize this by letting $k$ have $q$ components and $p$ have $r$ components, with $d=q+r$. The correlation is
\begin{align}
C &\sim \int \frac{d^qk \, d^r\!p}{k^4 + p^2}\nn
&\sim \int d^qk \int \frac{p^{(r-1)}dp}{k^4+p^2} \nn
&\sim \begin{cases} \int d^qk \log k^4, &r=2\\
    \int \frac{d^qk}{k^2}, & r=1,
\end{cases}
\end{align}
so that if $r=2$ the symmetry can never be broken, and if $r=1$ then $q=2$ is critical. Of course, if $r=0$ we have an isotropic quartic dispersion and $q=4$ is critical.

We can again appeal to the energy-scaling argument. This time, the total energy cost is
\begin{align}
E &\sim L_k^{q} L_p^{r-2} + L_k^{q-4} L_p^r,
\end{align}
so that $L_p \sim L_k^2$ as before, and $E\sim L_k^{q+2r-4}$. Combinations $(q,r)$ for which symmetry breaking cannot occur are
\begin{align}
&(0,0), (0,1), (0,2), \nn
&(1,0), (1,1), \nn
&(2,0), (2,1), \nn
&(3,0), \nn
&(4,0).
\end{align}
We see that the critical dimension $d=q+r$ can be 2, 3, or 4, depending on how many directions have quadratic or quartic dispersions.

\subsection{Haah}

In the Haah group, the three derivatives are
\begin{align}
D_1 &= \partial_1 + \partial_2 + \partial_3,\nn
D_2 &= \partial_1^2 + \partial_2^2 + \partial_3^2,\nn
D_3 &= \partial_1 \partial_2 + \partial_2 \partial_3 + \partial_3 \partial_1.
\end{align}
At this point, Gromov writes down the kinetic term as $g_{\alpha\beta} (D_\alpha \phi) (D_\beta \phi)$, and he also gauges it. Is there a unique way to turn this into a dispersion relation?

\end{document}

### 2021-11-14 multipoleSSB/manuscript.tex
first pull from overleaf

\documentclass[twocolumn, longbibliography]{revtex4-2}

\usepackage{amsmath}
\usepackage[dvipsnames]{xcolor}
\usepackage{hyperref}
\hypersetup{
	colorlinks=true,
	linkcolor=blue,   
	urlcolor=blue,
}
\usepackage{graphicx}
\usepackage{tikz}

\newcommand{\note}[1]{\textcolor{red}{#1}}
\newcommand{\outline}[1]{\textcolor{Plum}{#1}}
\newcommand{\todo}[1]{\textcolor{orange}{#1}}
\newcommand{\charlie}[1]{\textcolor{TealBlue}{#1}}
\newcommand{\RN}[1]{\textcolor{TealBlue}{#1}}
\newcommand{\op}[1]{\mathcal{O}^{(#1)}}
\newcommand{\nn}{\nonumber\\}
\newcommand{\goesto}{\rightarrow}
\renewcommand{\max}{\text{max}}
\newcommand{\eff}{\text{eff}}
\newcommand{\cl}{\text{cl}}
\newcommand{\q}{\text{q}}
\newcommand{\half}{\frac{1}{2}}

\begin{document}

\title{Spontaneous breaking of multipole symmetries}
\author{Charles Stahl}
\author{Rahul Nandkishore}
\affiliation{Department of Physics and Center for Theory of Quantum Matter, University of Colorado Boulder, Boulder CO 80309, USA}

\begin{abstract}
Multipole symmetries are of interest both as a window on fracton physics and as a crucial ingredient in realizing new universality classes for quantum dynamics. Here we address the question of whether and when multipole symmetries can be {\it spontaneously broken} in thermal equilibrium. We derive generalized Mermin-Wagner arguments for the total or partial breaking of multipolar symmetry groups and generalized Imry-Ma arguments for the robustness of such multipolar symmetry breaking to disorder. We present both general results and explicit examples. Our results should be directly applicable to quantum dynamics with multipolar symmetries and also provide a useful stepping stone to understanding the robustness of fracton phases to thermal fluctuations, quantum fluctuations, and disorder. 
\end{abstract}

\date{\today}

\maketitle

\section{Introduction}

Hamiltonians invariant under polynomial symmetry transformations conserve not only charge, but also various multipole moments of charge. Such `multipolar' symmetries are known to offer a robust route to ergodicity breaking \cite{PPN, KHN, Sala, Moudgalya, SLIOM, commutant}, and also to exotic universality classes of quantum dynamics \cite{Iaconis1, GLN, nonabelian, glorioso, MKH, Feldmeier, Iaconis2}. They are known to arise in `fracton' phases of quantum matter \cite{Chamon, Haah, VHF1, VHF2, NH, PretkoRadzihovsky}, the key dynamical properties of which are known to descend from conservation laws on multipole moments of charge \cite{Pretko1, BB,  Gromov2019}. They are also known to arise (in a prethermal sense) in various ultracold atom platforms \cite{KHN, Bakr, Aidelsburger}. There are thus multiple reasons for thinking about systems with multipolar symmetries. However, just because a symmetry is present in the {\it Hamiltonian} does not mean that it will be present in the {\it state}; there is always the possibility of spontaneous symmetry breaking (SSB).

For conventional symmetries, there exist general theorems which constrain the settings in which SSB can occur. In clean systems, the relevant theorem is due to Mermin and Wagner \cite{MerminWagner}, and involves the physics of (thermal or quantum) fluctuations of the Goldstone modes associated with SSB, whereas in disordered systems the key results are due to Imry and Ma \cite{ImryMa, Vojta2013}, and also Aizenman and Wehr \cite{Aizenman}, and involve the physics of order parameter deformation for local alignment with disorder. Multipolar symmetries, however, allow for a much richer pattern of possible symmetry breakings (including breaking some but not all of the multipolar symmetries), and the analogous theorems have not yet been derived, except in the special case of isotropic clean systems with total breaking of the symmetry \cite{Griffin2013Multi}.

In this work, we place generalized Mermin-Wagner and Imry-Ma constraints on the total and/or partial breaking of multipolar symmetries, in both clean and disordered systems. Along the way we also discuss the exotic Goldstone modes associated with total or partial SSB of multipolar symmetries. We will also provide explicit models of multipole symmetry breaking, to give intuition for these unusual forms of SSB.

Throughout, we consider only multipole groups where the underlying internal group is continuous and abelian. For concreteness, we will say it is $U(1)$. Multipole groups with a nonabelian underlying symmetry suffer a cascade effect where the dynamics in at least one direction must be trivial~\cite{nonabelian} and we shall not discuss them here. We note that specific examples of spontaneous symmetry breaking of multipolar symmetries have been discussed (sometimes in a dual language) in \cite{elastic1, elastic2, elastic3, elastic4, elastic5, FS1, FS2}. Our goal here is to place general constraints on when certain symmetry breaking phase transitions involving multipolar symmetries can occur.

This paper is organized as follows. In Section \ref{multipolegroup} we introduce the multipole group, how to build multipole-invariant field theories, and the generalized Mermin-Wagner argument for total breaking of the maximal multipole symmetry. In Section~\ref{sec:evade} we show how to evade the generalized Mermin-Wagner argument by considering phases in which multipole symmetries are partially broken.
We discuss generalized Imry-Ma arguments in the presence of quenched disorder in Section~\ref{sec:disord}. Finally, we consider an explicit lattice model that illustrates some of these ideas in Section~\ref{sec:example} before concluding with a discussion of open questions in \ref{sec:disc}.

\section{The multipole group}
\label{multipolegroup}

The multipole group is well-explained in Ref.~\cite{Gromov2019}. It generalizes abelian global symmetries such as $\phi(x) \goesto\phi (x) +c$ by allowing a shift by some set of polynomials, $\phi (x) \goesto \phi (x) + \lambda_\alpha P^\alpha(x)$. The variables $\lambda_\alpha$ are symmetry parameters, while $\alpha$ labels the set of polynomials $P^\alpha(x)$. These so-called polynomial shift symmetries~\cite{Griffin2015} all commute with each other. It is helpful to limit ourselves to homogeneous polynomials. We can label these as $P_a^{I_a}$, where $a$ is the degree of the polynomial and $I_a$ is an abstract index that runs over the polynomials of degree $a$.

The full structure of the multipole group comes into play when we also include spatial symmetries. For example,  translation in the $x_1$ direction will fail to commute with any polynomial shift where the polynomial is a function of $x_1$. Thus, if we want to consider a collection of polynomial shift symmetries, we must consider whether that collection closes under conjugation by translations and rotations. If it does not, we must either exclude the offending translations or rotations, or expand the set of polynomial shift symmetries. The result is a multipole symmetry group~\cite{Gromov2019}.

\subsection{Examples} \label{sub:examples}

Reference~\cite{Gromov2019} includes discussion of some multipole groups; we will review a few here. The simplest case is the maximal multipole group $\mathcal{M}^a_\text{max}$, which includes all shifts by polynomials of degree $a$ or less. Individual polynomials can be written as
\begin{align}
P_c^{I_c} = \mu^{I_c}_{i_1\dots i_c}x^{i_1}\dots x^{i_c}, \label{eqn:basis}
\end{align}
where each  matrix  $\mu^{I_c}_{i_1\dots i_c}$ is fully symmetric and $c\le a$. This group also includes all translations and rotations.

An example of a multipole group that contains all translations and rotations but is not the maximal multipole group is the group generated by shifts of the form
\begin{align}
\phi \goesto \phi + \lambda_0 P_0^0 + \lambda_{i} P^{i}_1 + \lambda_{I_2} P^{I_2}_2,
\end{align}
where the degree-0 polynomial is $P_0^0=1$. The other polynomials are
\begin{align}
P_1^{i} = x^i,\quad \quad P_2^{I_2} &= \mu^{I_2}_{ij} x^i x^j,
\end{align}
where $\mu^{I_2}$ are a basis for the rank-2 traceless symmetric $d\times d$ matrices. Let us call this group $\mathcal{M}^2_{\text{sym}}$. Recall that the maximal quadrupole group $\mathcal{M}^2_{\text{max}}$ is already only built from symmetric matrices $\mu$. The tracelessness condition thus only removes one polynomial from the set.

This set of polynomial shift symmetries is compatible with all rotations because no rotation will generate a traceful matrix from a traceless one. The set of symmetric matrices, seen as a representation of the group of rotations, decomposes into two independent irreducible representations. One of these is the set of traceless matrices while the other is the single matrix $\delta_{ij}$. In fact, the set of polynomial shift symmetries consisting of constant and linear shifts along with shifts of the form $\delta \phi \propto x^ix^i$ is also compatible with all rotations. We could call this group $\mathcal{M}^2_\text{tr}$.

There is one multipole group worth mentioning that does not include all rotations. This is the multipole group corresponding to Haah's U(1) code~\cite{Haah2017, BB, Gromov2019}. We will explain the correspondence in the next subsection. The group itself consists of all translations, a single rotation about the $(1,1,1)$ axis (on the cubic lattice), and shifts by five polynomials~\cite{Gromov2019}. These are
\begin{align}
P_0^0 &= 1, \nn
P_1^1 &= x_1 - x_2,\nn
P_1^2 &= x_1 + x_2 - 2x_3\nn
P_2^1 &= (x_1 - x_2) (x_1 + x_2 - 2x_3)\nn
P_2^2 &= (2x_1 - x_2 - x_3) (x_2 - x_3).
\end{align}
Although this looks complicated, we can simplify the presentation by first choosing a new spatial basis and then also new basis for the polynomials.

If we define new variables $x = (x_1 - x_2)/\sqrt{2}$, $y = (x_1 + x_2 - 2x_3)/\sqrt{6}$, and $z = (x_1 + x_2 + x_3)/\sqrt{3}$, then we can write the Haah group as all translations, a single rotation in the $x-y$ plane, and
\begin{align}
P_1^1 &= \sqrt{2} x , \nn
P_1^2 &= \sqrt{6} y, \nn
P_2^1 &= \sqrt{12} xy\nn
P_2^2 &= 6y^2 +2\sqrt{12} xy - 6x^2.
\end{align}
Finally, a basis change and redefinition for the polynomials allows us to write
\begin{align}
P_1^1 &= x , \nn
P_1^2 &= y, \nn
P_2^1 &= xy\nn
P_2^2 &= y^2 - x^2.
\end{align}
Note that no polynomials depend on $z$.

The Haah multipole group thus has a product structure, $\mathcal{M}_\text{Haah} = \mathbf{R} \times \mathcal{M}^2_\text{sym}$, where the $\mathbf{R}$ corresponds to translations along $z$. The second group $\mathcal{M}^2_\text{sym}$ is the submaximal quadrupole group in 2 dimensions, only containing quadrupole shifts corresponding to symmetric traceless tensors.

\subsection{Multipole field theories} \label{sub:field}

We can now consider building field theories that are invariant under various multipole groups. Reference~\cite{Gromov2019} describes the process in detail, so we will only sketch the outline. References~\cite{Griffin2015, Gromov2019} discuss quantum field theories where configurations are weighted by an action. We will also want to consider thermal field theories with Hamiltonian descriptions, so we will abstractly discuss ``kinetic terms" in this section.

The key step is to be able to find spatial derivatives $D$ that annihilate all the polynomials $P_a^{I_a}$ that we want to include in our multipole group. We say that $D$ is order-$s$ if it takes the form
\begin{align}
D = q^i\partial_i + q^{ij}\partial_1\partial_j + \cdots + q^{i_1\dots i_s}\partial_{i_1}\dots\partial_{i_s},
\end{align}
where each $q$ is a symmetric tensor. Although it is not generically possible~\cite{Gromov2019}, we can sometimes find a set of $D_\alpha$ ($\alpha$ is an abstract index) with $s\le a_{\text{max}}$, where $a_\text{max}$ is the highest degree of the polynomials. In this case, the effective field theory will be invariant under some non-maximal multipole symmetry.

Generically, we cannot find any $D$ with $s\le a_\text{max}$ such that $DP_a^{I_a} = 0$ for all the polynomials we want. In that case we can still write down a theory with derivatives 
\begin{align}
D = q^{i_1\dots i_n}\partial_{i_1}\dots \partial_{i_n},
\end{align}
with $n=a_\text{max}+1$. Then, the field theory will be invariant under the maximal multipole group of degree $a_\text{max}$.

Armed with the invariant derivatives $D_\alpha$ ($\alpha$ is another abstract index), we can start to construct a kinetic term. The most general term, at lowest order in derivatives, would be~\cite{Gromov2019}
\begin{align}
K[\phi(x)] &= g_{\alpha\beta} (D_\alpha \phi) (D_\beta \phi).
\end{align}
We will often write the Fourier transform of the kinetic term as $\phi_{-k} K_k \phi_k$.  Requiring some spatial symmetries restricts the choices of $g_{\alpha \beta}$. For the maximal multipole group, enforcing all rotation symmetries results in the kinetic term
\begin{align}
K_a[\phi(x)] &= (\partial_{i_1}\dots \partial_{i_{a+1}} \phi) (\partial_{i_1} \dots \partial_{i_{a+1}} \phi),
\end{align}
which is the kinetic terms studied in Ref.~\cite{Griffin2015}.

For $a>0$, $K_a$ can be split into multiple terms while still remaining rotationally invariant. For example, the kinetic term for $\mathcal{M}^1_\max$ is 
\begin{align}
K_1[\phi(x)] = g_1 (\partial^2\phi)^2 + g_2 \sum_{ij} (\partial_i \partial_j \phi)^2.
\end{align}
For the special cases of $g_2=0$ or $g_2=-g_1$, the symmetry group expands to $\mathcal{M}^2_\text{sym}$ or $\mathcal{M}^2_\text{tr}$, respectively. However, for generic $g_1,g_2$ the symmetry group is $\mathcal{M}^1_\max$. Similar statements can be made for larger $a$.

Finally, some of these multipole symmetries can be gauged to arrive at effective field theories for fracton phases. Of course, after gauging the theory will have gapless excitations due to the U(1) symmetry, but they can be Higgsed to arrive at a gapped phase. This is the sense in which the Haah group mentioned earlier corresponds to a field theory for Haah's code~\cite{BB}. See Ref.~\cite{Gromov2019} for the full story.

\subsection{Generalized Mermin-Wagner argument}

We are now ready to describe the generalized Mermin-Wagner argument for an arbitrary maximal multipole group. This already appears in Ref.~\cite{Griffin2015}, so we are simply reviewing it here in preparation for the more generic cases to follow. Here, we will discuss both classical and quantum systems.

Consider a clean classical system with a spontaneously broken maximal multipole symmetry of degree $a=n-1$.
Then the kinetic term is
\begin{align}
K[\phi(x)] = (\partial_{i_1} \dots \partial_{i_n} \phi)^2, \label{eqn:adisp}
\end{align} 
where $\phi$ is the order parameter field. 
Now assume we are in an ordered phase, but begin to nucleate domains of linear size $L$. Since the order parameter is continuous there will be a thick domain ``wall", in which the order parameter changes gradually from its value in the background phase to its value in the domain. The thickness will be of order $L$.
Equation~\ref{eqn:adisp} tells us the  energy cost of the field changing its value will be minimized if it varies as a polynomial with degree $a+1$. In that case the energy density will scale as $K_k \sim k^{2(a+1)} \sim L^{-2(a+1)}$. Integrating that energy density over a region of size $L$ gives a total energy cost to nucleating domains of $L^{d-2(a+1)}$.

In clean systems there is no energy gain from domain nucleation. When $d\le2(a+1)$, the energy cost is bounded, so the entropy gain favors domain creation. Thus, ordered phases are unstable and spontaneous symmetry breaking cannot occur for $d\le2(a+1)$.

We can reproduce this argument by considering correlation functions. As a warm-up, consider the standard monopole case. Assume there is a monopole U(1) symmetry that is spontaneously broken, leading to a Goldstone boson $\phi(x)$. The correlation function,
\begin{align}
C(x) = \left\langle \phi(x) \phi(0) \right\rangle = \int \frac{d^dk}{(2\pi)^d} \frac{e^{i k \cdot x}}{k^2}, 
\end{align}
diverges when $d\le 2$. Correlation functions should not diverge, so the interpretation is that the Goldstone boson fluctuates strongly enough that the field $\phi(x)$ cannot be well defined. In turn, this tells us the symmetry could not have been broken. This is the classical Mermin-Wagner argument. We should note that, since we only care about long-distance divergences, we don't need to include the complex exponential as long as we only look for divergence at small $k$. We will thus drop this dependence in future calculations.

When we consider a multipole symmetry, the dispersion changes to $K_k=k^{2(a+1)}$. Now the correlation functions will scale as 
\begin{align}
C \sim \int \frac{d^dk}{K_k} \sim \int \frac{d^dk}{k^{2(a+1)}}, \label{eqn:correl}
\end{align}
which diverges for $d\le2(a+1)$. As before, we interpret the divergence of the correlation function as a sign that the symmetry cannot be spontaneously broken. This allows us to reproduce our scaling argument that spontaneous symmetry breaking of the multipole group $\mathcal{M}^a_\max$ cannot occur for $d\le2(a+1)$.

If we look at the quantum (zero temperature) setting, the energy scaling no longer tells us where the critical dimension is. This is because even when there is no energy cost to forming domains, there is no entropy gain. Instead, quantum fluctuations must be the motivation for domain nucleation. To find the quantum critical dimension, we calculate the correlation function,
\begin{align}
C(x) = \left\langle \phi(x) \phi(0) \right\rangle &\sim \int \frac{d\omega d^dk}{\omega^2 + k^{2( a + 1 )}},
\end{align}
which now diverges at $d\le a+1$~\cite{Griffin2015}, where $d$ is the number of spatial dimensions. We have halved the critical dimension, so that spontaneous symmetry breaking cannot occur at $d\le a+1$. It is interesting to note that the dynamical critical exponent in these theories is $z = a+1$, so the classical and quantum critical dimensions are related by $d_\text{cl} = d_\text{q} + z$.

Furthermore, the structure of the quantum correlation function,
\begin{align}
C &\sim \int \frac{d\omega d^dk}{\omega^2 + K_k}\nn
&\sim \int \frac{d^dk}{\sqrt{K_k}},
\end{align}
suggests that, at least in some broad class of theories, the quantum critical dimension will be half the classical critical dimension. This is not true in general, though, as we will show for theories with quenched disorder.

\subsection{Non-maximal multipole group} \label{sub:nonmax}

We can also consider the fate of symmetry breaking for multipole groups other than the maximal multipole group. In general this will not match the critical dimension for the maximal group.

For concreteness, we will consider some examples. First, recall the group $\mathcal{M}^2_\text{sym}$ from Sec.~\ref{sub:examples}. The group contains polynomials of degree 2. However, since the polynomials are all traceless, the most relevant derivative is $D = \sum_i \partial^2_i$.  We can immediately see the dispersion is $(k^2)^2$, so that the critical dimension is $d_\cl = 4$ or $d_\q = 2$.

We can also consider an anisotropic multipole group. 
Let the conserved charges be the monopole moment and $d_2$ components of the dipole moment. Then the dispersion will be $k^4 + p^2$, where $k$ has $d_2$ components, $p$ has $d_1$ components, and $d = d_1+d_2$. Classically, the correlation function is
\begin{align}
C &\sim \int \frac{d^{d_2} k \, d^{d_1} p}{k^4 + p^2}\nn
&\sim \int x^{d_2/2 -1} p^{d_1-1} \frac{dx\, dp}{x^2 +p^2} \nn
&\sim \int q^{d_2/2 + d_1 - 2}\frac{dq}{q},
\end{align}
where $q^2 = x^2 + p^2 = k^4 + p^2$. 
If $d_1=2$ the symmetry can be broken for any $d_2>0$, and if $d_1=1$ then $d_2=2$  is critical. Of course, if $d_1=0$ we have an isotropic quartic dispersion and $d=d_2=4$ is critical.

We can recover this result in  the energy-scaling argument. Since the system is anisotropic, the domains will have different sizes in different directions. Consider forming a domain of linear size $L_p$ in the  direction with quadratic dispersion and $L_k$ in the quartic direction. When the gradient of the order parameter is in the quadratic direction, the order parameter has to change over a length $L_p$ so the energy density is $L_p^{-2}$. Similarly, when the order parameter is changing in the quartic direction its energy density is $L_k^{-4}$.

Both of these energy densities will be integrated over domain ``walls" with volume $L_k^{d_2} L_p^{d_1}$.
As a result, the total energy cost is
\begin{align}
E &\sim L_k^{d_2} L_p^{d_1-2} + L_k^{d_2-4} L_p^{d_1},
\end{align}
so that $L_p \sim L_k^2$ to make the terms match, and $E\sim L_p^{d_2/2+d_1-2}$. Recall that when $E$ does not increase for larger domains, the entropy gain will cause them to nucleate, destroying the ordered phase.
We see that the critical dimension $d= d_2 + d_1$ can be 2, 3, or 4, depending on how many directions have quadratic or quartic dispersions.

The quantum correlation function behaves as
\begin{align}
C &\sim \int \frac{d\omega d^{d_2} k \, d^{d_1} p}{\omega^2 + k^4 + p^2}\nn
&\sim \int k^{d_2 - 2} p^{ (d_1-1)} \frac{k\,dk\, dp}{\sqrt{k^4 + p^2}} \nn
&\sim \int x^{d_2/2-1} p^{d_1-1} \frac{dx\, dp}{\sqrt{x^2 + p^2}} \nn
&\sim \int q^{d_2/2 -1} q^{d_1 -1} dq,
\end{align}
where again $q^2 = x^2 + p^2 = k^4 + p^2$. This diverges when $d_2/2 + d_1 \le 1$.

The previous analysis suggests a procedure for finding the critical dimension in clean anisotropic systems for breaking the multipole symmetry to the trivial group. Sort each dimension by the degree of its dispersion relation, so that there are $d_n$ dimensions with dispersion $k^{2n}$. Define the effective dimension as 
\begin{align}
d_\eff = \sum_n \frac{d_n}{n}.
\end{align}
In a classical system, spontaneous symmetry breaking cannot occur if $d_\eff\le 2$, while in a quantum system the critical dimension is $d_\eff = 1$.

We should note that the statements in this section were framed in terms of the dispersion of the Goldstone modes rather than the structure of the symmetry group. It is always possible to find the dispersion given the multipole symmetry group, but it may require some basis changes (as in the Haah multipole group).

\section{Evading the generalized Mermin-Wagner argument} \label{sec:evade}

\subsection{Breaking to a subgroup} \label{sub:subgroup}

In the previous section we assumed we had an order parameter field that fully broke the multipole symmetry. Let us now consider the case where the order parameter breaks the symmetry from the original group $G$ to some subgroup $H$. For simplicity, we will let $G$ be the maximal multipole group $\mathcal{M}^{a}_\text{max}$ and $H$ be $\mathcal{M}^b_\text{max}$, with $b<a$.

The simplest case is when $a=1$ and $b=0$. In physical terms, we are taking a system that conserves total charge and total dipole moment, then condensing objects that carry a dipole moment but no total charge. We then recognize that these fundamental dipoles, which are of course the dipole charges of the monopole symmetry, are also the monopole charges of the dipole symmetry.  We should then expect symmetry breaking cannot occur for $d\le 2$ classically or $d\le 1$ quantum mechanically.

It is interesting to note that condensing an object that carries dipole charge must break rotation symmetry. In Sec.~\ref{sec:example} we show how to break the dipole symmetry while preserving a lattice rotation symmetry by considering multiple species of dipole. We are not aware of a way to condense a dipole while preserving continuous rotation symmetry.

In more generality, the order parameter for this pattern of symmetry breaking will be an object $\phi_{i_1\dots i_m}$, where $m=b+1$. Because $G$ is a maximal multipole group, we can still write the polynomials as in Eqn.~\ref{eqn:basis}. For our new order parameter, they act as
\begin{align}
\delta \phi_{i_1 \dots i_m} = \lambda_{I_c} \partial_{i_1} \dots \partial_{i_m} P_c^{I_c}, \label{eqn:partial}
\end{align}
so that $\phi_{i_1 \dots i_m}$ is invariant under $H=\mathcal{M}^b_\max$.

The kinetic term should take the form
\begin{align}
K[\phi(x)] &= (\partial_{i_{m+1}}\dots \partial_{i_n} \phi_{i_1 \dots i_m})^2,
\end{align}
which is invariant under shifts of the form of Eq.~\ref{eqn:partial}. The dispersion for the Goldstone modes is $K_k \sim k^{2(n-m)} = k^{2(a-b)}$. Following the correlation function calculation above, we find a classical critical dimension of $d = 2(a-b)$ and a quantum critical dimension of $d = a-b$. We recover the maximal generalized Mermin-Wagner argument if we let $b=-1$ denote the trivial group ($b=0$ denotes the monopole group).

\subsection{Mediated condensation} \label{sub:mediated}

We have seen that the generalized Mermin-Wagner argument provides critical dimensions for partial breaking that are lower than the critical dimensions for full symmetry breaking. In fact, we can use this observation to evade the generalized Mermin-Wagner theorem, completely breaking any multipole symmetry for $d>2$ classically, or $d>1$ quantum mechanically. This possibility has been pointed out in Refs.~\cite{Griffin2015Cascade, Griffin2015}, but we include a further interpretation in terms of what types of phase transitions are allowed in different dimensions. For simplicity, we will begin by considering the classical dipole-preserving theory with symmetry group $\mathcal{M}^2_\max$.

To be explicit, consider a monopole field $\phi(x)$ and a dipole field $\phi_i(x)$. Under the dipole symmetry they transform as 
\begin{align}
\phi(x) &\goesto \phi(x) + a_ix^i + b,\nn
\phi_i(x) &\goesto \phi_i(x) + a_i,
\end{align}
where we can more explicitly see that $\phi(x)$ transforms like a monopole field under the dipole part of the symmetry. The kinetic term will look like
\begin{align}
K[\phi,\phi_i] &= (\partial_i \partial_j \phi)^2 + (\partial_i \phi_j)^2,
\end{align}
so that $\phi$ has a $k^4$ dispersion and $\phi_i$ has a $k^2$ dispersion. We can also include an interaction term of the form $g(\partial_i \phi - \phi_i)^2$, which does not break the symmetry.  The $\phi_i$ field is well-defined whenever the dipole symmetry is broken. As long as $\phi_i$ is well-defined, integrating out fluctuations in $\phi_i$ endows the monopole field $\phi$ with a $k^2$ dispersion. The coupling $g$ is set by how far the theory sits from the dipole critical point.

With a $k^2$ dispersion for the monopole field, we now have a route to evade the (generalized) Mermin-Wagner argument and fully break the dipole symmetry in dimensions $d> 2$. First, partially break the symmetry so that $\phi_i$ is well defined, which is allowed for any $d>2$. Then, we can use the effective dispersion of the $\phi$ field to break the remaining part of the symmetry, also in any $d>2$. This procedure can be generalized, so that any multipole symmetry can be fully broken in $d>2$, provided we have enough symmetry-breaking order parameters. The discussion also translates to the quantum setting, so that the ultimate critical (space) dimension is always $d=1$.

The preceding procedure gives us insight into what the generalized Mermin-Wagner arguments were telling us. When we wrote down those arguments, we did not include the intermediate fields. Thus, the generalized Mermin-Wagner arguments tell us when a direct transition between two phases is or is not allowed. Consider again the dipole group $\mathcal{M}^2_\max$. The phase in which the symmetry is fully broken can exist in any dimension $d>2$ ($d>1$ for quantum systems), but there can only be direct phase transitions between this phase and the wholly symmetry unbroken phase in $d>4$ ($d>2$). In dimensions $2<d<4$ ($1<d<2$) the fully symmetry broken and the fully symmetry unbroken phase must be separated by an intermediate phase that breaks dipole but not monopole symmetry.

\section{Systems with quenched disorder} \label{sec:disord}

Let us now add some quenched disorder to our systems. In particular, we will consider disorder that explicitly breaks the symmetry locally but does not break the symmetry on average. Spatial disorder will also break translation and rotation symmetry, but again not on average. Of course, strong enough disorder can always destabilize the ordered phase, so will not consider that case. Weak disorder can discourage the ordered phase and raise the critical dimension at which spontaneous symmetry breaking is impossible. Theorems of this type originate with Imry and Ma~\cite{ImryMa} and were proved by Aizenman and Wehr~\cite{Aizenman}.

In a disordered classical system, the energy scaling argument gives a nice explanation for why the critical dimension changes. Consider the formation of a domain in an otherwise ordered phase fully breaking the maximal multipole group $\mathcal{M}^a_\max$. There are $\sim L^d$ disorder samples in the new domain. Since each sample is taken independently, the typical energy gain from forming the domain will be $\sim L^{d/2}$, by the central limit theorem. Comparing this to the cost of domain formation we calculated previously, the ordered phase will be unstable to domain nucleation when $d\le 4(a+1)$.

In clean systems, the transition from classical to quantum brought down the critical dimension because the argument depended on entropy. The Imry-Ma argument only appeals to energy considerations, so the quantum critical dimension in the presence of disorder remains the same as the classical critical dimension~\cite{Vojta2013}. In disordered quantum systems, spontaneous symmetry breaking is impossible for $d\le 4(a+1)$.

To reproduce this argument using a correlation function calculation~\cite{ImryMa} we need to calculate the response to disorder. Call the disorder field $\theta(x)$ and redefine $\phi(x)$ to be the variation away from the average value $\phi_0$. The relevant part of the Hamiltonian is 
\begin{align}
H &\sim \int d^dx \left[ \half K[ \phi(x)] - h \phi(x) \theta(x) \right] \nn
&\sim \int d^dk \left[ \half \phi_{-k} K_k \phi_k -h \phi_{-k} \theta_k \right] \nn
&\sim \int d^dk \left[ \half \phi_{-k}' K_k \phi_k' - \half h^2 \theta_{-k} K^{-1}_k \theta_k \right],
\end{align}
where $h$ sets the strength of the disorder, and we have written $\phi_k' = \phi_{-k} - h K^{-1}_k \theta_{-k}$. Then, from the expectation of $\phi_k$,
\begin{align}
\langle \phi_k \rangle &\sim \frac{1}{h} \frac{\delta Z}{\delta \theta_{-k} } \nn
&\sim h K_k^{-1} \theta_{k},
\end{align}
we can see that the disorder produces fluctuations mediated by the susceptibility.

We can then compute the correlation function for $\phi(x)$,
\begin{align}
\langle \phi(0) \phi(x) \rangle &\sim \int d^dk \, K_k^{-2}  h^2 \langle \theta_{-k} \theta_{k} \rangle e^{ik\cdot x},
\end{align}
where $h^2 \langle \theta_{-k} \theta_{k} \rangle e^{ik\cdot x}$ does not affect the divergence at small $k$, assuming the disorder is short range correlated. We can compare to Eqn.~\ref{eqn:correl} to see that in the presence of disorder that couples linearly to an order parameter fully breaking multipole symmetry, the critical dimension for a disordered system is twice the critical dimension for having direct full multipole symmetry breaking in a clean classical system.

We will now consider disorder that breaks the symmetry to a subgroup. For simplicity, let the symmetry of the system be the maximal dipole group and let the disorder break the dipole part of the group but not the monopole part. The kinetic term is $K[\phi(x)]= (\partial_i \partial_j \phi)^2$, while the coupling to disorder will be something like $(\partial_i\phi - \zeta_i)^2$, where $\zeta_i$ is the disorder field. 
Define the vector field $\xi_i = \partial_i\phi$, with kinetic term $K[\xi(x)]= (\partial_i \xi_j)^2$. In the ordered phase $\xi_i$ will have some constant value. It need not vanish because of the dipole symmetry. In the presence of weak disorder, it will want to follow the $\zeta_i$ field where possible. We can now follow the argument of the Imry-Ma theorem to say that the critical dimension will be $d=4$.

A more general case is a system with symmetry $G = \mathcal{M}_\max^a$ and disorder that breaks the symmetry to $\mathcal{M}_\max^b$ but preserves $G$ on average. In this system the critical dimension should be $d=4(a-b)$ for both classical and quantum systems. Again, let $b=-1$ denote the trivial group in order to recover our previous results.

As in the clean systems, we can have a sequence of phase transitions separated by intermediate phases in which the multipole symmetry is broken to a progressively smaller subgroup. With enough order parameter fields, any multipole symmetry can be fully broken in dimension $d>4$, regardless of which order parameter the disorder couples to. The same is true in quantum systems. Whether a certain phase transition is allowed or whether there needs to be an intermediate phase depends on the spatial dimension and the nature of the disorder.

\section{Explicit lattice example} \label{sec:example}

We now present a transparent quantum lattice example to illustrate the spontaneous breaking of multipolar symmetry. To this end, we take seriously the idea that ``dipole charges are monopole charges of a dipole symmetry," and include explicit secondary degrees of freedom that couple to dipole moments of the original degrees of freedom. This allows us to access phases corresponding to partial or complete symmetry breaking.

Consider a $d$-dimensional hypercubic lattice with sites labeled by $j$ and directions labeled by $\mu = 1,\dots,d$. Let there be bosonic quantum degrees of freedom $n_j$ on the sites and $n'_{j,j+\mu}$ on the edges. Thus, we can say there are $d$ species of edge-type bosons.

Define the symmetry operators
\begin{align}
\op{1}(\xi) &= \prod_je^{i\xi n_j}\nn
\op{2}_\mu(\xi) &= \prod_j e^{i\xi (j_\mu n_j)}e^{i\xi n'_{j,j+\mu}},
\end{align}
where $j_\mu$ is the $\mu$-th component of the site label $j$.
The first operator corresponds to conservation of the total number of site bosons, while the $\mu$-th component of the second operator corresponds to the sum of the $\mu$-th component of the total dipole moment of site bosons and the total number of the $\mu$-th type of edge bosons. These symmetries allow us to exchange a dipole of site bosons for an edge boson.

The most relevant Hamiltonian obeying these symmetries is  
\begin{align}
H_0  &= H_b + H_d + H_\text{int}\nn
H_b &= \sum_j \left[ \frac{U_b}{2}  n_j (n_j-1) - \mu_b \, n_j \right] \nn
&\qquad - \sum_{j,\mu,\nu} t_b \left[  b_j b^{\dag}_{j+\mu} b^{\dag}_{j+\nu} b_{j+\mu+\nu} + H.C. \right]\nn
H_d &= \sum_{j,\mu} \left[ \frac{U_d}{2} n'_{j,j+\mu} (n'_{j,j+\mu} - 1) -\mu \, n'_{j, j + \mu} \right]\nn
&\qquad  - \sum_{j,\mu,\nu} t_d \left[d^{\dag}_{j, j + \mu} d_{j + \nu, j+\mu + \nu} +H.C.\right]\nn
H_\text{int} &= \sum_{j,\mu} g\left[ b_j d_{j,j+\mu} b^\dag_{j+\mu} + H.C. \right],
\end{align}
where sums are taken over sites $j$ and lattice directions $\mu$ and $\nu$.  The operator $b^\dagger_j$ creates a boson on site $j$ while $d^\dagger_{j,j+\mu}$ creates an boson on the edge between $j$ and $j+\mu$. For simplicity, let us set 
\begin{align}
\frac{\mu_b}{ U_b} = \frac{\mu_d }{ U_d} = \frac{1}{2},
\end{align}
so there is a robust insulating phase at weak hopping $t_b, t_d$~\cite{Fisheretal}.

The Hamiltonian $H_b$ controls condensation of the site bosons. When $t_b$ is small, the site bosons are in number eigenstates while when $t_b$ is large the site bosons condense. Similarly, the edge bosons condense when $t_d$ is large. Note that all $d$ species of edge boson will condense equally, so that lattice rotation symmetry is not broken. The term $H_\text{int}$ is allowed by the symmetry because it simultaneously removes an edge boson and creates a dipole of site bosons. When the edge bosons are condensed, $H_\text{int}$ gives the site bosons an effective single particle hopping, and hence an effective $k^2$ dispersion, which would otherwise be disallowed by symmetry.

At sufficiently high dimension, this system will support three phases, which may be loosely described as the Mott insulator, the monopole superfluid where site bosons are condensed and the dipole superfluid where edge bosons (but not site bosons) are condensed. 
Note that the dipole superfluid does preserve lattice rotations.
In the monopole superfluid phase, the effective Hamiltonian contains bare creation and annihilation terms for the edge bosons, which automatically breaks number conservation of the edge bosons, and effectively induces a dipole superfluid as well. We can identify these phases with symmetry breaking patterns of the multipole group. In the trivial phase the group is unbroken. In the dipole superfluid, the group is broken to its monopole subgroup. The monopole superfluid then corresponds to the fully broken symmetry.

In the absence of disorder, neither symmetry breaking pattern can occur for $d\le 1$. There may still be interesting phases, such as a Luttinger liquid.
For $d>2$ dipole and monopole symmetry breaking may occur, so all three phases are intact. Furthermore, transitions between any two phases are allowed.
This leaves the intermediate case of $d=2$, in which all three phases can exist, but there can be no transitions between any symmetry-unbroken phase and the monopole superfluid. Instead, there must be an intermediate phase with dipole but not monopole superfluidity. Representative single-parameter phase diagrams are shown in Fig.~\ref{fig:phases}.

\begin{figure}
\centering

\begin{tikzpicture}[scale = .9]
\draw[<->] (0,0) node[left]{(a)}  -- (8,0) node[right]{$s$};
\draw (4,-.1) -- (4,.1);

\node [align=center] (A) at (2, -0.5) {unbroken};
\node [align=center] (C) at (6, -0.5) {fully\\broken};

\end{tikzpicture}
\begin{tikzpicture}[scale = .9]

\draw[<->] (0,0) node[left]{(b)}  -- (8,0) node[right]{$s$};
\draw (2.667,-.1) -- (2.667,.1);
\draw (5.333,-.1) -- (5.333,.1);

\node [align=center] (A) at (1.333, -0.5) {unbroken};
\node [align=center] (B) at (4, -0.5) {dipole\\broken};
\node [align=center] (C) at (6.667, -0.5) {fully\\broken};
\end{tikzpicture}

\caption{(a) For  $d>2$, there can be direct transitions between symmetry-unbroken and the fully-broken phase. Here $s$ is intended to represent a parameter along an arbitrary line in the phase diagram that crosses this transition. (b) At $d=2$, although the symmetry can be fully broken, the fully broken phase cannot touch the unbroken phase. Any line through the phase diagram that connects the two phases must pass though the dipole-broken phase. These phase diagrams are for translation invariant systems at zero temperature.}
\label{fig:phases}
\end{figure}

Let us now consider the same system, but with disorder.  The disorder Hamiltonian is
\begin{align}
H_\text{dis} &= \left[h_b\sum_j \sigma^*_jb_j + h_d \sum_{j,\mu} \sigma^*_{j,\mu}d_{j,j+\mu}\right] + H.C.,
\end{align}
where $h_b$ and $h_d$ control the magnitude of the disorder and each instance of $\sigma$ is a random phase. We will always consider $h_b,h_d<<1$.

At first, we will set $h_d=0$ but turn on $h_b$. This corresponds to the main considerations from Sec.~\ref{sec:disord}, where the disorder couples to the order parameter that fully breaks the symmetry. This raises the critical dimension for the superfluidity of the site bosons. The edge bosons may still enter a superfluid phase for $d>1$, but the site bosons may not condense with long range phase order until $d>4$. There cannot be a direct transition from the insulator to the monopole superfluid until $d>8$.  A phase akin to a Bose glass \cite{Fisheretal} of monopoles, where site bosons condense but without long-range order, can still exist in $1\le d<4$.

Now, let $h_d\ne 0$ while keeping $h_b=0$. The Imry-Ma argument for the edge bosons says that they should not be able to enter a superfluid phase when $d\le 4$. Once again, we can enter a Bose glass phase where $\langle d_{j, j+\mu} \rangle \ne 0$. In this case, the site bosons acquire a random single particle {\it hopping}. This may be expected to give rise to a Bose glass phase also for the site bosons. This implies that the monopoles cannot even undergo mediated condensation to superfluidity for $d\le 4$. For $d>4$, the dipoles may enter a superfluid phase, and the monopoles can enter a superfluid phase without mediation.

It is difficult to justify including including one disorder field but not the other. The standard reason to forbid disorder is by imposing translation invariance, but if we are already including one disorder field we are necessarily breaking translation symmetry. For $h_b,h_d\ne 0$, we can fully rely on the Imry-Ma argument. No symmetry breaking can occur for $d\le 4$, the monopole superfluid must be separated from any symmetry-unbroken phase by a dipole superfluid for $4<d\le 8$, and any symmetry-breaking transition can occur for $d>8$.

\section{Discussion} \label{sec:disc}

In this paper we analyzed the spontaneous symmetry breaking of various multipole groups, and discussed generalized Mermin-Wagner theorems, which (we argued) should be best understood as constraints on when a direct transition fully breaking a multipole group can occur, and when a symmetry unbroken phase and a (monopole) symmetry breaking phase must be separated by intermediate phases breaking higher multipole symmetries. We also considered multipole groups that are not the maximal multipole group, and the effect of quenched disorder. The disorder that we considered explicitly broke the symmetry, either fully or to a subgroup.

Of course, we could consider further combinations of effects. We could spontaneously break a symmetry from a group $G$ to a subgroup $H$ where either $G$ or $H$ are non-maximal. Or we could consider non-maximal groups with disorder. While the number of potential examples to consider is unlimited, they should all be analyzable using the ideas introduced herein.

We should be clear that the arguments in this paper are Mermin-Wagner or Imry-Ma arguments. These essentially amount to stability analyses about the Gaussian symmetry breaking fixed point. In principle, even when the Gaussian symmetry breaking fixed point is unstable, a non-trivial fixed point with long range order could arise (see e.g. \cite{TonerRadzihovsky}). Whether and when such non-trivial fixed points can be realized in models with multipolar symmetry is an important problem for future work.

We also emphasize that when the ordered phase does not exist, we did not provide any argument for what phase should replace it. For example, in 2 spatial dimensions there can be no ordered phases of continuous monopole (ordinary) symmetries. For $O(n)$ models with $n>2$, the result is that the disordered phase is the only phase \cite{polyakov}. For $n=2$, there can in addition be quasi-long-range-ordered phases. Determining what kind of phase {\it can} obtain in the absence of long range ordered symmetry breaking is beyond the scope of this work, and would at a minimum require understanding the nature and role of topological defects in the symmetry breaking order parameter, akin to vortices in the XY model. The range of possible symmetry unbroken phases could be even richer in the presence of disorder, where various glassy phases could also come into play \cite{Fisheretal}. Our discussion of disorder physics was also limited to quenched short-range correlated disorder. Extensions to disorder with long-range correlations, or annealed disorder, are left to future work.

We should also emphasize that our discussion has utilized standard concepts from statistical physics, which in turn amounts to assuming  ergodicity. However, quantum dynamics with multipolar symmetries can break ergodicity \cite{KHN, Sala}, in which case our analysis would not straightforwardly apply. It is however believed that the strict ergodicity breaking is limited to systems with strictly short range interactions (below some critical range) and that systems in which the interactions have long range tails (whether power law or exponential) should generically obey ergodicity at long times (although see \cite{NS}). Since long range tails are generic in physical systems, we believe our arguments should generically apply.

Another setting for generalized Mermin-Wagner-type arguments is higher form global symmetries~\cite{GKSW, Lake, Marvin}. It would also be interesting to see what sort of subtleties could exist in the spontaneous breaking of those symmetries, through partial symmetry breaking or disorder. Since the order parameters for higher-form symmetries are nonlocal, it is difficult to couple disorder directly to the order parameters. In the case of arbitrary perturbations the symmetry becomes broken microscopically, but emerges at long wavelengths. Could disorder have any effect on the Mermin-Wagner behavior of higher-form symmetries? We leave these questions for future work.

Finally, there exists a body of work on generalized Mermin-Wagner arguments in systems with subsystem symmetries \cite{Batista2005, SeibergA, SeibergB, SeibergC, Gorantla2021, Distler2021}. Subsystem symmetries are rather different in character to the multipolar symmetries discussed herein, but are also related to fracton phases via duality \cite{VHF2}. There can be theories with subsystem symmetries where symmetry breaking cannot occur even above the critical dimension, due to the UV/IR mixing~\cite{Gorantla2021}. However, it is always possible to write down theories that saturate the generalized Mermin-Wagner bound~\cite{Distler2021}.
Exploration of connections between the present work and the literature on subsystem symmetries would also be a fruitful topic for future work.

{\bf Acknowledgements} This work was supported by
the U.S. Department of Energy, Office of Science, Basic
Energy Sciences, under Award \# DE-SC0021346

\bibliography{big}

\end{document}

### 2021-11-14 multipoleSSB/mediation.tex
first pull from overleaf

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[margin=1in]{geometry} 
\usepackage{mathtools}
\usepackage{xcolor}

\newcommand{\note}[1]{\textcolor{red}{#1}}
\newcommand{\ex}[1]{\left\langle#1\right\rangle}
\newcommand{\nn}{\nonumber\\}
\renewcommand{\th}[1]{\frac{1}{#1}}
\newcommand{\half}{\th{2}}
\newcommand{\IR}{\text{IR}}
\renewcommand{\l}{\ell}
\newcommand{\cl}{\text{cl}}
\newcommand{\q}{\text{q}}
\newcommand{\goesto}{\rightarrow}

\title{Dipole mediated symmetry breaking}
\author{charlie}
\date{October 2021}

\begin{document}

\maketitle

Consider a system with an exact dipole symmetry, with fields $\phi$ and $\phi_\mu$ that transform as monopoles and dipole under the symmetry, respectively. They should have kinetic terms that look like $(\partial_\mu \partial_\nu \phi)^2$ and $(\partial_\mu \phi_\nu)^2$.  We can also allow them to have a coupling $g(\phi_\mu - \partial_\mu \phi)^2$

\end{document}

### 2021-11-14 multipoleSSB/multipole.tex
first pull from overleaf

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[margin=1in]{geometry} 
\usepackage{mathtools}
\usepackage{xcolor}

\newcommand{\note}[1]{\textcolor{red}{#1}}
\newcommand{\ex}[1]{\left\langle#1\right\rangle}
\newcommand{\nn}{\nonumber\\}
\renewcommand{\th}[1]{\frac{1}{#1}}
\newcommand{\half}{\th{2}}
\newcommand{\IR}{\text{IR}}
\renewcommand{\l}{\ell}
\newcommand{\cl}{\text{cl}}
\newcommand{\q}{\text{q}}
\newcommand{\goesto}{\rightarrow}

\title{CMW for Multipole Systems}
\author{charlie}
\date{May 2021}

\begin{document}

\maketitle

Throughout this document I will use $d$ for space dimensions and $D$ for spacetime dimensions.

\section{Review monopole systems}

Consider the Hamiltonian
\begin{align}
H &= \int d^dx\, \left( \partial_i\sigma \partial_i\sigma + m^2|\sigma|^2 + \lambda|\sigma|^4 + \cdots \right),
\end{align}
with the symmetry $\sigma(x)\goesto e^{i c} \sigma(x)$. There are two phases. For $m^2>0$ we have a disordered phase with $\ex{\sigma}=0$ and for $m^2<0$ we have an ordered phase with $\ex{\sigma} \ne 0$.

The diagnostic for symmetry breaking is the correlation function,
\begin{align}
C(x) &= \ex{\sigma(x)\sigma(x)} - \ex{\sigma(x)\sigma(0)}\dots
\end{align}

Deep in the ordered phase, we can ignore longitudinal fluctuations and write $\sigma(x) = e^{i\phi(x)}$, with
\begin{align}
H &= \int d^d x(\partial_i \phi \partial_i \phi),
\end{align}
which now has an honest shift symmetry $\phi(x)\goesto \phi(x)+c$. This is the Hamiltonian we use to evaluate the validity of the ordered phase. However, it is only valid deep in the ordered phase!

\section{Effective field theories}

Euclidean field theory.

For monopole symmetries, the relationship between the original and ssb effective field theories is straightforward. Say we have a field $\sigma(x)\in \mathbf{C}$ and a symmetry $\sigma(x)\goesto e^{i c} \sigma(x) $, with $c\in \mathbf{R}$ a constant. \note{This is NOT a shift symmetry. Does it make sense to call it a monopole symmetry?} Say the Hamiltonian is
\begin{align}
H &= \int d^dx\, \left( \partial_i\sigma \partial_i\sigma + V(|\sigma|) \right)\nn
&= \int d^dx\, \left( \partial_i\sigma \partial_i\sigma + m^2|\sigma|^2 + \lambda|\sigma|^4 + \cdots \right),
\end{align}
so that when $m^2>0$ the symmetry is unbroken. Now let $m^2<0$ and let the potential be deep enough that $\sigma$ can effectively be described as $\sigma=e^{i\phi(x)}$.

The effective Hamiltonian for $\phi$ is 
\begin{align}
H &= \int d^dx\,\left( \partial_i\phi \partial_i\phi + \cdots \right),
\end{align}
which now has a monopole symmetry that \emph{is} a shift symmetry, $\phi(x)\goesto \phi(x)+c$

On the other hand, if we start with the Hamiltonian
\begin{align}
H &= \int d^dx \left( \partial_i\partial_j \sigma \partial_i \partial_j \sigma + \cdots \right),
\end{align}
then the Hamiltonian in terms of $\phi$ is
\begin{align}
H &= \int d^dx \big( (\partial_i \partial_j \phi) (\partial_i \partial_j \phi) + (\partial_i \phi \partial_j \phi) (\partial_i \partial_j \phi) + (\partial_i \partial_j \phi) (\partial_i \phi \partial_j \phi) + (\partial_i \phi \partial_j \phi) (\partial_i \phi \partial_j \phi) \big),
\end{align}
which has a bunch of interaction terms. \note{Does that matter?}

\note{Does Gromov care about symmetry breaking? How do Griffin et al. handle this?}

\section{Quantum systems}

Ref.~\cite{Griffin2015} considers nonrelativistic quantum systems with the action
\begin{align}
S = \int dt\, d^dx \left\{ \half \dot{\phi}^2 -\half \zeta_n^2 (\partial_{i_1} \dots \partial_{i_n} \phi)^2 + \cdots \right\},
\end{align}
where $\cdots$ represents higher order terms. Such an action has a dynamical critical exponent $z = n$. Although this exponent is fixed, we will leave factors of $z$ explicit for clarity.

Since the action is dimensionless and $[dt\,d^dx] = -z-d$ in units of spatial momentum, we have 
\begin{align}
[\phi] = \frac{d-z}{2},\qquad [\zeta_n] = z-n = 0,
\end{align}
so that $\phi$ is dimensionless at $d=z=n$. Ref.~\cite{Griffin2015} then shows that the $\phi$ field cannot exist at or below $d=n$ as follows.

Consider the propagator for the $\phi$ field,
\begin{align}
\ex{\phi(t,x) \phi(0)} = \int\frac{d\omega\, d^dk}{(2 \pi)^{d + 1}} \frac{e^{ik\cdot x -\omega t}}{\omega^2 + k^{2d} + \mu^{2d}},
\end{align}
where $\mu$ is an IR regulator. For large $|x|$ we can set $t=0$, so that
\begin{align}
\ex{\phi(t,x) \phi(0)} &= \int \frac{d^dk}{(2 \pi)^{d + 1}}e^{ik\cdot x} \int_{-\infty}^{\infty} \frac{d \omega}{\omega^2 + (k^d + \mu^d)^2}\nn
&= \pi \int \frac{d^dk}{(2 \pi)^{d + 1}} \frac{e^{ik\cdot x}}{k^{d}+\mu^{d}}\nn
&= \th{2(2\pi)^d}\int d^d\l\, \frac{e^{i\mu\,\l\cdot x}}{\l^d + 1}\nn
&= \frac{\Omega_d}{2(2\pi)^d}\int_0^\infty \l^{d-1} d\l \int \frac{d\phi}{2\pi}\frac{e^{i\mu|x|\l\cos \phi}}{\l^d+1}\nn
&= \th{(4\pi)^{d/2}\Gamma(d/2)}\int_0^\infty \frac{d \l\, J_0( \mu|x|\l)}{\l+ \l^{1-d}},
\end{align}
which \note{I still don't know how to do}. After some manipulations,
\begin{align}
\ex{\phi(t,x) \phi(0)} &= -\th{(4\pi)^{d/2} \Gamma(d/2)}\log(\mu|x|),
\end{align}
matching Eqn.~2.25 in Ref.~\cite{Griffin2015}.

Personally, I prefer to introduce $\mu$ as a cutoff, so
\begin{align}
\ex{\phi(t,x) \phi(0)} &= \int \frac{d^dk}{(2 \pi)^{d + 1}}e^{ik\cdot x} \int_{-\infty}^{\infty} \frac{d \omega}{\omega^2 + k^{2d}}\nn
&\sim \int \frac{d^dk}{(2 \pi)^{d + 1}} \frac{e^{ik\cdot x}}{k^d}\nn
&\sim \int_\mu^\Lambda \frac{dk}{2\pi} \frac{J_0(k |x|)}{k}\nn
&\sim -\log(\mu|x|),
\end{align}
where we have ignored constants.

Regardless, the point is that the lower critical dimension is $d=z=n$, and that it doesn't matter how we introduce the IR regulator.

\section{Nontrivial dynamical critical exponents}

Usually, we talk about a correspondence between \begin{itemize}
    \item phase transitions in classical systems with $d_\cl$ space dimensions at finite temperature
    \item phase transitions in quantum systems with $d_q=d_\cl-1$ space dimensions at zero temperature.
\end{itemize}
The first step in the argument is to Wick rotate the quantum theory to obtain a Euclidean theory in $d_q+1$ space dimensions. That theory can then be reinterpreted as a classical theory where all dimensions are spatial.

A key part of the argument was that the Euclidean theory was isotropic, because the quantum theory was relativistic. If the original quantum theory instead had a nontrivial critical exponent $z\in \textbf{Z}, z\ne 1$, the argument is different.

The most naive solution is to let $d_\cl = d_\q + z$. Let's see if that works.

\section{Classical systems}

Compare to Griffin.

The Hamiltonian we care about is
\begin{align}
\beta H &= \half\int d^dx\, (\partial_{ i_1} \dots \partial_{ i_n}\phi) (\partial_{ i_1} \dots \partial_{ i_n}\phi) \nn
&= -\half\int d^dx\, \phi(\partial^2)^n \phi,
\end{align}
and we will set $\beta=1$.

\subsection{Naive energy scaling}

Start with the clean case, with no disorder. Then there is no energy gain from forming domains. Say we form domains anyway, with size $L$. In the original $n=1$ case the order parameter varies linearly over a size $L$, say in the $z$ direction. The energy cost is
\begin{align}
\Delta H &= \int d^{d-1}x \int dz (\partial_z \phi)^2\nn
&= L^{d-1} \int dx \th{L^2}\nn
&= L^{d-2},
\end{align}
so domains proliferate if $d\le 2$.

For larger $n$ we want to minimize $(\partial_z^n \phi)^2$. Naively, we can do this by gluing together regions where $\phi$ scales as an order-$n$ polynomial (\note{Can we do better?}). For example, for $n=2$ we would have
\begin{align}
\phi = \begin{cases}
    2\left(\frac{z}{L}\right)^2, & 0<z<\frac{L}{2}\\
    -2\left(\frac{z-L}{L}\right)^2+1, & \frac{L}{2}<z< 1
\end{cases},
\end{align}
so that $(\partial_z^2 \phi)^2=L^{-4}$ is constant. Then, the energy cost is
\begin{align}
\Delta H &= \int d^{d-1}x \int dz (\partial_z^n \phi)^2\nn
&= L^{d-1} \int dx L^{-2n}\nn
&= L^{d-2n},
\end{align}
so the critical dimension is $d=2n$. Recall that in the quantum case we had $d_\q=z=n$, so our naive energy scaling argument matches the naive prediction $d_\cl = d_\q+z = 2n$. Good!

Now say there is some Gaussian quenched disorder and the order parameter forms domains of size $L$. Regardless of $n$, the energy gains from these domains is $L^{d/2}$.  The gain and cost balance out when $d/2 = d-2n$, so the critical dimension is now $d=4n$. That seems large, but it matches the $n=1$ case so it's at least not crazy.

\subsection{Diverging correlation function}

We want to use the correlation function to diagnose long-range order. If the correlation function remains finite at large distance, then the phase is ordered.
The correlation function that we care about is  
\begin{align}
C(x) &= \half \ex{(\phi(x)-\phi(0))^2} =  \langle\phi(0)\phi(0) - \phi(x)\phi(0)\rangle\nn
&= \int\frac{d^dk}{(2\pi)^2} \frac{1-e^{ik \cdot x}}{k^{2n}} \nn
&= \int dk\, k^{d-1-2n} \int_0^{2\pi} \frac{d\phi}{2\pi}(1 - e^{ikx \cos\phi} )\nn
&= \int dk\, k^{d-1-2n} (1 - J_0(kx)),
\end{align}
where $J_0(kx)$ is a Bessel function and we're ignoring constants. 
%We care about the asymptotic forms $J_0(kx) = 1$ for $kx<<1$ and $J_0(kx) = \sqrt{2/(\pi kx)}\cos(kx-\pi/4)$ for $kx>>1$. 
% The infrared regulator enters either in the form of a cutoff $\mu_\IR$ or by modifying the propagator to
% \begin{align}
% G(x) = \int\frac{d^dk}{(2\pi)^d}\frac{e^{ik\cdot x}}{k^{2n} + \mu_\IR^{2n}},
% \end{align}
% where we now see that $\mu_\IR$ has the interpretation of mass only when $n=1$.
We will introduce an IR cutoff into the integral, rather than as a regulator.

If we take $n=1$ we should recover the CMW theorem,
\begin{align}
C(x) &= \int_\mu^\Lambda dk\, k^{d-3} (1 - J_0(kx)) \nn
&= x^{2-d} \int_{\mu x}^{\Lambda x} dy\, y^{d-3} (1 - J_0(y)) \nn
&= \begin{cases}
\Lambda^{d-2}, & d>2,\\
\log(\Lambda x) - \log(\mu x), & d=2,\\
x^{2-d}, & d<2,
\end{cases}
\end{align}
matching Griffin 2.21. \note{Redo for regulator analysis?}

For general $n$ everything goes through the same way and we get
\begin{align}
C(x) &= \int_\mu^\Lambda dk\, k^{d-1-2n} (1 - J_0(kx)) \nn
&= \begin{cases}
\Lambda^{d-2}, & d>2n,\\
\log(\Lambda x) - \log(\mu x), & d=2n,\\
x^{2-d}, & d<2n,
\end{cases}
\end{align}
so that now the critical dimension is $2n$, matching the energy scaling. The result is that above the critical dimension, the correlation function asymptotes to a finite value, so the symmetry must be spontaneously broken. Below the critical dimension, the correlation function diverges and the symmetry is preserved.

Now, in the case of quenched disorder we will follow the conventions of Imry and Ma~\cite{ImryMa} where
\begin{align}
\Delta H = -\int d^dx\, h(x) \sigma(x),
\end{align}
with $\sigma(x) = e^{i\phi(x)}$ and $h(x)\in \mathbf{C}$. Then 
\begin{align}
G_\perp(k) &= \th{k^{2n}},
\end{align}
and
\begin{align}
\ex{m_\perp(x)m_\perp(0)} &= \int d^dk\, G_\perp^2 e^{ik \cdot x}\nn
&= \int dk\, k^{d-4n-1}J_0(kx)
\end{align}
which diverges for $d\le 4n$, reproducing the naive energy scaling.

\bibliographystyle{unsrt}
\bibliography{big}
\end{document}

### 2021-11-14 multipoleSSB/overview.tex
first pull from overleaf

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[margin=1in]{geometry} 
\usepackage{mathtools}
\usepackage{xcolor}

\newcommand{\note}[1]{\textcolor{red}{#1}}
\newcommand{\ex}[1]{\left\langle#1\right\rangle}
\newcommand{\nn}{\nonumber\\}
\renewcommand{\th}[1]{\frac{1}{#1}}
\newcommand{\half}{\th{2}}
\newcommand{\IR}{\text{IR}}
\renewcommand{\l}{\ell}
\newcommand{\cl}{\text{cl}}
\newcommand{\q}{\text{q}}
\newcommand{\goesto}{\rightarrow}

\title{Project Overview}
\author{charlie}
\date{May 2021}

\begin{document}

\maketitle

I will use this document to keep track of the goals of this project and to keep track of the documents in this overleaf folder.

\section{The main goal}

Various phases that were thought to be beyond the symmetry-breaking paradigm have recently been described using the spontaneous breaking of exotic symmetries, such as subsystem symmetries, higher symmetries, and multipole symmetries. For some of these symmetries, CMW-like theorems exist in the literature. None have Imry-Ma-type theorems.

Some systems, like the X-cube model, have alternative descriptions in terms of multipole or higher-form symmetries. This type of system can provide a non-trivial check on whatever symmetry-breaking theorems we come up with.

\section{Files}

\begin{itemize}
\item \texttt{multipole.tex} has some CMW and Imry-Ma theorems.

\item \texttt{dipole1d.tex} has attempts at finding a microscopic theory that supports partial multipole symmetry breaking.
\end{itemize}

\bibliographystyle{unsrt}
\bibliography{big}
\end{document}

### 2021-11-14 multipoleSSB/partialbreaking.tex
first pull from overleaf

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[margin=1in]{geometry} 
\usepackage{mathtools}
\usepackage{xcolor}

\newcommand{\note}[1]{\textcolor{red}{#1}}
\newcommand{\ex}[1]{\left\langle#1\right\rangle}
\newcommand{\nn}{\nonumber\\}
\renewcommand{\th}[1]{\frac{1}{#1}}
\newcommand{\half}{\th{2}}
\newcommand{\IR}{\text{IR}}
\renewcommand{\l}{\ell}
\newcommand{\cl}{\text{cl}}
\newcommand{\q}{\text{q}}
\renewcommand{\max}{\text{max}}
\newcommand{\goesto}{\rightarrow}

\title{Breaking monopole symmetry to dipole symmetry}
\author{charlie}
\date{October 2021}

\begin{document}

\maketitle

\section{Fields}

We have a field $\phi$ that transforms into $\phi + a^\mu x_\mu + b$ under a dipole symmetry. If $\phi$ is well-defined and constant, then the symmetry is fully broken. We only want to consider fields with constant values. How can we define a field that breaks the symmetry group to the monopole group? It must change under the $a^\mu$ part of the transformation, but somehow be rotationally invariant.

It is tempting to use a field like $\partial_\mu \phi$. The problem then is that if $\partial_\mu \phi$ is well-defined, then $\phi$ must be as well. Alternatively, it is tempting to use a vector-valued field $\theta_\mu$, but this will necessarily break rotation symmetry. Something like $\theta_\mu \theta^\mu$ would be rotationally invariant, but the ground state would still have to choose a value of $\theta_\mu$.

Maybe it would help to look at the group structure. The group $\mathcal{M}^2_\max$ has a space subgroup, a dipole subgroup $U(1)^d$, and a monopole subgroup $U(1)$. The space subgroup and the monopole subgroup combine to form $\mathcal{M}^1_\max$, but the dipole and space subgroups cannot combine without including the monopole subgroup. This should not be a problem though, as $\mathcal{M}^1_\max$ is a perfectly valid subgroup.

If we don't want to break rotational invariance, our order parameter needs to be a scalar\dots

\section{On the lattice}

Ignoring the question of whether we can do this in the continuum, can we partially break the symmetry on the lattice? The system
\begin{align}
H_0  &= H_b + H_d + H_\text{int}\nn
H_b &= \sum_j \left[ A  n_j + B n_j^2 \right] \nn
&\qquad + \sum_{j,\mu,\nu} C \left[  b_j b^{\dag}_{j+\mu} b^{\dag}_{j+\nu} b_{j+\mu+\nu} + H.C. \right]\nn
H_d &= \sum_{j,\mu} \left[ A' n'_{j,j+\mu} + B' (n'_{j, j + \mu})^2 \right]\nn
&\qquad  + \sum_{j,\mu,\nu} C' \left[d^{\dag}_{j, j + \mu} d_{j + \nu, j+\mu + \nu} +H.C.\right]\nn
H_\text{int} &= \sum_{j,\mu} g\left[ b_j d_{j,j+\mu} b^\dag_{j+\mu} + H.C. \right],
\end{align}
\emph{almost} manages to break dipole without breaking rotations. Say that in the dipole condensate we have $d_1=e^{i\phi_1}$, $d_2=e^{i\phi_2}$, and $d_3=e^{i\phi_3}$.

\end{document}

### 2021-11-14 multipoleSSB/rotors.tex
first pull from overleaf

\documentclass[12pt]{article}
\usepackage[margin=1in]{geometry} 
\usepackage{mathtools}
\usepackage{xcolor}
\usepackage{hyperref}

\newcommand{\note}[1]{\textcolor{red}{#1}}
\newcommand{\nn}{\nonumber\\}
\newcommand{\goesto}{\rightarrow}
\renewcommand{\int}{\text{int}}
\newcommand{\ex}[1]{\left\langle#1\right\rangle}
\newcommand{\op}[1]{\mathcal{O}^{(#1)}}

\title{Quantum rotors with multipole symmetry}
\author{charlie}
\date{August 2021}

\begin{document}

\maketitle

\section{In 1 dimension}

Let's start with the simpler case of 1d, so that there is only one type of link. Obviously CMW theorems and Imre-Ma stuff need more dimensions, but 1d is simple.

On sites, we have degrees of freedom with states $|\theta\rangle$, $0\le \theta<2\pi$. The operators are $\theta$ and $L = -i\frac{\partial}{\partial\theta}$. On links, let the operators be $\phi$ and $J = -i\frac{\partial}{\partial\phi}$. One symmetry operator is 
\begin{align}
\mathcal{O}^{(2)}(\xi) &= \prod_j e^{i \xi( jL_j + J_{j, j+1} )}
\end{align}
which acts by rotating all variables in the same direction. On the links, this looks like a regular monopole U(1) symmetry operator. But on the sites it acts as a multipole symmetry operator. The other symmetry operator
\begin{align}
\mathcal{O}^{(1)}(\xi) = \prod_j e^{i \xi L_j},
\end{align}
acts as a regular monopole operator on the sites.

\note{Note that the most trivial version of this system would split $\op{2}$ into two symmetry operators, a monopole on the link rotors and a dipole on the sites. This would fully decouple the symmetries.}

The Hamiltonian we are interested in is 
\begin{align}
H_b &= \sum_i \; A  L_i + B L_i^2 + C \cos \left( \theta_{i-1} - 2\theta_i+\theta_{i+1} \right)\nn
H_d &= \sum_i \; A' J_{i,i+1} + B' J_{i,i+1}^2 + C' \cos \left( \phi_{i-1,i}-\phi_{i,i+1} \right) \nn
H_{\int} &= g \sum_i \; \cos \left( \theta_{i}+\phi_{i,i+1} - \theta_{i+1} \right) \nn
&= \frac{g}{2}\sum_i\left( e^{i\left( \theta_{i}+\phi_{i,i+1} - \theta_{i+1} \right)} + e^{-i\left( \theta_{i}+\phi_{i,i+1} - \theta_{i+1} \right) } \right)
\end{align}
The ground states are either number eigenstates or angle eigenstates. The advantage here is that the Hamiltonian is exactly solvable. The disadvantage is that the physical picture of condensed link variables facilitating hopping is not present.

\subsection{Noninteracting phases}

The noninteracting phases ($g=0$) are somewhat uninteresting. This is because symmetry breaking for the site systems behaves like dipole symmetry breaking, while the link systems undergo monopole symmetry breaking. To see this, consider the four phases.

In the trivial phase, with $C,C'<<A,B,A',B'$, all rotors are in number eigenstates. The state is gapped and both symmetry operators act as an overall phase. We can write the ground state as $|n;m\rangle$, with energy
\begin{align}
E_{n;m} &= N\left(An+Bn^2+A'm+B'm^2\right),
\end{align}
where $N$ is the number of sites. If, for example, one of the sites were in state $|n+1\rangle$ instead of $|n\rangle$, the energy would be 
\begin{align}
N\left(An+Bn^2+A'm+B'm^2\right) + A+2Bn+B,
\end{align}
which would be higher than $E_{n;m}$ by a finite amount if $|n;m\rangle$ is the ground state. Call this phase I.

If we let $C'>>A',B'$, then the link states will be angle eigenstates. Let us write the ground state (call it phase II) as $|n;\Phi\rangle$, with energy
\begin{align}
E_{n;\Phi} = N(An+Bn^2).
\end{align}
Clearly, we have a Goldstone mode. More generally, if we let $\tilde{\Phi}(j)$ be an arbitrary function, then
\begin{align}
E_{n;\Phi+\tilde{\Phi}(x)} = E_{n;\Phi} + \sum_jC' \cos( \Delta \tilde{\Phi}(j)),
\end{align}
with a dispersion quadratic in the first derivative of the angle. \note{Is this a monopole symmetry breaking phase or a dipole symmetry breaking phase?} The operator $\mathcal{O}^{(1)}$ acts as a phase but $\mathcal{O}^{(2)}$ moves within the ground state manifold. I guess the interesting question is: which CMW theorem does this phase obey?

In phase III, if we instead let $C>>A,B$, the site variables condense. Since we have the lattice second derivative in the cosine, the dispersion is now quadratic in the second derivative of the angle. Ground states can be written as 
\begin{align}
|\Theta_1,\Theta_2; m\rangle = \bigotimes_j |\Theta_1 + j\Theta_2\rangle_j |m\rangle_{j,j+1},
\end{align}
so that either symmetry operator changes the state within the ground state manifold. Thus, we have true dipole symmetry breaking.

Nothing extra happens when we let both variable types condense, in a state we will call phase IV. The takeaway is that the $g=0$ system should display two types of CMW-type theorems, but they should match either system independently. \note{That being said, should we be careful about operators like $\op{2}$ that act as a dipole operator on one sector and a monopole operator on another sector?}

\subsection{CMW in the noninteracting case}

\note{Obviously we need to generalize the system to higher dimensions before we can talk about CMW. But assuming it goes through in the simplest way possible\dots}

Phase I should exist in any dimension. Since phase II has a $k^2$ dispersion, it should obey the monopole CMW theorem and exist for $d>1$. Note that this is true even though the dipole symmetry operator appears to be broken.

Naively, I would expect phases III and IV to exist for $d>2$, since they both have Goldstone bosons with $k^4$ dispersions coming from the site variables. At $d=2$, the site variables would become disordered, so that phase III would be demoted to phase I and phase IV to phase II. At $d=1$ the only valid phase is phase I (modulo topological defects \note{what is that behavior called? Luttinger liquid?})

\subsection{Interacting phases}

What happens if we now turn on $g>0$? There can be no fixed points with finite $g, A,$ and $B$, but \note{think about which way RG would flow}. Perturbation theory?

In phase IV, we can include the interaction term nonperturbatively. The Hamiltonian,
\begin{align}
H = C \cos \left( \theta_{i-1} - 2\theta_i+\theta_{i+1} \right) + C' \cos \left( \phi_{i-1,i}-\phi_{i,i+1} \right) + g \cos \left( \theta_{i}+\phi_{i,i+1} - \theta_{i+1} \right),
\end{align}
has eigenstates $|\Theta_1,\Theta_2;\Phi\rangle$ with energy $E = Ng\cos(\Theta_2-\Phi)$. For a general state the energy looks like
\begin{align}
E &= C\cos\left( \Delta^2 \Theta \right) + C' \cos \left( \Delta \Phi \right) + g \cos \left( \Delta \Theta - \Phi \right)\nn
&\simeq C(\partial^2\Theta)^2 + C'(\partial\Phi)^2 + g(\partial\Theta - \Phi)^2
\end{align}
\note{Understand relevance of different operators and NGB dispersions.}

\bibliographystyle{unsrt}
\bibliography{big}
\end{document}

### 2021-11-14 multipoleSSB/subsystem.tex
first pull from overleaf

--- 2021-11-14 multipoleSSB/partialbreaking.tex
@@
 \newcommand{\q}{\text{q}}
-\renewcommand{\max}{\text{max}}
 \newcommand{\goesto}{\rightarrow}
 
-\title{Breaking monopole symmetry to dipole symmetry}
+\title{Subsystem vs Multipole}
 \author{charlie}
@@
 
-\section{Fields}
-
-We have a field $\phi$ that transforms into $\phi + a^\mu x_\mu + b$ under a dipole symmetry.
-If $\phi$ is well-defined and constant, then the symmetry is fully broken.
-We only want to consider fields with constant values.
-How can we define a field that breaks the symmetry group to the monopole group?
-It must change under the $a^\mu$ part of the transformation, but somehow be rotationally invariant.
-
-It is tempting to use a field like $\partial_\mu \phi$.
-The problem then is that if $\partial_\mu \phi$ is well-defined, then $\phi$ must be as well.
-Alternatively, it is tempting to use a vector-valued field $\theta_\mu$, but this will necessarily break rotation symmetry.
-Something like $\theta_\mu \theta^\mu$ would be rotationally invariant, but the ground state would still have to choose a value of $\theta_\mu$.
-
-Maybe it would help to look at the group structure.
-The group $\mathcal{M}^2_\max$ has a space subgroup, a dipole subgroup $U(1)^d$, and a monopole subgroup $U(1)$.
-The space subgroup and the monopole subgroup combine to form $\mathcal{M}^1_\max$, but the dipole and space subgroups cannot combine without including the monopole subgroup.
-This should not be a problem though, as $\mathcal{M}^1_\max$ is a perfectly valid subgroup.
-
-If we don't want to break rotational invariance, our order parameter needs to be a scalar\dots
-
-\section{On the lattice}
-
-Ignoring the question of whether we can do this in the continuum, can we partially break the symmetry on the lattice? The system
+For subsystem symmetries we have the conservation law
 \begin{align}
-H_0  &= H_b + H_d + H_\text{int}\nn
-H_b &= \sum_j \left[ A  n_j + B n_j^2 \right] \nn
-&\qquad + \sum_{j,\mu,\nu} C \left[  b_j b^{\dag}_{j+\mu} b^{\dag}_{j+\nu} b_{j+\mu+\nu} + H.C. \right]\nn
-H_d &= \sum_{j,\mu} \left[ A' n'_{j,j+\mu} + B' (n'_{j, j + \mu})^2 \right]\nn
-&\qquad  + \sum_{j,\mu,\nu} C' \left[d^{\dag}_{j, j + \mu} d_{j + \nu, j+\mu + \nu} +H.C.\right]\nn
-H_\text{int} &= \sum_{j,\mu} g\left[ b_j d_{j,j+\mu} b^\dag_{j+\mu} + H.C. \right],
+\frac{d}{dt} \int_{y=a} dx \; \rho = \frac{d}{dt} \int_{x=b} dy \; \rho = 0.
 \end{align}
-\emph{almost} manages to break dipole without breaking rotations. Say that in the dipole condensate we have $d_1=e^{i\phi_1}$, $d_2=e^{i\phi_2}$, and $d_3=e^{i\phi_3}$.
+This is equivalent to the conservation law
+\begin{align}
+0 &= \frac{d}{dt} \int dxdy \, \big ( f(x) + g(y) \big )\rho,
+\end{align}
+for arbitrary functions $f$ and $g$.
+We can then Taylor expand these functions.
+This makes the conservation law look like a conservation law for a multipole symmetry where the polynomials are $x, x^2, x^3, \dots$ and $y, y^2, y^3, \dots$
 

### 2021-11-26 multipoleSSB/manuscript.tex
ready for submission
starting edits from rahul and ethan

@@
-\documentclass[twocolumn, longbibliography]{revtex4-2}
+\documentclass[pra,aps,twocolumn, amsfonts,amsmath,amssymb,nofootinbib,superscriptaddress]{revtex4}
 
@@
 	linkcolor=blue,   
-	urlcolor=blue,
+	urlcolor=blue
 }
@@
 \newcommand{\todo}[1]{\textcolor{orange}{#1}}
-\newcommand{\charlie}[1]{\textcolor{TealBlue}{#1}}
+\newcommand{\charlie}[1]{\textcolor{Blue}{#1}}
 \newcommand{\RN}[1]{\textcolor{TealBlue}{#1}}
@@
 \author{Charles Stahl}
+\affiliation{Department of Physics and Center for Theory of Quantum Matter, University of Colorado Boulder, Boulder CO 80309, USA}
+\author{Ethan Lake}
+\affiliation{Department of Physics, Massachusetts Institute of Technology, Cambridge, MA, 02139}
 \author{Rahul Nandkishore}
@@
 In Section \ref{multipolegroup} we introduce the multipole group, how to build multipole-invariant field theories, and the generalized Mermin-Wagner argument for total breaking of the maximal multipole symmetry.
-In Section~\ref{sec:evade} we show how to evade the generalized Mermin-Wagner argument by considering phases in which multipole symmetries are partially broken.
-We discuss generalized Imry-Ma arguments in the presence of quenched disorder in Section~\ref{sec:disord}.
+In Section~\ref{sec:partial} we explore phases in which multipole symmetries are partially broken.
+Then, we discuss generalized Imry-Ma arguments in the presence of quenched disorder in Section~\ref{sec:disord}.
 Finally, we consider an explicit lattice model that illustrates some of these ideas in Section~\ref{sec:example} before concluding with a discussion of open questions in \ref{sec:disc}.
@@
 This is the sense in which the Haah group mentioned earlier corresponds to a field theory for Haah's code~\cite{BB}.
-See Ref.~\cite{Gromov2019} for the full story.
+See Ref.~\cite{BB, Gromov2019} for the full story.
 
@@
 It is interesting to note that the dynamical critical exponent in these theories is $z = a+1$, so the classical and quantum critical dimensions are related by $d_\text{cl} = d_\text{q} + z$.
+
+\charlie{Is it more useful to say $d_\text{cl} = d_\text{q} + z$ or $d_\text{cl} = 2 d_\text{q}$, or not try to make any general statement at all?}
 
@@
 
-\section{Evading the generalized Mermin-Wagner argument} \label{sec:evade}
+\section{Partial breaking of multipole symmetries} \label{sec:partial}
 
@@
 In physical terms, we are taking a system that conserves total charge and total dipole moment, then condensing objects that carry a dipole moment but no total charge.
-We then recognize that these fundamental dipoles, which are of course the dipole charges of the monopole symmetry, are also the monopole charges of the dipole symmetry.
+\note{If $\phi(x)$ is charged under the monopole symmetry, then the operator $\partial_i \phi(x)$ is charged under the dipole part of the symmetry group.
+\charlie{Do we want correlation functions of $\partial_i\phi$ or $\exp(\partial_i\phi)$?} In a classical system at $T>0$ the correlation functions of this operator are
+\begin{align}
+\left\langle \partial_i\phi(x) \partial_j\phi(0) \right\rangle &\sim \delta_{ij} \int \frac{d^dk\, k^2}{k^4},
+\end{align}
+so that the critical dimension is $d=2$, as in the standard Mermin-Wagner theorem. }
+
+%We then recognize that these fundamental dipoles, which are of course the dipole charges of the monopole symmetry, are also the monopole charges of the dipole symmetry.
 We should then expect symmetry breaking cannot occur for $d\le 2$ classically or $d\le 1$ quantum mechanically.
@@
 
-In more generality, the order parameter for this pattern of symmetry breaking will be an object $\phi_{i_1\dots i_m}$, where $m=b+1$.
+\note{In more generality, for this pattern of symmetry breaking there will be a charged operator $\partial_{i_1} \dots \partial{i_m} \phi$, where $m=b+1$.
 Because $G$ is a maximal multipole group, we can still write the polynomials as in Eqn.~\ref{eqn:basis}.
-For our new order parameter, they act as
+As long as $\phi$ transforms as
 \begin{align}
-\delta \phi_{i_1 \dots i_m} = \lambda_{I_c} \partial_{i_1} \dots \partial_{i_m} P_c^{I_c}, \label{eqn:partial}
+\delta \phi = \lambda_{I_c} P_c^{I_c}, \label{eqn:partial}
 \end{align}
-so that $\phi_{i_1 \dots i_m}$ is invariant under $H=\mathcal{M}^b_\max$.
+then $\partial_{i_1} \dots \partial_{i_m} \phi$ is invariant under $H=\mathcal{M}^b_\max$. }
 
-The kinetic term should take the form
+\note{Classical correlation functions for these operators take the form
 \begin{align}
-K[\phi(x)] &= (\partial_{i_{m+1}}\dots \partial_{i_n} \phi_{i_1 \dots i_m})^2,
+C &= \left\langle \partial_{i_1} \dots \partial_{i_m} \phi(x) \partial_{j_1} \dots \partial_{j_m} \phi(0) \right\rangle \nn
+&\sim \int \frac{d^dk\, k^{2m}}{k^{2n}},
 \end{align}
-which is invariant under shifts of the form of Eq.~\ref{eqn:partial}.
+and diverge when $d \le 2(n-m) = 2(a-b)\equiv d_\text{cl}$. On the other hand, the $T=0$ correlation functions behave as
+\begin{align}
+C &= \left\langle \partial_{i_1} \dots \partial_{i_m} \phi(x) \partial_{j_1} \dots \partial_{j_m} \phi(0) \right\rangle \nn
+&\sim \int \frac{d\omega\, d^dk\, k^{2m}}{\omega^2 + k^{2n}}\nn
+&\sim \int \frac{d^dk\, k^{2m}}{k^n},
+\end{align}
+so the quantum critical dimension is $d_\text{q} = n - 2m = a - 2b - 1$.
+We recover the maximal generalized Mermin-Wagner argument if we let $b=-1$ denote the trivial group ($b=0$ denotes the monopole group).
+Furthermore, the critical dimensions again satisfy $d_\cl = d_\text{q} + z$.
+\charlie{Distinguish $z$ and $z_\max$?}
+}
+
+% The kinetic term should take the form
+% \begin{align}
+% K[\phi(x)] &= (\partial_{i_{m+1}}\dots \partial_{i_n} \phi_{i_1 \dots i_m})^2,
+% \end{align}
+% which is invariant under shifts of the form of Eq.~\ref{eqn:partial}.
 The dispersion for the Goldstone modes is $K_k \sim k^{2(n-m)} = k^{2(a-b)}$.
 Following the correlation function calculation above, we find a classical critical dimension of $d = 2(a-b)$ and a quantum critical dimension of $d = a-b$.
-We recover the maximal generalized Mermin-Wagner argument if we let $b=-1$ denote the trivial group ($b=0$ denotes the monopole group).
 
-\subsection{Mediated condensation} \label{sub:mediated}
+\subsection{No mediated condensation} \label{sub:mediated}
 
 We have seen that the generalized Mermin-Wagner argument provides critical dimensions for partial breaking that are lower than the critical dimensions for full symmetry breaking.
-In fact, we can use this observation to evade the generalized Mermin-Wagner theorem, completely breaking any multipole symmetry for $d>2$ classically, or $d>1$ quantum mechanically.
+\note{Can we use this behavior to try to evade the Mermin-Wagner argument for the full symmetry group?}
+
+%\remove{In fact, we can use this observation to evade the generalized Mermin-Wagner theorem, completely breaking any multipole symmetry for $d>2$ classically, or $d>1$ quantum mechanically}.
 This possibility has been pointed out in Refs.~\cite{Griffin2015Cascade, Griffin2015}, but we include a further interpretation in terms of what types of phase transitions are allowed in different dimensions.
@@
 
-To be explicit, consider a monopole field $\phi(x)$ and a dipole field $\phi_i(x)$. Under the dipole symmetry they transform as 
+To be explicit, consider a monopole field $\phi(x)$ and a dipole field $\varphi_i(x)$. Under the dipole symmetry they transform as 
 \begin{align}
 \phi(x) &\goesto \phi(x) + a_ix^i + b,\nn
-\phi_i(x) &\goesto \phi_i(x) + a_i,
+\varphi_i(x) &\goesto \varphi_i(x) + a_i.
 \end{align}
-where we can more explicitly see that $\phi(x)$ transforms like a monopole field under the dipole part of the symmetry. The kinetic term will look like
+The leading terms in the Lagrangian will be
 \begin{align}
-K[\phi,\phi_i] &= (\partial_i \partial_j \phi)^2 + (\partial_i \phi_j)^2,
+\mathcal{L} &= (\partial_t \phi)^2 + (\partial_t \varphi_i)^2 + (\partial_i \partial_j \phi)^2 + (\partial_i \varphi_j)^2,
 \end{align}
-so that $\phi$ has a $k^4$ dispersion and $\phi_i$ has a $k^2$ dispersion.
-We can also include an interaction term of the form $g(\partial_i \phi - \phi_i)^2$, which does not break the symmetry.
-The $\phi_i$ field is well-defined whenever the dipole symmetry is broken.
-As long as $\phi_i$ is well-defined, integrating out fluctuations in $\phi_i$ endows the monopole field $\phi$ with a $k^2$ dispersion.
+so that $\phi$ has a $k^4$ dispersion and $\varphi_i$ has a $k^2$ dispersion.
+We can also include an interaction term of the form $g(\partial_i \phi - \varphi_i)^2$, which does not break the symmetry.
+The $\varphi_i$ field is well-defined whenever the dipole symmetry is broken.
+
+\note{We can then redefine the $\varphi_i$ field to $\tilde{\varphi}_i = \varphi_i - \partial_i \phi$, so that it is neutral under the symmetry.
+The interaction term becomes $g\tilde{\varphi}_i^2$, so that the $\tilde{\varphi}_i$ field becomes massive.
+The only remaining massless field is $\phi$, which remains charged under the full dipole group and keeps its $k^4$ dispersion.
+As a result, the critical dimension for breaking the full dipole group is not affected by the possibility of partially breaking.
+}
+\charlie{What else to say here? Compare to Griffin/Grosvenor?}
+
+%As long as $\phi_i$ is well-defined, integrating out fluctuations in $\phi_i$ endows the monopole field $\phi$ with a $k^2$ dispersion.
 The coupling $g$ is set by how far the theory sits from the dipole critical point.
 
-With a $k^2$ dispersion for the monopole field, we now have a route to evade the (generalized) Mermin-Wagner argument and fully break the dipole symmetry in dimensions $d> 2$.
+%With a $k^2$ dispersion for the monopole field, we now have a route to evade the (generalized) Mermin-Wagner argument and fully break the dipole symmetry in dimensions $d> 2$.
 First, partially break the symmetry so that $\phi_i$ is well defined, which is allowed for any $d>2$.
@@
 
-The preceding procedure gives us insight into what the generalized Mermin-Wagner arguments were telling us.
+%The preceding procedure gives us insight into what the generalized Mermin-Wagner arguments were telling us.
 When we wrote down those arguments, we did not include the intermediate fields.
@@
 
-As in the clean systems, we can have a sequence of phase transitions separated by intermediate phases in which the multipole symmetry is broken to a progressively smaller subgroup.
+%\todo{No more mediated criticality}
+%As in the clean systems, we can have a sequence of phase transitions separated by intermediate phases in which the multipole symmetry is broken to a progressively smaller subgroup.
 With enough order parameter fields, any multipole symmetry can be fully broken in dimension $d>4$, regardless of which order parameter the disorder couples to.
@@
 &\qquad - \sum_{j,\mu,\nu} t_b \left[  b_j b^{\dag}_{j+\mu} b^{\dag}_{j+\nu} b_{j+\mu+\nu} + H.C. \right]\nn
-H_d &= \sum_{j,\mu} \left[ \frac{U_d}{2} n'_{j,j+\mu} (n'_{j,j+\mu} - 1) -\mu \, n'_{j, j + \mu} \right]\nn
+H_d &= \sum_{j,\mu} \left[ \frac{U_d}{2} n'_{j,j+\mu} (n'_{j,j+\mu} - 1) -\mu_d \, n'_{j, j + \mu} \right]\nn
 &\qquad  - \sum_{j,\mu,\nu} t_d \left[d^{\dag}_{j, j + \mu} d_{j + \nu, j+\mu + \nu} +H.C.\right]\nn
@@
 When the edge bosons are condensed, $H_\text{int}$ gives the site bosons an effective single particle hopping, and hence an effective $k^2$ dispersion, which would otherwise be disallowed by symmetry.
+\todo{Yes to effective hopping, no to $k^2$ dispersion.
+Discuss SF/SC?}
 
+\todo{Redo discussion of phases. Reframe in terms of symmetry-breaking patterns, not phases.}
+
+\todo{Define multiple dipole-breaking patterns at the effective level. Relevance of cosine terms and stability to crystal phase?}
 At sufficiently high dimension, this system will support three phases, which may be loosely described as the Mott insulator, the monopole superfluid where site bosons are condensed and the dipole superfluid where edge bosons (but not site bosons) are condensed.
@@
 
-In the absence of disorder, neither symmetry breaking pattern can occur for $d\le 1$. There may still be interesting phases, such as a Luttinger liquid.
-For $d>2$ dipole and monopole symmetry breaking may occur, so all three phases are intact. Furthermore, transitions between any two phases are allowed.
-This leaves the intermediate case of $d=2$, in which all three phases can exist, but there can be no transitions between any symmetry-unbroken phase and the monopole superfluid.
+% In the absence of disorder, neither symmetry breaking pattern can occur for $d\le 1$. There may still be interesting phases, such as a Luttinger liquid.
+% For $d>2$ dipole and monopole symmetry breaking may occur, so all three phases are intact. Furthermore, transitions between any two phases are allowed.
+% This leaves the intermediate case of $d=2$, in which all three phases can exist, but there can be no transitions between any symmetry-unbroken phase and the monopole superfluid.
 Instead, there must be an intermediate phase with dipole but not monopole superfluidity.
@@
 
-\begin{figure}
-\centering
+% \begin{figure}
+% \centering
 
-\begin{tikzpicture}[scale = .9]
-\draw[<->] (0,0) node[left]{(a)}  -- (8,0) node[right]{$s$};
-\draw (4,-.1) -- (4,.1);
+% \begin{tikzpicture}[scale = .9]
+% \draw[<->] (0,0) node[left]{(a)}  -- (8,0) node[right]{$s$};
+% \draw (4,-.1) -- (4,.1);
 
-\node [align=center] (A) at (2, -0.5) {unbroken};
-\node [align=center] (C) at (6, -0.5) {fully\\broken};
+% \node [align=center] (A) at (2, -0.5) {unbroken};
+% \node [align=center] (C) at (6, -0.5) {fully\\broken};
 
-\end{tikzpicture}
-\begin{tikzpicture}[scale = .9]
+% \end{tikzpicture}
+% \begin{tikzpicture}[scale = .9]
 
-\draw[<->] (0,0) node[left]{(b)}  -- (8,0) node[right]{$s$};
-\draw (2.667,-.1) -- (2.667,.1);
-\draw (5.333,-.1) -- (5.333,.1);
+% \draw[<->] (0,0) node[left]{(b)}  -- (8,0) node[right]{$s$};
+% \draw (2.667,-.1) -- (2.667,.1);
+% \draw (5.333,-.1) -- (5.333,.1);
 
-\node [align=center] (A) at (1.333, -0.5) {unbroken};
-\node [align=center] (B) at (4, -0.5) {dipole\\broken};
-\node [align=center] (C) at (6.667, -0.5) {fully\\broken};
-\end{tikzpicture}
+% \node [align=center] (A) at (1.333, -0.5) {unbroken};
+% \node [align=center] (B) at (4, -0.5) {dipole\\broken};
+% \node [align=center] (C) at (6.667, -0.5) {fully\\broken};
+% \end{tikzpicture}
 
-\caption{(a) For  $d>2$, there can be direct transitions between symmetry-unbroken and the fully-broken phase.
+% \caption{\todo{Remove figure} \todo{New figure with different dipole-breaking phases?}(a) For  $d>2$, there can be direct transitions between symmetry-unbroken and the fully-broken phase.
 Here $s$ is intended to represent a parameter along an arbitrary line in the phase diagram that crosses this transition.
@@
 These phase diagrams are for translation invariant systems at zero temperature.}
-\label{fig:phases}
-\end{figure}
+% \label{fig:phases}
+% \end{figure}
 

### 2021-12-08 multipoleSSB/ethans_edits.tex
updates from Ethan

--- 2021-11-26 multipoleSSB/manuscript.tex
@@
-\documentclass[pra,aps,twocolumn, amsfonts,amsmath,amssymb,nofootinbib,superscriptaddress]{revtex4}
+\documentclass[pra,aps,twocolumn, amsfonts,amsmath,amssymb,nofootinbib,superscriptaddress]{revtex4-2}
 
@@
 \newcommand{\half}{\frac{1}{2}}
+\newcommand{\mmax}[1]{\mathcal{M}^{#1}_\max}
+
+\newcommand{\ethan}[1]{ { \color{blue} \footnotesize \textsf{ethan: \textsl{#1}} }}
+\newcommand\be            {\begin{equation}}
+\newcommand\ee            {\end{equation}}
+\newcommand\ba            {\begin{aligned}}
+\newcommand\ea            {\end{aligned}}
+\newcommand{\p}{\partial}
+\renewcommand{\t}{\theta}
+\newcommand{\mcm}{\mathcal{M}}
 
@@
 Multipole symmetries are of interest both as a window on fracton physics and as a crucial ingredient in realizing new universality classes for quantum dynamics.
-Here we address the question of whether and when multipole symmetries can be {\it spontaneously broken} in thermal equilibrium.
+Here we address the question of whether and when multipole symmetries can be spontaneously broken, both in thermal equilibrium and at zero temperature.
 We derive generalized Mermin-Wagner arguments for the total or partial breaking of multipolar symmetry groups and generalized Imry-Ma arguments for the robustness of such multipolar symmetry breaking to disorder.
@@
 This paper is organized as follows.
-In Section \ref{multipolegroup} we introduce the multipole group, how to build multipole-invariant field theories, and the generalized Mermin-Wagner argument for total breaking of the maximal multipole symmetry.
-In Section~\ref{sec:partial} we explore phases in which multipole symmetries are partially broken.
-Then, we discuss generalized Imry-Ma arguments in the presence of quenched disorder in Section~\ref{sec:disord}.
-Finally, we consider an explicit lattice model that illustrates some of these ideas in Section~\ref{sec:example} before concluding with a discussion of open questions in \ref{sec:disc}.
+In Section \ref{multipolegroup} we introduce the multipole group, and how to build field theories invariant under it.
+In section \ref{sec:full_breaking} we discuss generalized Mermin-Wagner arguments for situations in which a multipole group is spontaneously broken down to the trivial group, while in Section~\ref{sec:partial} we explore the more subtle case where a nontrivial subgroup is preserved.
+This analysis allows us to construct examples where a single symmetry-breaking pattern can be described by several distinct types of Goldstone modes, as well as an example where a continuous symmetry is spontaneously broken in one dimension at $T=0$.
+We then discuss generalized Imry-Ma arguments in the presence of quenched disorder in Section~\ref{sec:disord}.
+Finally, in Section~\ref{sec:example} we consider an explicit lattice model that illustrates some of the ideas of the previous sections.
+We conclude with a discussion of open questions in \ref{sec:disc}.
 
@@
 The multipole group is well-explained in Ref.~\cite{Gromov2019}.
-It generalizes abelian global symmetries such as $\phi(x) \goesto\phi (x) +c$ by allowing a shift by some set of polynomials, $\phi (x) \goesto \phi (x) + \lambda_\alpha P^\alpha(x)$.
+For concreteness, and since we will be interested in situations where a multipolar symmetry group is spontaneously broken, we will describe the multipole group in terms of its action on a compact scalar field $\phi$, which will play the role of a Goldstone boson for the broken multipolar symmetry.
+We will always imagine $\phi$ as constituting the phase mode for some microscopic order parameter $\psi \sim e^{i\phi}$, with $\psi$ transforming by various $U(1)$ phases under the multipole group.
+\ethan{are there any examples showing that this interpretation is too restrictive?}
+
+ The multipole group generalizes the internal shift symmetry $\phi(x) \goesto\phi (x) +c$ by allowing a shift by some set of polynomials involving the spatial coordinates, viz.
+$\phi (x) \goesto \phi (x) + \lambda_\alpha P^\alpha(x)$.
 The variables $\lambda_\alpha$ are symmetry parameters, while $\alpha$ labels the set of polynomials $P^\alpha(x)$.
@@
 
-An example of a multipole group that contains all translations and rotations but is not the maximal multipole group is the group generated by shifts of the form
+An example of a multipole group that contains all translations and rotations but is \emph{not} the maximal multipole group is the group generated by shifts of the form
 \begin{align}
@@
 
-We can now consider building field theories that are invariant under various multipole groups.
-Reference~\cite{Gromov2019} describes the process in detail, so we will only sketch the outline.
-References~\cite{Griffin2015, Gromov2019} discuss quantum field theories where configurations are weighted by an action.
+We can now consider building IR field theories for phases in which a multipolar symmetry is completely spontaneously broken, with no residual unbroken subgroup.
+Reference~\cite{Gromov2019} describes the process in detail, so in the following we will be somewhat succinct.
+We will continue to write things down in terms of the compact scalar $\phi$, which (at in the cases for which SSB is not preempted by strong fluctuations) is to be viewed as the Goldstone for the sponatneously broken symmetry.
+% References~\cite{Griffin2015, Gromov2019} discuss quantum field theories where configurations are weighted by an action.
 We will also want to consider thermal field theories with Hamiltonian descriptions, so we will abstractly discuss ``kinetic terms" in this section.
 
-The key step is to be able to find spatial derivatives $D$ that annihilate all the polynomials $P_a^{I_a}$ that we want to include in our multipole group.
-We say that $D$ is order-$s$ if it takes the form
+Since we are only interested in the low-energy physics of the putative symmetry-broken phase, constructing an appropriate IR field theory amounts to nothing more than constructing an appropriate kinetic term for $\phi$.
+To find a kinetic term invariant under a given multipole group $\mathcal{M}$, we need to find operators $D$ built out of spatial derivatives that annihilate all the polynomials in $\mathcal{M}$.
+If $a_\text{max}$ is the highest degree of the polynomials in $\mathcal{M}$, the simplest derivative operators which generically do the job are of the form \ethan{may or may not want to switch to the composite index notation $q^{I_{a_{\rm max}+1}} \p_{I_{a_{\rm max}+1}}$ employed later}
 \begin{align}
-D = q^i\partial_i + q^{ij}\partial_1\partial_j + \cdots + q^{i_1\dots i_s}\partial_{i_1}\dots\partial_{i_s},
+D = q^{i_1\dots i_{a_\text{max}+1}}\partial_{i_1}\dots \partial_{i_{a_\text{max}+1}},
 \end{align}
-where each $q$ is a symmetric tensor.
-Although it is not generically possible~\cite{Gromov2019}, we can sometimes find a set of $D_\alpha$ ($\alpha$ is an abstract index) with $s\le a_{\text{max}}$, where $a_\text{max}$ is the highest degree of the polynomials.
+for some symmetric tensor $q^{i_1\dots i_{a_\text{max}+1}}$.
+Although it is not generically possible to do so~\cite{Gromov2019}, we can sometimes find a set of $D_\alpha$ ($\alpha$ is an abstract index) with $s\le a_{\text{max}}$, where $a_\text{max}$ is the highest degree of the polynomials.
 In this case, the effective field theory will be invariant under some non-maximal multipole symmetry.
+% We say that $D$ is order-$s$ if the maximum number of derivatives appearing in $D$ is $s$.
 
-Generically, we cannot find any $D$ with $s\le a_\text{max}$ such that $DP_a^{I_a} = 0$ for all the polynomials we want. In that case we can still write down a theory with derivatives 
+Writing the invariant derivative operators as $D_\alpha$, the most general kinetic term is ~\cite{Gromov2019} 
 \begin{align}
-D = q^{i_1\dots i_n}\partial_{i_1}\dots \partial_{i_n},
+K[\phi(x)] &= g_{\alpha\beta} (D_\alpha \phi) (D_\beta \phi)
 \end{align}
-with $n=a_\text{max}+1$. Then, the field theory will be invariant under the maximal multipole group of degree $a_\text{max}$.
-
-Armed with the invariant derivatives $D_\alpha$ ($\alpha$ is another abstract index), we can start to construct a kinetic term.
-The most general term, at lowest order in derivatives, would be~\cite{Gromov2019}
-\begin{align}
-K[\phi(x)] &= g_{\alpha\beta} (D_\alpha \phi) (D_\beta \phi).
-\end{align}
+for some symmetric tensor $g_{\alpha\beta}$.
 We will often write the Fourier transform of the kinetic term as $\phi_{-k} K_k \phi_k$.
@@
 
-\subsection{Generalized Mermin-Wagner argument}
+\section{Generalized Mermin-Wagner: full multipole breaking}\label{sec:full_breaking}
 
-We are now ready to describe the generalized Mermin-Wagner argument for an arbitrary maximal multipole group.
+We are now ready to describe the generalized Mermin-Wagner argument in the case where an arbitrary maximal multipole group is spontaneously broken down to the trivial subgroup.
 This already appears in Ref.~\cite{Griffin2015}, so we are simply reviewing it here in preparation for the more generic cases to follow.
-Here, we will discuss both classical and quantum systems.
+Here, we will discuss both thermal and $T=0$ systems.
 
-Consider a clean classical system with a spontaneously broken maximal multipole symmetry of degree $a=n-1$.
-Then the kinetic term is
+Consider first a clean classical system at $T>0$ with a spontaneously broken maximal multipole symmetry of degree $a$.
+Then the kinetic term is proportional to 
 \begin{align}
-K[\phi(x)] = (\partial_{i_1} \dots \partial_{i_n} \phi)^2, \label{eqn:adisp}
+K[\phi(x)] = (\partial_{i_1} \dots \partial_{i_{a+1}} \phi)^2. \label{eqn:adisp}
 \end{align} 
-where $\phi$ is the order parameter field. 
-Now assume we are in an ordered phase, but begin to nucleate domains of linear size $L$.
-Since the order parameter is continuous there will be a thick domain ``wall", in which the order parameter changes gradually from its value in the background phase to its value in the domain.
+Now consider nucleating domains of linear size $L$ in which the expectation value of the order parameter --- namely $\langle e^{i\phi}\rangle$ --- varies by an amount of order 1.
+Since the order parameter is continuous there will be a thick domain ``wall" over which the order parameter changes gradually from its value in the background phase to its value in the domain.
 The thickness will be of order $L$.
@@
 When $d\le2(a+1)$, the energy cost is bounded, so the entropy gain favors domain creation.
-Thus, ordered phases are unstable and spontaneous symmetry breaking cannot occur for $d\le2(a+1)$.
+Thus, ordered phases are unstable and SSB cannot occur for $d\le2(a+1)$.
 
-We can reproduce this argument by considering correlation functions.
-As a warm-up, consider the standard monopole case.
-Assume there is a monopole U(1) symmetry that is spontaneously broken, leading to a Goldstone boson $\phi(x)$.
-The correlation function,
+We can also reproduce this argument by considering correlation functions.
+As a warm-up, consider the standard case where a monopole U(1) symmetry is spontaneously broken, leading to a Goldstone boson $\phi(x)$.
+The correlation function
 \begin{align}
@@
 
-When we consider a multipole symmetry, the dispersion changes to $K_k=k^{2(a+1)}$. Now the correlation functions will scale as 
+When we consider a (maximal) multipole symmetry, the dispersion changes to $K_k=k^{2(a+1)}$. Now the correlation functions will scale as 
 \begin{align}
@@
 As before, we interpret the divergence of the correlation function as a sign that the symmetry cannot be spontaneously broken.
-This allows us to reproduce our scaling argument that spontaneous symmetry breaking of the multipole group $\mathcal{M}^a_\max$ cannot occur for $d\le2(a+1)$.
+This allows us to reproduce our scaling argument that SSB of the multipole group $\mathcal{M}^a_\max$ cannot occur for $d\le2(a+1)$.
 
-If we look at the quantum (zero temperature) setting, the energy scaling no longer tells us where the critical dimension is.
+If we look at the zero temperature setting, the energy scaling no longer tells us where the critical dimension is.
 This is because even when there is no energy cost to forming domains, there is no entropy gain.
 Instead, quantum fluctuations must be the motivation for domain nucleation.
-To find the quantum critical dimension, we calculate the correlation function,
+To find the quantum critical dimension, we calculate the correlation function by way of the (imaginary-time) IR Lagrangian
 \begin{align}
-C(x) = \left\langle \phi(x) \phi(0) \right\rangle &\sim \int \frac{d\omega d^dk}{\omega^2 + k^{2( a + 1 )}},
+     \mathcal{L}[\phi] = (\partial_\tau\phi)^2 + K[\phi].
+\end{align} 
+This gives 
+\begin{align}
+\label{tzerocorr}C(x) = \left\langle \phi(x) \phi(0) \right\rangle &\sim \int \frac{d\omega d^dk}{\omega^2 + k^{2( a + 1 )}},
 \end{align}
 which now diverges at $d\le a+1$~\cite{Griffin2015}, where $d$ is the number of spatial dimensions.
-We have halved the critical dimension, so that spontaneous symmetry breaking cannot occur at $d\le a+1$.
-It is interesting to note that the dynamical critical exponent in these theories is $z = a+1$, so the classical and quantum critical dimensions are related by $d_\text{cl} = d_\text{q} + z$.
-
-\charlie{Is it more useful to say $d_\text{cl} = d_\text{q} + z$ or $d_\text{cl} = 2 d_\text{q}$, or not try to make any general statement at all?}
+We have halved the critical dimension, so that SSB cannot occur at $d\le a+1$.
+Note that the dynamical critical exponent in these theories is $z = a+1$, so that as expected, the classical and quantum critical dimensions are related by $d_\text{cl} = d_\text{q} + z$.
 
@@
 The previous analysis suggests a procedure for finding the critical dimension in clean anisotropic systems for breaking the multipole symmetry to the trivial group.
-Sort each dimension by the degree of its dispersion relation, so that there are $d_n$ dimensions with dispersion $k^{2n}$.
-Define the effective dimension as
+First, sort each dimension by the degree of its dispersion relation, so that there are $d_n$ dimensions with dispersion $k^{2n}$.
+Then define the effective dimension as
 \begin{align}
@@
 \end{align}
-In a classical system, spontaneous symmetry breaking cannot occur if $d_\eff\le 2$, while in a quantum system the critical dimension is $d_\eff = 1$.
+In a classical system, we then conclude that SSB cannot occur if $d_\eff\le 2$, while in a quantum system the critical dimension is $d_\eff = 1$.
 
-We should note that the statements in this section were framed in terms of the dispersion of the Goldstone modes rather than the structure of the symmetry group.
+We should emphasize that the statements in this subsection were framed in terms of the dispersion of the Goldstone modes rather than the structure of the symmetry group.
 It is always possible to find the dispersion given the multipole symmetry group, but it may require some basis changes (as in the Haah multipole group).
@@
 
-\subsection{Breaking to a subgroup} \label{sub:subgroup}
+In the previous section we studied situations in which a multipole group was fully broken.
+In this section we turn our attention to examining what happens when the symmetry breaking is incomplete, with a nontrivial subgroup remaining unbroken.
+A simple example that illustrates why this problem is nontrivial is the following.
 
-In the previous section we assumed we had an order parameter field that fully broke the multipole symmetry.
-Let us now consider the case where the order parameter breaks the symmetry from the original group $G$ to some subgroup $H$.
-For simplicity, we will let $G$ be the maximal multipole group $\mathcal{M}^{a}_\text{max}$ and $H$ be $\mathcal{M}^b_\text{max}$, with $b<a$.
+Consider a situation in which a dipolar symmetry $\mcm^1_{\rm max}$ is spontaneously broken to its monopolar subgroup $\mcm^0$, and momentarily specialize to the case of $d=3$ and $T>0$.
+In the spirit of Goldstone's theorem, the most naive thing to do when analyzing the symmetry broken phase would be to write down an action involving a set of $d=3$ Goldstone fields $\theta_j$, with $\t_j$ transforming linearly under the dipolar part of the symmetry group, but remaining invariant under the unbroken monopole subgroup, so that $\theta_j \mapsto \theta_j + \beta_j$ under a transformation parameterized by $P(x) = \alpha + \beta_j x^j$.
+In this case, the IR field theory for the putative symmetry-breaking phase would be of the form
+\begin{align}
+\label{vector} \mathcal{L}= \sum_j  (\partial_\tau \theta_j)^2 + \sum_{i,j} g_{ij} (\partial_i \theta_j)^2.
+\end{align}
+As we are working in $d=3$ and at $T>0$, the theory \eqref{vector} indeed has SSB for the dipole symmetry, with $e^{i\t_j}$ developing long-range order.
 
-The simplest case is when $a=1$ and $b=0$.
-In physical terms, we are taking a system that conserves total charge and total dipole moment, then condensing objects that carry a dipole moment but no total charge.
-\note{If $\phi(x)$ is charged under the monopole symmetry, then the operator $\partial_i \phi(x)$ is charged under the dipole part of the symmetry group.
-\charlie{Do we want correlation functions of $\partial_i\phi$ or $\exp(\partial_i\phi)$?} In a classical system at $T>0$ the correlation functions of this operator are
+The theory \eqref{vector} is not the only possibility consistent with this pattern of symmetry breaking, however. Indeed, consider instead the scalar field theory 
 \begin{align}
-\left\langle \partial_i\phi(x) \partial_j\phi(0) \right\rangle &\sim \delta_{ij} \int \frac{d^dk\, k^2}{k^4},
-\end{align}
-so that the critical dimension is $d=2$, as in the standard Mermin-Wagner theorem. }
+\label{singlescalar}    \mathcal{L} = (\partial_\tau\phi)^2 + g_{ij} (\partial_i\partial_j\phi)^2, \end{align}
+where $\phi$ transforms under the symmetry as $\phi \mapsto \phi + \alpha + \beta_jx^j$. 
+ In the scalar theory \eqref{singlescalar}, the analysis of section \ref{sec:full_breaking} shows that the monopole symmetry is {\it not} spontaneously broken, with $e^{i\phi}$ possessing short-ranged correlation functions.
+However, the dipolar symmetry does indeed have SSB, and it is easy to check that the operators $e^{i\p_j\phi}$ --- which transform under the symmetry in the same way as $e^{i\t_j}$ --- have long-range order.
+Thus in this case, the two theories \eqref{vector} and \eqref{singlescalar} exhibit the same pattern of symmetry breaking, despite having a different number of Goldstone modes.\footnote{We will refer to the mode $\phi$ in the theory \eqref{singlescalar} as the ``Goldstone mode'' of the spontaneously broken dipole symmetry, since it is the only mode in the theory.
+Note however that it is $\p_j\phi$ (and not $\phi$ itself) which transforms under the symmetry action in the way expected of Goldstone bosons.} In order to formulate the correct version of the Mermin-Wagner theorem in the partial breaking case, it will be necessary to understand why this is so.
+
+\subsection{Goldstone counting} \label{sub:subgroup}
+
+That the number of Goldstone modes is not uniquely determined by the pattern of symmetry breaking is of course not in contradiction with Goldstone's theorem, as illustrated e.g.
+by the example of the ferromagnet vs the antiferromagnet.
+Indeed, correctly determining the number of Goldstone modes realized in a given situation depends not just on the symmetry breaking pattern, but also on information relating to how the various charge generators act on the ground state, as nicely described in \cite{watanabe2013redundancies}.
+
+A general formal framework for understanding the counting of Goldstone modes in the present context can be developed by following the analysis of \cite{watanabe2013redundancies}, but here we will simply content ourselves with a heuristic physical argument.
+For simplicity, we will focus on systems in which the total symmetry group $\mcm_{\rm max}^a$ is spontaneously broken down to some nontrivial subgroup.
+
+Let us define the compressibilities $\kappa_c$ by 
+\be \label{kappadef} \kappa_c^{I_c} = \frac{d\langle n_c^{I_c}\rangle}{d\mu_c^{I_c}},\ee 
+where $n_c$ is the number density of $c$-pole charges, $\mu_c$ is a conjugate chemical potential, and $I_c = \{i_1,\dots,i_c\}$ is a composite index.
+
+Let $b\leq a$ denote the smallest degree of multipole charge with nonzero compressibility, i.e.
+consider a situation in which $\kappa_d>0$ for all $b\leq d\leq a$, while $\kappa_d=0$ for all $d<b$.
+Then the minimal IR Lagrangian for the symmetry broken phase looks schematically like
+\be \label{generall} \mathcal{L} = \kappa_b (\partial_\tau \phi_{I_b})^2 + g_{J_{a-b+1}}(\partial_{J_{a-b+1}}\phi_{I_b})^2,\ee 
+where 
+%$I_b = \{i_1,\dots,i_b\}$ and $J_{a-b+1}=\{j_1,\dots,j_{a-b+1}\}$ are composite indices with 
+$\p_{J_{a-b+1}} \equiv \p_{j_1} \cdots \p_{j_{a-b+1}}$, and where $\phi_{I_b}$ is a field that shifts by a polynomial of degree $a-b$ under the action of $\mmax a$.
+The compressibility appears as the coefficient of the time derivative term since $\p_\tau \phi_{I_b}$ is the momentum conjugate to $\phi_{I_b}$ in the quantum theory, and hence correlation functions of fluctuations in $n_b$ are determined by those of $\p_\tau \phi_{I_b}$.
+
+A necessary condition for $\mmax b$ to be spontaneously broken is that $\kappa_b \neq 0$.
+Indeed, if $\mmax b$ is spontaneously broken then $n_b$ creates Goldstones when acting on the ground state, thereby ensuring that the density-density correlator appearing in the calculation of \eqref{kappadef} must be nonzero at long wavelengths.
+\ethan{is this too laconic?}
+
+However, a nonzero $\kappa_b$ does not {\it necessarily} imply that $\mmax b$ is spontaneously broken.
+$\kappa_b\neq 0$ means that the system possesses gapless modes created by $n_b$, but it is possible for fluctuations to be strong enough so that these modes cannot be associated with the Goldstone modes of a spontaneously broken $\mmax b$.
+Thus the theory \eqref{generall} does not necessarily describe a phase where $\mmax b$ is spontaneously broken, and the exact pattern of symmetry breaking depends on the spatial dimension and whether or not $T$ is nonzero.
+For example, we can again consider the theory in \eqref{singlescalar} at $d=3,T>0$: even though the charge compressibility $\kappa_0$ (the coefficient of $(\p_\tau\phi)^2$ in \eqref{singlescalar}) is nonzero, $\phi$ fluctuates strongly enough that the monopole symmetry is not spontaneously broken.
+
+\subsection{Generalized Mermin-Wagner: partial multipole breaking}
+
+In the previous subsection, we saw that the number of Goldstone modes in the symmetry-breaking theory is determined by the smallest degree of multipole charge $b$ such that $\kappa_b\neq0$.
+We now turn to determining the symmetry breaking pattern that occurs for a given choice of $b,d,$ and $T$, thereby allowing us to formulate a generalized version of the Mermin-Wagner theorem for the case of partial multipole breaking.
+
+Since $\kappa_d=0$ if $d<b$, none of the subgroups $\mmax {d<b}$ can be spontaneously broken.
+Suppose then that the symmetry is spontaneously broken down to $\mmax {c}$, where $c\geq b$.
+To find the allowed values of $c$ consistent with this assumption, we need to calculate correlation functions of operators of the form $\exp(i\p_{J_{c-b}}\phi_{I_b})$, these being the operators transforming nontrivially under $\mmax c$ with the longest-range correlation functions.
+We have
+\be\ba \langle \p_{J_{c-b}}\phi_{I_b}(x) \p_{J_{c-b}}\phi_{I_b}(0)\rangle & \sim \int \frac{d\omega \, dk\, k^{d-1+2(c-b)}}{\omega^2 + k^{2(a-b+1)}} \\ 
+& \sim \int dk\, k^{d-1+2(c-b) - \varsigma_T(a-b+1)}, \ea\ee 
+where we have defined the function $\varsigma_T$ such that $\varsigma_T = 1$ if $T=0$ and $\varsigma_T = 2$ if $T>0$.
+Consequently, we find that $\mmax c$ can be spontaneously broken only in dimensions $d$ such that
+\be \label{pbreak_mw} d>2(b-c) + \varsigma_T(a-b+1).\ee 
+For a given choice of symmetry breaking pattern (captured by $a$ and $c$) and a given choice of which compressibilities are nonzero (captured by $b$), the quantity on the RHS determines the lower critical dimension for symmetry breaking, thereby generalizing the Mermin-Wagner theorem to the partial breaking case.
+As a sanity check on \eqref{pbreak_mw}, note that when we set $c=b=0$ (full multipole breaking) we recover the lower critical dimension of $\varsigma_T(a+1)$ obtained in the previous section.
+Also note that as expected, the lower critical dimensions in the quantum $(\varsigma_T = 1)$ and classical $(\varsigma_T = 2)$ cases are related by $d_{\rm cl} = d_{\rm q} + z$, where $z=a-b+1$ is the dynamical exponent of the Goldstone theory \eqref{generall}.
+
+Finally, as an interesting example, consider the case where the dipolar group $(a=1)$ is broken down to the monopole group $(b=0,c=1)$.
+At $T=0$, we see from \eqref{pbreak_mw} that this is possible provided that $d>0$.
+Thus the partial breaking of dipole symmetry provides us with an example where a continuous symmetry can be spontaneously broken in $d=1$ at zero temperature.
+
+%In a classical system at $T>0$ the correlation functions of this operator are
+%\begin{align}
+%\left\langle \partial_i\phi(x) \partial_j\phi(0) \right\rangle &\sim \delta_{ij} \int \frac{d^dk\, k^2}{k^4},
+%\end{align}
+%so that the critical dimension is $d=2$, as in the standard Mermin-Wagner theorem for monopole symmetries.
+This allows us to interpret the dipole charges of the monopole symmetry as monopole charges of the dipole part of the symmetry group.
+%However, this interpretation does not hold for  quantum systems. 
+%The correlation function scales as
+%\begin{align}
+%\left\langle \partial_i\phi(x) \partial_j\phi(0) \right\rangle &\sim \delta_{ij} \int \frac{d\omega\, d^dk\, k^2}{\omega^2 + k^4}\nn
+%&= \delta_{ij} \int \frac{d^dk\, k^2}{k^2},
+%\end{align}
+%which is always IR-finite. We thus conclude that at $T=0$, $\mcm^1_{\rm max}$ can be spontaneously broken down to $\mcm^0$ in any dimension
 
@@
 We should then expect symmetry breaking cannot occur for $d\le 2$ classically or $d\le 1$ quantum mechanically.
-
-It is interesting to note that condensing an object that carries dipole charge must break rotation symmetry.
-In Sec.~\ref{sec:example} we show how to break the dipole symmetry while preserving a lattice rotation symmetry by considering multiple species of dipole.
-We are not aware of a way to condense a dipole while preserving continuous rotation symmetry.
-
-\note{In more generality, for this pattern of symmetry breaking there will be a charged operator $\partial_{i_1} \dots \partial{i_m} \phi$, where $m=b+1$.
+%
+%
+%In more generality, for $G = \mathcal{M}^{a}_\text{max}$ and $H = \mathcal{M}^b_\text{max}$ there will be a charged operator $\partial_{i_1} \dots \partial_{i_{b+1}}\phi$ with a dispersion as in Eqn.~\ref{eqn:adisp}.
 Because $G$ is a maximal multipole group, we can still write the polynomials as in Eqn.~\ref{eqn:basis}.
+Recall that $P_c^{I_c}$ is a degree-$c$ monomial.
 As long as $\phi$ transforms as
-\begin{align}
-\delta \phi = \lambda_{I_c} P_c^{I_c}, \label{eqn:partial}
-\end{align}
-then $\partial_{i_1} \dots \partial_{i_m} \phi$ is invariant under $H=\mathcal{M}^b_\max$. }
-
-\note{Classical correlation functions for these operators take the form
-\begin{align}
-C &= \left\langle \partial_{i_1} \dots \partial_{i_m} \phi(x) \partial_{j_1} \dots \partial_{j_m} \phi(0) \right\rangle \nn
-&\sim \int \frac{d^dk\, k^{2m}}{k^{2n}},
-\end{align}
-and diverge when $d \le 2(n-m) = 2(a-b)\equiv d_\text{cl}$. On the other hand, the $T=0$ correlation functions behave as
-\begin{align}
-C &= \left\langle \partial_{i_1} \dots \partial_{i_m} \phi(x) \partial_{j_1} \dots \partial_{j_m} \phi(0) \right\rangle \nn
-&\sim \int \frac{d\omega\, d^dk\, k^{2m}}{\omega^2 + k^{2n}}\nn
-&\sim \int \frac{d^dk\, k^{2m}}{k^n},
-\end{align}
-so the quantum critical dimension is $d_\text{q} = n - 2m = a - 2b - 1$.
+%\begin{align}
+%\delta \phi = \lambda_{I_c} P_c^{I_c}, \label{eqn:partial}
+%\end{align}
+%then $\partial_{i_1} \dots \partial_{i_{b+1}} \phi$ is invariant under $H=\mathcal{M}^b_\max$. 
+%
+%Classical correlation functions for these operators take the form
+%\begin{align}
+%C &= \left\langle \partial_{i_1} \dots \partial_{i_{b+1}} \phi(x) \partial_{j_1} \dots \partial_{j_{b+1}} \phi(0) \right\rangle \nn
+%&\sim \int \frac{d^dk\, k^{2(b+1)}}{k^{2(a+1)}},
+%\end{align}
+%and diverge when $d \le 2(a-b)\equiv d_\text{cl}$. On the other hand, the $T=0$ correlation functions behave as
+%\begin{align}
+%C &= \left\langle \partial_{i_1} \dots \partial_{i_{b+1}} \phi(x) \partial_{j_1} \dots \partial_{j_{b+1}} \phi(0) \right\rangle \nn
+%&\sim \int \frac{d\omega\, d^dk\, k^{2({b+1})}}{\omega^2 + k^{2(a+1)}}\nn
+%&\sim \int \frac{d^dk\, k^{2(b+1)}}{k^{a+1}},
+%\end{align}
+%so the quantum critical dimension is $d_\text{q} = a - 2b - 1$.
 We recover the maximal generalized Mermin-Wagner argument if we let $b=-1$ denote the trivial group ($b=0$ denotes the monopole group).
 Furthermore, the critical dimensions again satisfy $d_\cl = d_\text{q} + z$.
-\charlie{Distinguish $z$ and $z_\max$?}
-}
-
-% The kinetic term should take the form
-% \begin{align}
-% K[\phi(x)] &= (\partial_{i_{m+1}}\dots \partial_{i_n} \phi_{i_1 \dots i_m})^2,
-% \end{align}
-% which is invariant under shifts of the form of Eq.~\ref{eqn:partial}.
+%
+%
+%\subsection{General formulation of Mermin-Wagner}
+%% The kinetic term should take the form
+%% \begin{align}
+%% K[\phi(x)] &= (\partial_{i_{m+1}}\dots \partial_{i_n} \phi_{i_1 \dots i_m})^2,
+%% \end{align}
+%% which is invariant under shifts of the form of Eq.~\ref{eqn:partial}.
 The dispersion for the Goldstone modes is $K_k \sim k^{2(n-m)} = k^{2(a-b)}$.
 Following the correlation function calculation above, we find a classical critical dimension of $d = 2(a-b)$ and a quantum critical dimension of $d = a-b$.
+%
+%\subsection{\note{Different phases for the same subgroup}} \label{sub:phases}
+%
+%The previous subsection shows that when the kinetic term in the effective action is $K\sim (\partial_i\partial_j\phi)^2$, there are intermediate dimensions ($d=2$ for $T>0$ or $d=1,2$ for $T=0$) where the symmetry breaking pattern preserves the monopole group.
+For smaller dimensions there is no symmetry breaking, while for larger dimensions (with this effective action) the group is fully broken.
+%We can also write down theories where the kinetic term will be $(\partial_i\theta_j)^2$.
+The latter type of theory will preserve the monopole group in any large enough dimension.
+Since the latter theory is quadratic, it behaves like an ordinary theory, with a MW theorem prohibiting symmetry breaking for $d\le 2$ at $T>0$ or $d\le 1$ at zero temperature.
+%
+%We can organize these different theories that have the same SSB structure by their compressibilities.
+Since we are considering theories that conserve the dipole group, there will be dipole charges and monopole charges.
+Call the compressibility for the monopole charges $\mu_0$ and the compressibility for the dipole charges $\mu_1$.
+In any number of dimensions, there will be a Mott-insulator-like phase where $\mu_0=\mu_1  = 0$.
+The kinetic term $(\partial_i\theta_j)^2$ assumes $\mu_0=0$ and $\mu_1>0$, while the term $(\partial_i \partial_j \phi)^2$ assumes both compressibilies are nonzero.
+%
+%The list of compressibilies only tells us which fields are available for writing down an effective action.
+As in the previous subsection, which operators develop long-range order depends on the dimension.
+The calculations are analogous to previous correlation functions, and the results are as follows.
+At nonzero temperature, the order $b$ of the preserved subgroup $\mmax{b}$ is
+%\begin{align}
+%\begin{tabular}{l|r|r|r}
+%     & $d=1,2$ & $d=3,4$, & $d>4$ \\
+%    $(\partial_i\partial_j\phi)^2$ &  1 &  0 & -1 \\
+%    $(\partial_i\theta_j)^2$       &  1 &  0 &  0 \\
+%    Mott                           &  1 &  1 &  1
+%\end{tabular}
+%\end{align}
+%where we are, as always, denoting the trivial group by $b=-1$.
+Because the correlation calculations do not have any $\omega$ integral, the critical dimensions for a given symmetry-breaking pattern are the same in different theories.
 
-\subsection{No mediated condensation} \label{sub:mediated}
+%The same is not true at $T=0$. Similar correlation function calculations show that the symmetry-breaking patterns are 
+%\begin{align}
+%\begin{tabular}{l|r|r|r}
+%     & $d=1$ & $d=2$, & $d>2$ \\
+%    $(\partial_i\partial_j\phi)^2$ &  0 &  0 & -1 \\
+%    $(\partial_i\theta_j)^2$       &  1 &  0 &  0 \\
+%    Mott                           &  1 &  1 &  1
+%\end{tabular} \label{eqn:zerophases}
+%\end{align}
+%The interesting difference from finite temperature is that the theory with a $\phi$ field can preserve $\mmax{0}$ at $d=1$, while the theory with a $\theta_j$ field cannot.
+%Both the $T>0$ and the $T=0$ table can easily be generalized to higher values of $a$, the order of the preserved multipole group.
 
-We have seen that the generalized Mermin-Wagner argument provides critical dimensions for partial breaking that are lower than the critical dimensions for full symmetry breaking.
-\note{Can we use this behavior to try to evade the Mermin-Wagner argument for the full symmetry group?}
+%Whenever there are distinct theories with the same symmetry-breaking pattern, some of the theories have operators with quasi-long-range order.
+\todo{QLRO vs SRO} For example, the $(\partial_i\theta_j)^2$  theory at $d=1$ acts like a Luttinger liquid of dipoles.
+A full analysis of quasi-long-range ordered fields is beyond the scope of this paper.
+Furthermore, we have only considered free fixed points.
+There certainly may be interacting theories with distinct phases and symmetry-breaking patterns~\cite{TonerRadzihovsky}.
+\ethan{how do you have interacting goldstones for an abelian symmetry?}
 
-%\remove{In fact, we can use this observation to evade the generalized Mermin-Wagner theorem, completely breaking any multipole symmetry for $d>2$ classically, or $d>1$ quantum mechanically}.
-This possibility has been pointed out in Refs.~\cite{Griffin2015Cascade, Griffin2015}, but we include a further interpretation in terms of what types of phase transitions are allowed in different dimensions.
-For simplicity, we will begin by considering the classical dipole-preserving theory with symmetry group $\mathcal{M}^2_\max$.
+% ------------------------
 
-To be explicit, consider a monopole field $\phi(x)$ and a dipole field $\varphi_i(x)$. Under the dipole symmetry they transform as 
-\begin{align}
-\phi(x) &\goesto \phi(x) + a_ix^i + b,\nn
-\varphi_i(x) &\goesto \varphi_i(x) + a_i.
-\end{align}
-The leading terms in the Lagrangian will be
-\begin{align}
-\mathcal{L} &= (\partial_t \phi)^2 + (\partial_t \varphi_i)^2 + (\partial_i \partial_j \phi)^2 + (\partial_i \varphi_j)^2,
-\end{align}
-so that $\phi$ has a $k^4$ dispersion and $\varphi_i$ has a $k^2$ dispersion.
-We can also include an interaction term of the form $g(\partial_i \phi - \varphi_i)^2$, which does not break the symmetry.
-The $\varphi_i$ field is well-defined whenever the dipole symmetry is broken.
+% Once we have identified a symmetry group $G$ and its preserved subgroup $H$ we still have not fully specified a phase of matter.
+As an example, consider the previous case of partially breaking the maximal dipole group $\mathcal{M}^1_\max$ to the monopole group $\mathcal{M}^0_\max$.
+There is one theory with Lagrangian
+% \begin{align}
+% \mathcal{L}_2 = (\partial_t \phi)^2 + \sum_{ij} (\partial_i \partial_j \phi)^2,
+% \end{align}
+% which has critical dynamical exponent $z=2$. There is another theory with Lagrangian
+% \begin{align}
+% \mathcal{L}_1 &= (\partial_t \varphi_i)^2  + \sum_{ij} (\partial_i \varphi_j)^2,
+% \end{align}
+% and $z=1$. The two theories differ in that, if $\phi$ transforms as $\phi \goesto \phi + a_ix^i + b$, then $\varphi_i$ transforms as $\varphi_i \goesto \varphi_i + a_i$
 
-\note{We can then redefine the $\varphi_i$ field to $\tilde{\varphi}_i = \varphi_i - \partial_i \phi$, so that it is neutral under the symmetry.
-The interaction term becomes $g\tilde{\varphi}_i^2$, so that the $\tilde{\varphi}_i$ field becomes massive.
-The only remaining massless field is $\phi$, which remains charged under the full dipole group and keeps its $k^4$ dispersion.
-As a result, the critical dimension for breaking the full dipole group is not affected by the possibility of partially breaking.
-}
-\charlie{What else to say here? Compare to Griffin/Grosvenor?}
+% \charlie{Does this mean the phase transition (if it exists) has to be topological,  \ethan{in what sense?} in the sense that the symmetry group and preserved subgroups are the same/ beyond Landau?}
 
-%As long as $\phi_i$ is well-defined, integrating out fluctuations in $\phi_i$ endows the monopole field $\phi$ with a $k^2$ dispersion.
-The coupling $g$ is set by how far the theory sits from the dipole critical point.
+% For both theories, the classical critical dimension is $d_\cl = 2$.
+However, they have different quantum critical dimensions, $d_{\q, 1} = 1$ and $d_{\q,2}=0$.
+Both theories obey $d_\cl = d_\q + z$.
+Of course, for $d>2$ the $z=2$ theory develops long-range order for $e^{i\phi(x)}$, breaking the monopole symmetry in addition to the dipole symmetry.
 
-%With a $k^2$ dispersion for the monopole field, we now have a route to evade the (generalized) Mermin-Wagner argument and fully break the dipole symmetry in dimensions $d> 2$.
-First, partially break the symmetry so that $\phi_i$ is well defined, which is allowed for any $d>2$.
-Then, we can use the effective dispersion of the $\phi$ field to break the remaining part of the symmetry, also in any $d>2$.
-This procedure can be generalized, so that any multipole symmetry can be fully broken in $d>2$, provided we have enough symmetry-breaking order parameters.
-The discussion also translates to the quantum setting, so that the ultimate critical (space) dimension is always $d=1$.
+% We can now generalize to arbitrary $a$ and $b$. To do so, we define a field $\phi_{i_1\dots i_k}$ which transforms under $\mathcal{M}^a_\max$ as 
+% \begin{align}
+% \delta \phi_{i_1\dots i_k} = \lambda_{I_c} \partial_{i_1} \dots \partial_{i_k} P_c^{I_c}.
+% \end{align}
+% The kinetic term for this field will be
+% \begin{align}
+% K \sim (\partial_{i_{k+1}} \dots \partial_{i_{a+1}} \phi_{i_1\dots i_k})^2, \label{eqn:atok}
+% \end{align}
+% so that $z = 1+a-k$.
+Although $\phi_{i_1\dots i_k}$ is invariant under $\mathcal{M}^k_\max$, that field may not be able to develop long-range order.
+The fields that do develop LRO may be invariant under larger groups.
+In particular, $\partial_{i_{k+1}} \dots \partial_{i_{b+1}} \phi_{i_1\dots i_k} $ is invariant under $\mathcal{M}^b_\max$
 
-%The preceding procedure gives us insight into what the generalized Mermin-Wagner arguments were telling us.
-When we wrote down those arguments, we did not include the intermediate fields.
-Thus, the generalized Mermin-Wagner arguments tell us when a direct transition between two phases is or is not allowed.
-Consider again the dipole group $\mathcal{M}^2_\max$.
-The phase in which the symmetry is fully broken can exist in any dimension $d>2$ ($d>1$ for quantum systems), but there can only be direct phase transitions between this phase and the wholly symmetry unbroken phase in $d>4$ ($d>2$).
-In dimensions $2<d<4$ ($1<d<2$) the fully symmetry broken and the fully symmetry unbroken phase must be separated by an intermediate phase that breaks dipole but not monopole symmetry.
+% At finite temperature, the field $\phi_{i_1\dots i_k}$ develops long-range order when $d>2(a-k)$, so that the preserved subgroup is $\mmax{k}$.
+When $2(a-b)<d\le 2(a-b+1)$, the preserved subgroup is $\mmax{b}$.
+The upshot is that there are $N^d_{ab}$ to spontaneously break the symmetry from $\mmax{a}$ to $\mmax{b}$ in $d$ spacial dimensions.
+The function is
+% \begin{align}
+% N^d_{ab} = \begin{cases} 0, \quad d\le 2(a-b) \\ 2 \\ 
+% \end{cases}
+% \end{align}
+
+% All of these theories will have $d_\cl = 2(a-b)$ and $d_\q = 2(a-b) - z$.
+
+% In addition to having different critical exponents $z$, these different theories with the same symmetry breaking patterns have different numbers of Goldstone modes.
+For the dipole examples considered above, the $z=1$ theory has $d$ massless modes (the components of $\varphi_i$) while the $z=2$ theory only has a single massless mode, $\phi$.
+\charlie{What else should we say about this?}
+
+% \todo{How should we address the question of stability?}
+% \ethan{I think my comments about stability are actually not relevant here --- they only imply that the transition between the ordered and disordered phases cannot be studied using by doing a perturbative analysis of the effects of vortex operators in the ordered phase}
 
@@
 Of course, strong enough disorder can always destabilize the ordered phase, so will not consider that case.
-Weak disorder can discourage the ordered phase and raise the critical dimension at which spontaneous symmetry breaking is impossible.
+Weak disorder can discourage the ordered phase and raise the critical dimension at which SSB is impossible.
 Theorems of this type originate with Imry and Ma~\cite{ImryMa} and were proved by Aizenman and Wehr~\cite{Aizenman}.
@@
 The Imry-Ma argument only appeals to energy considerations, so the quantum critical dimension in the presence of disorder remains the same as the classical critical dimension~\cite{Vojta2013}.
-In disordered quantum systems, spontaneous symmetry breaking is impossible for $d\le 4(a+1)$.
+In disordered quantum systems, SSB is impossible for $d\le 4(a+1)$.
 
 To reproduce this argument using a correlation function calculation~\cite{ImryMa} we need to calculate the response to disorder.
-Call the disorder field $\theta(x)$ and redefine $\phi(x)$ to be the variation away from the average value $\phi_0$.
+Call the disorder field $h(x)$ and redefine $\phi(x)$ to be the variation away from the average value $\phi_0$.
 The relevant part of the Hamiltonian is
 \begin{align}
-H &\sim \int d^dx \left[ \half K[ \phi(x)] - h \phi(x) \theta(x) \right] \nn
-&\sim \int d^dk \left[ \half \phi_{-k} K_k \phi_k -h \phi_{-k} \theta_k \right] \nn
-&\sim \int d^dk \left[ \half \phi_{-k}' K_k \phi_k' - \half h^2 \theta_{-k} K^{-1}_k \theta_k \right],
+H &\sim \int d^dx \left[ \half K[ \phi(x)] - \phi(x) h(x) \right] \nn
+&\sim \int d^dk \left[ \half \phi_{-k} K_k \phi_k - \phi_{-k} h_k \right] \nn
+&\sim \int d^dk \left[ \half \phi_{-k}' K_k \phi_k' - \half h_{-k} K^{-1}_k h_k \right],
 \end{align}
-where $h$ sets the strength of the disorder, and we have written $\phi_k' = \phi_{-k} - h K^{-1}_k \theta_{-k}$. Then, from the expectation of $\phi_k$,
+where we have written $\phi_k' = \phi_{-k} - K^{-1}_k h_{-k}$. Then, from the expectation of $\phi_k$,
 \begin{align}
-\langle \phi_k \rangle &\sim \frac{1}{h} \frac{\delta Z}{\delta \theta_{-k} } \nn
-&\sim h K_k^{-1} \theta_{k},
+\langle \phi_k \rangle &\sim \frac{\delta Z}{\delta h_{-k} } \nn
+&\sim K_k^{-1} h_{k},
 \end{align}
@@
 \begin{align}
-\langle \phi(0) \phi(x) \rangle &\sim \int d^dk \, K_k^{-2}  h^2 \langle \theta_{-k} \theta_{k} \rangle e^{ik\cdot x},
+\langle \phi(0) \phi(x) \rangle &\sim \int d^dk \, K_k^{-2} \langle h_{-k} h_{k} \rangle e^{ik\cdot x},
 \end{align}
-where $h^2 \langle \theta_{-k} \theta_{k} \rangle e^{ik\cdot x}$ does not affect the divergence at small $k$, assuming the disorder is short range correlated.
+where $\langle h_{-k} h_{k} \rangle e^{ik\cdot x}$ does not affect the divergence at small $k$, assuming the disorder is short range correlated.
 We can compare to Eqn.~\ref{eqn:correl} to see that in the presence of disorder that couples linearly to an order parameter fully breaking multipole symmetry, the critical dimension for a disordered system is twice the critical dimension for having direct full multipole symmetry breaking in a clean classical system.
@@
 A more general case is a system with symmetry $G = \mathcal{M}_\max^a$ and disorder that breaks the symmetry to $\mathcal{M}_\max^b$ but preserves $G$ on average.
-In this system the critical dimension should be $d=4(a-b)$ for both classical and quantum systems.
+In this system the critical dimension for long-range order of a field that breaks $\mmax{a}$ but preserves $\mmax{b}$ should be $d=4(a-b)$ for both classical and quantum systems.
 Again, let $b=-1$ denote the trivial group in order to recover our previous results.
@@
 Consider a $d$-dimensional hypercubic lattice with sites labeled by $j$ and directions labeled by $\mu = 1,\dots,d$.
-Let there be bosonic quantum degrees of freedom $n_j$ on the sites and $n'_{j,j+\mu}$ on the edges.
-Thus, we can say there are $d$ species of edge-type bosons.
+Let there be bosonic degrees of freedom annihilated by $b_j$ on the sites, and `dipolar' bosonic degrees of freedom annihilated by $d_{j,j+\mu}$ on the edges.
+We can thus think of there as being $d$ species of edge-type bosons.
 
@@
 \op{1}(\xi) &= \prod_je^{i\xi n_j}\nn
-\op{2}_\mu(\xi) &= \prod_j e^{i\xi (j_\mu n_j)}e^{i\xi n'_{j,j+\mu}},
+\op{2}_\mu(\xi) &= \prod_j e^{i\xi (j_\mu n_j)}e^{i\xi n^d_{j,j+\mu}},
 \end{align}
-where $j_\mu$ is the $\mu$-th component of the site label $j$.
+where $j_\mu$ is the $\mu$-th component of the site label $j$, $n$ is the site boson number operator, and $n^d$ is the edge boson number operator.
 The first operator corresponds to conservation of the total number of site bosons, while the $\mu$-th component of the second operator corresponds to the sum of the $\mu$-th component of the total dipole moment of site bosons and the total number of the $\mu$-th type of edge bosons.
@@
 
-The most relevant Hamiltonian obeying these symmetries is  
+The minimal Hamiltonian obeying these symmetries is  
 \begin{align}
@@
 &\qquad - \sum_{j,\mu,\nu} t_b \left[  b_j b^{\dag}_{j+\mu} b^{\dag}_{j+\nu} b_{j+\mu+\nu} + H.C. \right]\nn
-H_d &= \sum_{j,\mu} \left[ \frac{U_d}{2} n'_{j,j+\mu} (n'_{j,j+\mu} - 1) -\mu_d \, n'_{j, j + \mu} \right]\nn
+H_d &= \sum_{j,\mu} \left[ \frac{U_d}{2} n^d_{j,j+\mu} (n^d_{j,j+\mu} - 1) -\mu_d \, n^d_{j, j + \mu} \right]\nn
 &\qquad  - \sum_{j,\mu,\nu} t_d \left[d^{\dag}_{j, j + \mu} d_{j + \nu, j+\mu + \nu} +H.C.\right]\nn
@@
 where sums are taken over sites $j$ and lattice directions $\mu$ and $\nu$.  
-The operator $b^\dagger_j$ creates a boson on site $j$ while $d^\dagger_{j,j+\mu}$ creates an boson on the edge between $j$ and $j+\mu$.
+%The operator $b^\dagger_j$ creates a boson on site $j$ while $d^\dagger_{j,j+\mu}$ creates an boson on the edge between $j$ and $j+\mu$. 
 For simplicity, let us set 
@@
 so there is a robust insulating phase at weak hopping $t_b, t_d$~\cite{Fisheretal}.
+A detailed discussion of the phase diagram and phenomenology of this model will be given in~\cite{dbhm}; in the following we will simply discuss the symmetry breaking patterns realized in various different limits.
 
 The Hamiltonian $H_b$ controls condensation of the site bosons.
-When $t_b$ is small, the site bosons are in number eigenstates while when $t_b$ is large the site bosons condense.
+When $t_b$ is small, the site bosons are in number eigenstates while when $t_b$ is large the site bosons condense (although note that since the term proportional to $t_b$ is quartic in the site boson operators, the condensation transition may be nonstandard).
 Similarly, the edge bosons condense when $t_d$ is large.
-Note that all $d$ species of edge boson will condense equally, so that lattice rotation symmetry is not broken.
+Note that lattice rotation symmetry will be broken if only a subset of the edge bosons condense, while if all $d$ species of condense equally then lattice rotation symmetry is preserved.
+We are not aware of a way to condense objects with dipolar charge while preserving {\it continuous} rotation symmetry.
+\ethan{when doing mean field for the dipoles the potential may be such as to favor a case where only one species condenses.
+The quartic part of the potential is a complicated function that depends on the dimension and parent Mott insulator, so it's better to just discuss both possibilities}
+
 The term $H_\text{int}$ is allowed by the symmetry because it simultaneously removes an edge boson and creates a dipole of site bosons.
-When the edge bosons are condensed, $H_\text{int}$ gives the site bosons an effective single particle hopping, and hence an effective $k^2$ dispersion, which would otherwise be disallowed by symmetry.
-\todo{Yes to effective hopping, no to $k^2$ dispersion.
-Discuss SF/SC?}
+Note that when the edge bosons are condensed, $H_\text{int}$ gives the site bosons an effective single particle hopping: thus condensation of single bosons can either proceed directly (by making $t_b$ large) or indirectly (by first condensing dipoles, and making the effective single-particle hopping term large).
+%and hence an effective $k^2$ dispersion, which would otherwise be disallowed by symmetry.
+\charlie{What would the effective dispersion actually be?
+Does it make sense to say that it is $k^2$ at short distances and $k^4$ at long distance?}
 
-\todo{Redo discussion of phases. Reframe in terms of symmetry-breaking patterns, not phases.}
+%\charlie{Relevance of cosine terms and stability to crystal phase?}
+%\ethan{at large enough hopping strength we will generically just get a condensate and don't have to worry about this stuff}
 
-\todo{Define multiple dipole-breaking patterns at the effective level. Relevance of cosine terms and stability to crystal phase?}
-At sufficiently high dimension, this system will support three phases, which may be loosely described as the Mott insulator, the monopole superfluid where site bosons are condensed and the dipole superfluid where edge bosons (but not site bosons) are condensed.
-Note that the dipole superfluid does preserve lattice rotations.
-In the monopole superfluid phase, the effective Hamiltonian contains bare creation and annihilation terms for the edge bosons, which automatically breaks number conservation of the edge bosons, and effectively induces a dipole superfluid as well.
+Consider first the phase in which only the edge bosons have condensed, with the site bosons remaining gapped.
+In this case the dipole compressibility $\kappa_1$ is nonzero, while the charge compressibility vanishes, $\kappa_0=0$.
+The condensed phase is therefore described in the IR by the $z=1$ theory \eqref{vector}, and the dipolar symmetry is spontaneously broken down to the monopole subgroup provided that $d>1$ (for $T=0$) or $d>2$ (for $T>0$).
+
+Next, consider the phase where the site bosons have condensed.
+In this case both $\kappa_1,\kappa_0$ are nonzero, and deep in the condensed phase the IR physics is described by the $z=2$ theory \eqref{singlescalar}.
+The dipolar symmetry is spontaneously broken in $d>0$ if $T=0$ and in $d>2$ at $T>0$, while the monopole subgroup is spontaneously broken in $d>2$ at $T=0$ and $d>4$ at $T>0$.
+This model therefore provides us with a way of realizing symmetry-broken phases described by both of our earlier example theories \eqref{vector} and \eqref{singlescalar}.
+
+%The phases of this model will be those in Eqn.~\ref{eqn:zerophases}.
+The values of $t_b$ and $t_d$ will determine which fields are available for the IR action.
+When the site bosons are condensed the $\phi$ field exists but might not be long-range ordered.
+Similarly when the edge bosons are condensed the $\theta_j$ field exists but might not be long-range ordered.
+For $d<2$ any symmetry-breaking pattern is possible.
+For $0<d\le 2$ the symmetry can not be completely broken, but the dipole part can be broken.
+
+%At sufficiently high dimension, this system will support three phases, which may be loosely described as the Mott insulator, the monopole superfluid where site bosons are condensed and the dipole superfluid where edge bosons (but not site bosons) are condensed.
+%Note that the dipole superfluid does preserve lattice rotations.
+%In the monopole superfluid phase, the effective Hamiltonian contains bare creation and annihilation terms for the edge bosons, which automatically breaks number conservation of the edge bosons, and effectively induces a dipole superfluid as well.
 We can identify these phases with symmetry breaking patterns of the multipole group.
@@
 \end{align}
-where $h_b$ and $h_d$ control the magnitude of the disorder and each instance of $\sigma$ is a random phase. We will always consider $h_b,h_d<<1$.
+where $h_b$ and $h_d$ control the magnitude of the disorder and each instance of $\sigma$ is a random phase. We will always consider $h_b,h_d\ll1$.
 
-At first, we will set $h_d=0$ but turn on $h_b$.
+%At first, we will set $h_d=0$ but turn on $h_b$.
 This corresponds to the main considerations from Sec.~\ref{sec:disord}, where the disorder couples to the order parameter that fully breaks the symmetry.
-This raises the critical dimension for the superfluidity of the site bosons.
+This raises the critical dimension for the long-range order of the site bosons.
 The edge bosons may still enter a superfluid phase for $d>1$, but the site bosons may not condense with long range phase order until $d>4$.
@@
 
-Now, let $h_d\ne 0$ while keeping $h_b=0$.
+%Now, let $h_d\ne 0$ while keeping $h_b=0$.
 The Imry-Ma argument for the edge bosons says that they should not be able to enter a superfluid phase when $d\le 4$.
@@
 
-It is difficult to justify including including one disorder field but not the other.
+%It is difficult to justify including including one disorder field but not the other.
 The standard reason to forbid disorder is by imposing translation invariance, but if we are already including one disorder field we are necessarily breaking translation symmetry.
 For $h_b,h_d\ne 0$, we can fully rely on the Imry-Ma argument.
-No symmetry breaking can occur for $d\le 4$, the monopole superfluid must be separated from any symmetry-unbroken phase by a dipole superfluid for $4<d\le 8$, and any symmetry-breaking transition can occur for $d>8$.
+No symmetry breaking can occur for $d\le 4$, the dipole part of the symmetry may be broken for $4<d\le 8$, and any symmetry-breaking phase can occur for $d>8$.
 
@@
 
-In this paper we analyzed the spontaneous symmetry breaking of various multipole groups, and discussed generalized Mermin-Wagner theorems, which (we argued) should be best understood as constraints on when a direct transition fully breaking a multipole group can occur, and when a symmetry unbroken phase and a (monopole) symmetry breaking phase must be separated by intermediate phases breaking higher multipole symmetries.
+\charlie{Should we say anything about Ref.~\cite{Argurio2021}, which doesn't consider multipole symmetry, but does look at restricted mobility and crystal-like phases from abnormal kinetic terms?}
+
+In this paper we analyzed the spontaneous symmetry breaking of various multipole groups and discussed generalized Mermin-Wagner theorems, which we argued should be best understood as constraints on when a direct transition fully breaking a multipole group can occur, and when a symmetry unbroken phase and a (monopole) symmetry breaking phase must be separated by intermediate phases breaking higher multipole symmetries \ethan{is this still the best way of thinking about things?
+(also by ``direct transition'' I am assuming you mean ``direct continuous transition''?)}.
 We also considered multipole groups that are not the maximal multipole group, and the effect of quenched disorder.
@@
 Of course, we could consider further combinations of effects.
-We could spontaneously break a symmetry from a group $G$ to a subgroup $H$ where either $G$ or $H$ are non-maximal.
-Or we could consider non-maximal groups with disorder.
-While the number of potential examples to consider is unlimited, they should all be analyzable using the ideas introduced herein.
+For example, we could spontaneously break a symmetry from a group $G$ to a subgroup $H$, where one or both of $G,H$ are non-maximal; we could also consider non-maximal groups with disorder.
+While the number of potential examples to consider is large, they should all be analyzable using the ideas introduced herein.
 
@@
 These essentially amount to stability analyses about the Gaussian symmetry breaking fixed point.
-In principle, even when the Gaussian symmetry breaking fixed point is unstable, a non-trivial fixed point with long range order could arise (see e.g.
+\ethan{I don't really understand how these are ``stability'' arguments --- stability with respect to which perturbations?
+The free theories we wrote down are at least all stable in the RG sense (the core energies of any vortex configurations can be made arbitrarily large, which means that the Gaussian actions can describe the IR physics even when the fluctuations are strong enough to prevent the symmetries from being spontaneously broken).
+Maybe there's something with disorder that I'm missing?} In principle, even when the Gaussian symmetry breaking fixed point is unstable, a non-trivial fixed point with long range order could arise (see e.g.
 \cite{TonerRadzihovsky}).
@@
 Determining what kind of phase {\it can} obtain in the absence of long range ordered symmetry breaking is beyond the scope of this work, and would at a minimum require understanding the nature and role of topological defects in the symmetry breaking order parameter, akin to vortices in the XY model.
-The range of possible symmetry unbroken phases could be even richer in the presence of disorder, where various glassy phases could also come into play \cite{Fisheretal}.
+\ethan{I think that since all the symmetries we are considering are Abelian, we can always have QLRO in the dimensions where the Gaussian action allows it (we can always work in a limit where the vortex core energy is infinitely large, meaning that we can always find a regime in which the description in terms of a phase mode is legitimate).
+The case of $O(n)$ models with $n>2$ is different because the sigma model for the putative broken phase is non-Abelian, with the stiffness flowing to zero.
+In the present Abelian case there is no such flow, and vortices are the only thing standing in the way of a Gaussian description.
+But as mentioned above, they can be gotten rid of (in any dimension) by making their core energies large.} The range of possible symmetry unbroken phases could be even richer in the presence of disorder, where various glassy phases could also come into play \cite{Fisheretal}.
 Our discussion of disorder physics was also limited to quenched short-range correlated disorder.
@@
 the U.S. Department of Energy, Office of Science, Basic
-Energy Sciences, under Award \# DE-SC0021346
+Energy Sciences, under Award \# DE-SC0021346. EL was supported by the Hertz Fellowship.
 
@@
 
+\appendix
+
+\section{No mediated condensation} \label{app:mediated}
+
+\ethan{not sure if this appendix is necessary, given that the paper is about MW theorems?}
+
+In Sec.~\ref{sec:partial} we saw that the generalized Mermin-Wagner argument provides critical dimensions for partial breaking that are lower than the critical dimensions for full symmetry breaking.
+\note{Can we use this behavior to try to evade the Mermin-Wagner argument for the full symmetry group?}
+
+%\remove{In fact, we can use this observation to evade the generalized Mermin-Wagner theorem, completely breaking any multipole symmetry for $d>2$ classically, or $d>1$ quantum mechanically}.
+This possibility has been pointed out in Refs.~\cite{Griffin2015Cascade, Griffin2015}, but we include a further interpretation in terms of what types of phase transitions are allowed in different dimensions.
+For simplicity, we will begin by considering the classical dipole-preserving theory with symmetry group $\mathcal{M}^2_\max$.
+
+To be explicit, consider a monopole field $\phi(x)$ and a dipole field $\varphi_i(x)$. Under the dipole symmetry they transform as 
+\begin{align}
+\phi(x) &\goesto \phi(x) + a_ix^i + b,\nn
+\varphi_i(x) &\goesto \varphi_i(x) + a_i.
+\end{align}
+The leading terms in the Lagrangian will be
+\begin{align}
+\mathcal{L} &= (\partial_t \phi)^2 + (\partial_t \varphi_i)^2 + (\partial_i \partial_j \phi)^2 + (\partial_i \varphi_j)^2,
+\end{align}
+so that $\phi$ has a $k^4$ dispersion and $\varphi_i$ has a $k^2$ dispersion.
+We can also include an interaction term of the form $g(\partial_i \phi - \varphi_i)^2$, which does not break the symmetry.
+The $\varphi_i$ field is well-defined whenever the dipole symmetry is broken.
+
+\note{We can then redefine the $\varphi_i$ field to $\tilde{\varphi}_i = \varphi_i - \partial_i \phi$, so that it is neutral under the symmetry.
+The interaction term becomes $g\tilde{\varphi}_i^2$, so that the $\tilde{\varphi}_i$ field becomes massive.
+The only remaining massless field is $\phi$, which remains charged under the full dipole group and keeps its $k^4$ dispersion.
+As a result, the critical dimension for breaking the full dipole group is not affected by the possibility of partially breaking.
+}
+\charlie{What else to say here? Compare to Griffin/Grosvenor?}
+\todo{Compare to fracton superfluids}
+
+%As long as $\phi_i$ is well-defined, integrating out fluctuations in $\phi_i$ endows the monopole field $\phi$ with a $k^2$ dispersion.
+The coupling $g$ is set by how far the theory sits from the dipole critical point.
+
+%With a $k^2$ dispersion for the monopole field, we now have a route to evade the (generalized) Mermin-Wagner argument and fully break the dipole symmetry in dimensions $d> 2$.
+First, partially break the symmetry so that $\phi_i$ is well defined, which is allowed for any $d>2$.
+Then, we can use the effective dispersion of the $\phi$ field to break the remaining part of the symmetry, also in any $d>2$.
+This procedure can be generalized, so that any multipole symmetry can be fully broken in $d>2$, provided we have enough symmetry-breaking order parameters.
+The discussion also translates to the quantum setting, so that the ultimate critical (space) dimension is always $d=1$.
+
+%The preceding procedure gives us insight into what the generalized Mermin-Wagner arguments were telling us.
+When we wrote down those arguments, we did not include the intermediate fields.
+Thus, the generalized Mermin-Wagner arguments tell us when a direct transition between two phases is or is not allowed.
+Consider again the dipole group $\mathcal{M}^2_\max$.
+The phase in which the symmetry is fully broken can exist in any dimension $d>2$ ($d>1$ for quantum systems), but there can only be direct phase transitions between this phase and the wholly symmetry unbroken phase in $d>4$ ($d>2$).
+In dimensions $2<d<4$ ($1<d<2$) the fully symmetry broken and the fully symmetry unbroken phase must be separated by an intermediate phase that breaks dipole but not monopole symmetry.
+
 \end{document}

### 2021-12-10 multipoleSSB/charlies_edits.tex
charlies edits

--- 2021-12-08 multipoleSSB/ethans_edits.tex
@@
 The simplest case is the maximal multipole group $\mathcal{M}^a_\text{max}$, which includes all shifts by polynomials of degree $a$ or less.
-Individual polynomials can be written as
+Individual monomials can be written as
 \begin{align}
-P_c^{I_c} = \mu^{I_c}_{i_1\dots i_c}x^{i_1}\dots x^{i_c}, \label{eqn:basis}
+P_c^{\gamma} = \mu^{\gamma}_{I_c}x^{I_c}, \label{eqn:basis}
 \end{align}
-where each  matrix  $\mu^{I_c}_{i_1\dots i_c}$ is fully symmetric and $c\le a$. This group also includes all translations and rotations.
+where $\gamma$ is an index that runs over monomials of degree $c$ and $c\le a$.
+The composite index $I_c = \{i_1,\dots,i_c\}$  expands as $x^{I_c}=x^{i_1}\cdots x^{i_c}$, and similarly for $\partial_{I_c}$.
+Each  matrix  $\mu^{\gamma}$ is fully symmetric on its $I_c$ indices.
+This group also includes all translations and rotations.
 
@@
 \begin{align}
-\phi \goesto \phi + \lambda_0 P_0^0 + \lambda_{i} P^{i}_1 + \lambda_{I_2} P^{I_2}_2,
+\phi \goesto \phi + \lambda_0 P_0^0 + \lambda_{i} P^{i}_1 + \lambda_{\beta} P^{\beta}_2,
 \end{align}
@@
 \begin{align}
-P_1^{i} = x^i,\quad \quad P_2^{I_2} &= \mu^{I_2}_{ij} x^i x^j,
+P_1^{i} = x^i,\quad \quad P_2^{\beta} &= \mu^{\beta}_{ij} x^i x^j,
 \end{align}
-where $\mu^{I_2}$ are a basis for the rank-2 traceless symmetric $d\times d$ matrices.
+where $\mu^{\beta}$ are a basis for the rank-2 traceless symmetric $d\times d$ matrices.
 Let us call this group $\mathcal{M}^2_{\text{sym}}$.
@@
 We will continue to write things down in terms of the compact scalar $\phi$, which (at in the cases for which SSB is not preempted by strong fluctuations) is to be viewed as the Goldstone for the sponatneously broken symmetry.
-% References~\cite{Griffin2015, Gromov2019} discuss quantum field theories where configurations are weighted by an action.
-We will also want to consider thermal field theories with Hamiltonian descriptions, so we will abstractly discuss ``kinetic terms" in this section.
 
@@
 To find a kinetic term invariant under a given multipole group $\mathcal{M}$, we need to find operators $D$ built out of spatial derivatives that annihilate all the polynomials in $\mathcal{M}$.
-If $a_\text{max}$ is the highest degree of the polynomials in $\mathcal{M}$, the simplest derivative operators which generically do the job are of the form \ethan{may or may not want to switch to the composite index notation $q^{I_{a_{\rm max}+1}} \p_{I_{a_{\rm max}+1}}$ employed later}
+If $a_\text{max}$ is the highest degree of the polynomials in $\mathcal{M}$, the simplest derivative operators which generically do the job are of the form
 \begin{align}
-D = q^{i_1\dots i_{a_\text{max}+1}}\partial_{i_1}\dots \partial_{i_{a_\text{max}+1}},
+D = q^{I_{a_{\rm max}+1}} \p_{I_{a_{\rm max}+1}},
 \end{align}
-for some symmetric tensor $q^{i_1\dots i_{a_\text{max}+1}}$.
+for some symmetric tensor $q^{I_{a_{\rm max}+1}}$.
 Although it is not generically possible to do so~\cite{Gromov2019}, we can sometimes find a set of $D_\alpha$ ($\alpha$ is an abstract index) with $s\le a_{\text{max}}$, where $a_\text{max}$ is the highest degree of the polynomials.
 In this case, the effective field theory will be invariant under some non-maximal multipole symmetry.
-% We say that $D$ is order-$s$ if the maximum number of derivatives appearing in $D$ is $s$.
 
@@
 \begin{align}
-K_a[\phi(x)] &= (\partial_{i_1}\dots \partial_{i_{a+1}} \phi) (\partial_{i_1} \dots \partial_{i_{a+1}} \phi),
+K_a[\phi(x)] &= (\partial_{I_{a+1}} \phi) (\partial_{I_{a+1}} \phi),
 \end{align}
@@
 
-Consider first a clean classical system at $T>0$ with a spontaneously broken maximal multipole symmetry of degree $a$.
+Consider first a clean system at $T>0$ with a spontaneously broken maximal multipole symmetry of degree $a$.
 Then the kinetic term is proportional to 
 \begin{align}
-K[\phi(x)] = (\partial_{i_1} \dots \partial_{i_{a+1}} \phi)^2. \label{eqn:adisp}
+K[\phi(x)] = (\partial_{I_{a+1}} \phi)^2. \label{eqn:adisp}
 \end{align} 
@@
 Then the dispersion will be $k^4 + p^2$, where $k$ has $d_2$ components, $p$ has $d_1$ components, and $d = d_1+d_2$.
-Classically, the correlation function is
+At nonzero temperature, the correlation function is
 \begin{align}
@@
 
-The quantum correlation function behaves as
+The $T=0$ correlation function behaves as
 \begin{align}
@@
 \end{align}
-In a classical system, we then conclude that SSB cannot occur if $d_\eff\le 2$, while in a quantum system the critical dimension is $d_\eff = 1$.
+For $T>0$, we then conclude that SSB cannot occur if $d_\eff\le 2$, while at $T=0$ the critical dimension is $d_\eff = 1$.
 
@@
 \be \label{kappadef} \kappa_c^{I_c} = \frac{d\langle n_c^{I_c}\rangle}{d\mu_c^{I_c}},\ee 
-where $n_c$ is the number density of $c$-pole charges, $\mu_c$ is a conjugate chemical potential, and $I_c = \{i_1,\dots,i_c\}$ is a composite index.
+where $n_c$ is the number density of $c$-pole charges, $\mu_c$ is a conjugate chemical potential, and $I_c = \{i_1,\dots,i_c\}$ is a composite index as before.
 
@@
 where 
-%$I_b = \{i_1,\dots,i_b\}$ and $J_{a-b+1}=\{j_1,\dots,j_{a-b+1}\}$ are composite indices with 
 $\p_{J_{a-b+1}} \equiv \p_{j_1} \cdots \p_{j_{a-b+1}}$, and where $\phi_{I_b}$ is a field that shifts by a polynomial of degree $a-b$ under the action of $\mmax a$.
@@
 Thus the theory \eqref{generall} does not necessarily describe a phase where $\mmax b$ is spontaneously broken, and the exact pattern of symmetry breaking depends on the spatial dimension and whether or not $T$ is nonzero.
-For example, we can again consider the theory in \eqref{singlescalar} at $d=3,T>0$: even though the charge compressibility $\kappa_0$ (the coefficient of $(\p_\tau\phi)^2$ in \eqref{singlescalar}) is nonzero, $\phi$ fluctuates strongly enough that the monopole symmetry is not spontaneously broken.
+For example, we can again consider the theory in \eqref{singlescalar} at $d=3, T>0$: even though the charge compressibility $\kappa_0$ (the coefficient of $(\p_\tau\phi)^2$ in \eqref{singlescalar}) is nonzero, $\phi$ fluctuates strongly enough that the monopole symmetry is not spontaneously broken.
 
@@
 Thus the partial breaking of dipole symmetry provides us with an example where a continuous symmetry can be spontaneously broken in $d=1$ at zero temperature.
-
-%In a classical system at $T>0$ the correlation functions of this operator are
-%\begin{align}
-%\left\langle \partial_i\phi(x) \partial_j\phi(0) \right\rangle &\sim \delta_{ij} \int \frac{d^dk\, k^2}{k^4},
-%\end{align}
-%so that the critical dimension is $d=2$, as in the standard Mermin-Wagner theorem for monopole symmetries.
-This allows us to interpret the dipole charges of the monopole symmetry as monopole charges of the dipole part of the symmetry group.
-%However, this interpretation does not hold for  quantum systems. 
-%The correlation function scales as
-%\begin{align}
-%\left\langle \partial_i\phi(x) \partial_j\phi(0) \right\rangle &\sim \delta_{ij} \int \frac{d\omega\, d^dk\, k^2}{\omega^2 + k^4}\nn
-%&= \delta_{ij} \int \frac{d^dk\, k^2}{k^2},
-%\end{align}
-%which is always IR-finite. We thus conclude that at $T=0$, $\mcm^1_{\rm max}$ can be spontaneously broken down to $\mcm^0$ in any dimension
-
-%We then recognize that these fundamental dipoles, which are of course the dipole charges of the monopole symmetry, are also the monopole charges of the dipole symmetry.
-We should then expect symmetry breaking cannot occur for $d\le 2$ classically or $d\le 1$ quantum mechanically.
-%
-%
-%In more generality, for $G = \mathcal{M}^{a}_\text{max}$ and $H = \mathcal{M}^b_\text{max}$ there will be a charged operator $\partial_{i_1} \dots \partial_{i_{b+1}}\phi$ with a dispersion as in Eqn.~\ref{eqn:adisp}.
-Because $G$ is a maximal multipole group, we can still write the polynomials as in Eqn.~\ref{eqn:basis}.
-Recall that $P_c^{I_c}$ is a degree-$c$ monomial.
-As long as $\phi$ transforms as
-%\begin{align}
-%\delta \phi = \lambda_{I_c} P_c^{I_c}, \label{eqn:partial}
-%\end{align}
-%then $\partial_{i_1} \dots \partial_{i_{b+1}} \phi$ is invariant under $H=\mathcal{M}^b_\max$. 
-%
-%Classical correlation functions for these operators take the form
-%\begin{align}
-%C &= \left\langle \partial_{i_1} \dots \partial_{i_{b+1}} \phi(x) \partial_{j_1} \dots \partial_{j_{b+1}} \phi(0) \right\rangle \nn
-%&\sim \int \frac{d^dk\, k^{2(b+1)}}{k^{2(a+1)}},
-%\end{align}
-%and diverge when $d \le 2(a-b)\equiv d_\text{cl}$. On the other hand, the $T=0$ correlation functions behave as
-%\begin{align}
-%C &= \left\langle \partial_{i_1} \dots \partial_{i_{b+1}} \phi(x) \partial_{j_1} \dots \partial_{j_{b+1}} \phi(0) \right\rangle \nn
-%&\sim \int \frac{d\omega\, d^dk\, k^{2({b+1})}}{\omega^2 + k^{2(a+1)}}\nn
-%&\sim \int \frac{d^dk\, k^{2(b+1)}}{k^{a+1}},
-%\end{align}
-%so the quantum critical dimension is $d_\text{q} = a - 2b - 1$.
-We recover the maximal generalized Mermin-Wagner argument if we let $b=-1$ denote the trivial group ($b=0$ denotes the monopole group).
-Furthermore, the critical dimensions again satisfy $d_\cl = d_\text{q} + z$.
-%
-%
-%\subsection{General formulation of Mermin-Wagner}
-%% The kinetic term should take the form
-%% \begin{align}
-%% K[\phi(x)] &= (\partial_{i_{m+1}}\dots \partial_{i_n} \phi_{i_1 \dots i_m})^2,
-%% \end{align}
-%% which is invariant under shifts of the form of Eq.~\ref{eqn:partial}.
-The dispersion for the Goldstone modes is $K_k \sim k^{2(n-m)} = k^{2(a-b)}$.
-Following the correlation function calculation above, we find a classical critical dimension of $d = 2(a-b)$ and a quantum critical dimension of $d = a-b$.
-%
-%\subsection{\note{Different phases for the same subgroup}} \label{sub:phases}
-%
-%The previous subsection shows that when the kinetic term in the effective action is $K\sim (\partial_i\partial_j\phi)^2$, there are intermediate dimensions ($d=2$ for $T>0$ or $d=1,2$ for $T=0$) where the symmetry breaking pattern preserves the monopole group.
-For smaller dimensions there is no symmetry breaking, while for larger dimensions (with this effective action) the group is fully broken.
-%We can also write down theories where the kinetic term will be $(\partial_i\theta_j)^2$.
-The latter type of theory will preserve the monopole group in any large enough dimension.
-Since the latter theory is quadratic, it behaves like an ordinary theory, with a MW theorem prohibiting symmetry breaking for $d\le 2$ at $T>0$ or $d\le 1$ at zero temperature.
-%
-%We can organize these different theories that have the same SSB structure by their compressibilities.
-Since we are considering theories that conserve the dipole group, there will be dipole charges and monopole charges.
-Call the compressibility for the monopole charges $\mu_0$ and the compressibility for the dipole charges $\mu_1$.
-In any number of dimensions, there will be a Mott-insulator-like phase where $\mu_0=\mu_1  = 0$.
-The kinetic term $(\partial_i\theta_j)^2$ assumes $\mu_0=0$ and $\mu_1>0$, while the term $(\partial_i \partial_j \phi)^2$ assumes both compressibilies are nonzero.
-%
-%The list of compressibilies only tells us which fields are available for writing down an effective action.
-As in the previous subsection, which operators develop long-range order depends on the dimension.
-The calculations are analogous to previous correlation functions, and the results are as follows.
-At nonzero temperature, the order $b$ of the preserved subgroup $\mmax{b}$ is
-%\begin{align}
-%\begin{tabular}{l|r|r|r}
-%     & $d=1,2$ & $d=3,4$, & $d>4$ \\
-%    $(\partial_i\partial_j\phi)^2$ &  1 &  0 & -1 \\
-%    $(\partial_i\theta_j)^2$       &  1 &  0 &  0 \\
-%    Mott                           &  1 &  1 &  1
-%\end{tabular}
-%\end{align}
-%where we are, as always, denoting the trivial group by $b=-1$.
-Because the correlation calculations do not have any $\omega$ integral, the critical dimensions for a given symmetry-breaking pattern are the same in different theories.
-
-%The same is not true at $T=0$. Similar correlation function calculations show that the symmetry-breaking patterns are 
-%\begin{align}
-%\begin{tabular}{l|r|r|r}
-%     & $d=1$ & $d=2$, & $d>2$ \\
-%    $(\partial_i\partial_j\phi)^2$ &  0 &  0 & -1 \\
-%    $(\partial_i\theta_j)^2$       &  1 &  0 &  0 \\
-%    Mott                           &  1 &  1 &  1
-%\end{tabular} \label{eqn:zerophases}
-%\end{align}
-%The interesting difference from finite temperature is that the theory with a $\phi$ field can preserve $\mmax{0}$ at $d=1$, while the theory with a $\theta_j$ field cannot.
-%Both the $T>0$ and the $T=0$ table can easily be generalized to higher values of $a$, the order of the preserved multipole group.
-
-%Whenever there are distinct theories with the same symmetry-breaking pattern, some of the theories have operators with quasi-long-range order.
-\todo{QLRO vs SRO} For example, the $(\partial_i\theta_j)^2$  theory at $d=1$ acts like a Luttinger liquid of dipoles.
-A full analysis of quasi-long-range ordered fields is beyond the scope of this paper.
-Furthermore, we have only considered free fixed points.
-There certainly may be interacting theories with distinct phases and symmetry-breaking patterns~\cite{TonerRadzihovsky}.
-\ethan{how do you have interacting goldstones for an abelian symmetry?}
-
-% ------------------------
-
-% Once we have identified a symmetry group $G$ and its preserved subgroup $H$ we still have not fully specified a phase of matter.
-As an example, consider the previous case of partially breaking the maximal dipole group $\mathcal{M}^1_\max$ to the monopole group $\mathcal{M}^0_\max$.
-There is one theory with Lagrangian
-% \begin{align}
-% \mathcal{L}_2 = (\partial_t \phi)^2 + \sum_{ij} (\partial_i \partial_j \phi)^2,
-% \end{align}
-% which has critical dynamical exponent $z=2$. There is another theory with Lagrangian
-% \begin{align}
-% \mathcal{L}_1 &= (\partial_t \varphi_i)^2  + \sum_{ij} (\partial_i \varphi_j)^2,
-% \end{align}
-% and $z=1$. The two theories differ in that, if $\phi$ transforms as $\phi \goesto \phi + a_ix^i + b$, then $\varphi_i$ transforms as $\varphi_i \goesto \varphi_i + a_i$
-
-% \charlie{Does this mean the phase transition (if it exists) has to be topological,  \ethan{in what sense?} in the sense that the symmetry group and preserved subgroups are the same/ beyond Landau?}
-
-% For both theories, the classical critical dimension is $d_\cl = 2$.
-However, they have different quantum critical dimensions, $d_{\q, 1} = 1$ and $d_{\q,2}=0$.
-Both theories obey $d_\cl = d_\q + z$.
-Of course, for $d>2$ the $z=2$ theory develops long-range order for $e^{i\phi(x)}$, breaking the monopole symmetry in addition to the dipole symmetry.
-
-% We can now generalize to arbitrary $a$ and $b$. To do so, we define a field $\phi_{i_1\dots i_k}$ which transforms under $\mathcal{M}^a_\max$ as 
-% \begin{align}
-% \delta \phi_{i_1\dots i_k} = \lambda_{I_c} \partial_{i_1} \dots \partial_{i_k} P_c^{I_c}.
-% \end{align}
-% The kinetic term for this field will be
-% \begin{align}
-% K \sim (\partial_{i_{k+1}} \dots \partial_{i_{a+1}} \phi_{i_1\dots i_k})^2, \label{eqn:atok}
-% \end{align}
-% so that $z = 1+a-k$.
-Although $\phi_{i_1\dots i_k}$ is invariant under $\mathcal{M}^k_\max$, that field may not be able to develop long-range order.
-The fields that do develop LRO may be invariant under larger groups.
-In particular, $\partial_{i_{k+1}} \dots \partial_{i_{b+1}} \phi_{i_1\dots i_k} $ is invariant under $\mathcal{M}^b_\max$
-
-% At finite temperature, the field $\phi_{i_1\dots i_k}$ develops long-range order when $d>2(a-k)$, so that the preserved subgroup is $\mmax{k}$.
-When $2(a-b)<d\le 2(a-b+1)$, the preserved subgroup is $\mmax{b}$.
-The upshot is that there are $N^d_{ab}$ to spontaneously break the symmetry from $\mmax{a}$ to $\mmax{b}$ in $d$ spacial dimensions.
-The function is
-% \begin{align}
-% N^d_{ab} = \begin{cases} 0, \quad d\le 2(a-b) \\ 2 \\ 
-% \end{cases}
-% \end{align}
-
-% All of these theories will have $d_\cl = 2(a-b)$ and $d_\q = 2(a-b) - z$.
-
-% In addition to having different critical exponents $z$, these different theories with the same symmetry breaking patterns have different numbers of Goldstone modes.
-For the dipole examples considered above, the $z=1$ theory has $d$ massless modes (the components of $\varphi_i$) while the $z=2$ theory only has a single massless mode, $\phi$.
-\charlie{What else should we say about this?}
-
-% \todo{How should we address the question of stability?}
-% \ethan{I think my comments about stability are actually not relevant here --- they only imply that the transition between the ordered and disordered phases cannot be studied using by doing a perturbative analysis of the effects of vortex operators in the ordered phase}
 
@@
 Note that when the edge bosons are condensed, $H_\text{int}$ gives the site bosons an effective single particle hopping: thus condensation of single bosons can either proceed directly (by making $t_b$ large) or indirectly (by first condensing dipoles, and making the effective single-particle hopping term large).
-%and hence an effective $k^2$ dispersion, which would otherwise be disallowed by symmetry.
-\charlie{What would the effective dispersion actually be?
-Does it make sense to say that it is $k^2$ at short distances and $k^4$ at long distance?}
-
-%\charlie{Relevance of cosine terms and stability to crystal phase?}
-%\ethan{at large enough hopping strength we will generically just get a condensate and don't have to worry about this stuff}
 
@@
 
-%The phases of this model will be those in Eqn.~\ref{eqn:zerophases}.
-The values of $t_b$ and $t_d$ will determine which fields are available for the IR action.
-When the site bosons are condensed the $\phi$ field exists but might not be long-range ordered.
-Similarly when the edge bosons are condensed the $\theta_j$ field exists but might not be long-range ordered.
-For $d<2$ any symmetry-breaking pattern is possible.
-For $0<d\le 2$ the symmetry can not be completely broken, but the dipole part can be broken.
-
-%At sufficiently high dimension, this system will support three phases, which may be loosely described as the Mott insulator, the monopole superfluid where site bosons are condensed and the dipole superfluid where edge bosons (but not site bosons) are condensed.
-%Note that the dipole superfluid does preserve lattice rotations.
-%In the monopole superfluid phase, the effective Hamiltonian contains bare creation and annihilation terms for the edge bosons, which automatically breaks number conservation of the edge bosons, and effectively induces a dipole superfluid as well.
-We can identify these phases with symmetry breaking patterns of the multipole group.
-In the trivial phase the group is unbroken.
-In the dipole superfluid, the group is broken to its monopole subgroup.
-The monopole superfluid then corresponds to the fully broken symmetry.
-
-% In the absence of disorder, neither symmetry breaking pattern can occur for $d\le 1$. There may still be interesting phases, such as a Luttinger liquid.
-% For $d>2$ dipole and monopole symmetry breaking may occur, so all three phases are intact. Furthermore, transitions between any two phases are allowed.
-% This leaves the intermediate case of $d=2$, in which all three phases can exist, but there can be no transitions between any symmetry-unbroken phase and the monopole superfluid.
-Instead, there must be an intermediate phase with dipole but not monopole superfluidity.
-Representative single-parameter phase diagrams are shown in Fig.~\ref{fig:phases}.
-
-% \begin{figure}
-% \centering
-
-% \begin{tikzpicture}[scale = .9]
-% \draw[<->] (0,0) node[left]{(a)}  -- (8,0) node[right]{$s$};
-% \draw (4,-.1) -- (4,.1);
-
-% \node [align=center] (A) at (2, -0.5) {unbroken};
-% \node [align=center] (C) at (6, -0.5) {fully\\broken};
-
-% \end{tikzpicture}
-% \begin{tikzpicture}[scale = .9]
-
-% \draw[<->] (0,0) node[left]{(b)}  -- (8,0) node[right]{$s$};
-% \draw (2.667,-.1) -- (2.667,.1);
-% \draw (5.333,-.1) -- (5.333,.1);
-
-% \node [align=center] (A) at (1.333, -0.5) {unbroken};
-% \node [align=center] (B) at (4, -0.5) {dipole\\broken};
-% \node [align=center] (C) at (6.667, -0.5) {fully\\broken};
-% \end{tikzpicture}
-
-% \caption{\todo{Remove figure} \todo{New figure with different dipole-breaking phases?}(a) For  $d>2$, there can be direct transitions between symmetry-unbroken and the fully-broken phase.
-Here $s$ is intended to represent a parameter along an arbitrary line in the phase diagram that crosses this transition.
-(b) At $d=2$, although the symmetry can be fully broken, the fully broken phase cannot touch the unbroken phase.
-Any line through the phase diagram that connects the two phases must pass though the dipole-broken phase.
-These phase diagrams are for translation invariant systems at zero temperature.}
-% \label{fig:phases}
-% \end{figure}
-
 Let us now consider the same system, but with disorder.  The disorder Hamiltonian is
@@
 
-%At first, we will set $h_d=0$ but turn on $h_b$.
-This corresponds to the main considerations from Sec.~\ref{sec:disord}, where the disorder couples to the order parameter that fully breaks the symmetry.
-This raises the critical dimension for the long-range order of the site bosons.
-The edge bosons may still enter a superfluid phase for $d>1$, but the site bosons may not condense with long range phase order until $d>4$.
-There cannot be a direct transition from the insulator to the monopole superfluid until $d>8$.
-A phase akin to a Bose glass \cite{Fisheretal} of monopoles, where site bosons condense but without long-range order, can still exist in $1\le d<4$.
-
-%Now, let $h_d\ne 0$ while keeping $h_b=0$.
-The Imry-Ma argument for the edge bosons says that they should not be able to enter a superfluid phase when $d\le 4$.
-Once again, we can enter a Bose glass phase where $\langle d_{j, j+\mu} \rangle \ne 0$.
-In this case, the site bosons acquire a random single particle {\it hopping}.
-This may be expected to give rise to a Bose glass phase also for the site bosons.
-This implies that the monopoles cannot even undergo mediated condensation to superfluidity for $d\le 4$.
-For $d>4$, the dipoles may enter a superfluid phase, and the monopoles can enter a superfluid phase without mediation.
-
-%It is difficult to justify including including one disorder field but not the other.
-The standard reason to forbid disorder is by imposing translation invariance, but if we are already including one disorder field we are necessarily breaking translation symmetry.
 For $h_b,h_d\ne 0$, we can fully rely on the Imry-Ma argument.

### 2021-12-13 multipoleSSB/manuscript_Nov28.tex
changed name for version control

--- 2021-11-26 multipoleSSB/manuscript.tex
@@
 Multipole symmetries are of interest both as a window on fracton physics and as a crucial ingredient in realizing new universality classes for quantum dynamics.
-Here we address the question of whether and when multipole symmetries can be {\it spontaneously broken} in thermal equilibrium.
+Here we address the question of whether and when multipole symmetries can be {\it spontaneously broken} in thermal equilibrium \charlie{or at zero temperature}.
 We derive generalized Mermin-Wagner arguments for the total or partial breaking of multipolar symmetry groups and generalized Imry-Ma arguments for the robustness of such multipolar symmetry breaking to disorder.
@@
 The multipole group is well-explained in Ref.~\cite{Gromov2019}.
-It generalizes abelian global symmetries such as $\phi(x) \goesto\phi (x) +c$ by allowing a shift by some set of polynomials, $\phi (x) \goesto \phi (x) + \lambda_\alpha P^\alpha(x)$.
+For concreteness, and since we will be interested in situations where a multipolar symmetry group is spontaneously broken, we will describe the multipole group in terms of its action on a compact scalar field $\phi(x) \approx \phi(x) + 2\pi$.
+\ethan{are there any examples where $\phi$ is noncompact?} The multipole group generalizes the internal shift symmetry $\phi(x) \goesto\phi (x) +c$ by allowing a shift by some set of polynomials involving the spatial coordinates, viz.
+$\phi (x) \goesto \phi (x) + \lambda_\alpha P^\alpha(x)$.
 The variables $\lambda_\alpha$ are symmetry parameters, while $\alpha$ labels the set of polynomials $P^\alpha(x)$.
@@
 
-We can now consider building field theories that are invariant under various multipole groups.
-Reference~\cite{Gromov2019} describes the process in detail, so we will only sketch the outline.
-References~\cite{Griffin2015, Gromov2019} discuss quantum field theories where configurations are weighted by an action.
+We can now consider building IR field theories for phases in which a multipolar symmetry is completely spontaneously broken, with no residual unbroken subgroup.
+Reference~\cite{Gromov2019} describes the process in detail, so in the following we will be somewhat succinct.
+We will continue to write things down in terms of the compact scalar $\phi$, which (in the cases for which SSB is not preempted by strong fluctuations) is to be viewed as the Goldstone for the sponatneously broken symmetry.
+% References~\cite{Griffin2015, Gromov2019} discuss quantum field theories where configurations are weighted by an action.
 We will also want to consider thermal field theories with Hamiltonian descriptions, so we will abstractly discuss ``kinetic terms" in this section.
 
-The key step is to be able to find spatial derivatives $D$ that annihilate all the polynomials $P_a^{I_a}$ that we want to include in our multipole group.
-We say that $D$ is order-$s$ if it takes the form
+Since we are only interested in the low-energy physics of the putative symmetry-broken phase, constructing an appropriate IR field theory amounts to nothing more than constructing an appropriate kinetic term for $\phi$.
+To find a kinetic term invariant under a given multipole group $\mathcal{M}$, we need to find operators $D$ built out of spatial derivatives that annihilate all the polynomials in $\mathcal{M}$.
+If $a_\text{max}$ is the highest degree of the polynomials in $\mathcal{M}$, the simplest derivative operators which generically do the job are of the form
 \begin{align}
-D = q^i\partial_i + q^{ij}\partial_1\partial_j + \cdots + q^{i_1\dots i_s}\partial_{i_1}\dots\partial_{i_s},
+D = q^{i_1\dots i_{a_\text{max}+1}}\partial_{i_1}\dots \partial_{i_{a_\text{max}+1}},
 \end{align}
-where each $q$ is a symmetric tensor.
-Although it is not generically possible~\cite{Gromov2019}, we can sometimes find a set of $D_\alpha$ ($\alpha$ is an abstract index) with $s\le a_{\text{max}}$, where $a_\text{max}$ is the highest degree of the polynomials.
+for some symmetric tensor $q^{i_1\dots i_{a_\text{max}+1}}$.
+Although it is not generically possible to do so~\cite{Gromov2019}, we can sometimes find a set of $D_\alpha$ ($\alpha$ is an abstract index) with $s\le a_{\text{max}}$, where $a_\text{max}$ is the highest degree of the polynomials.
 In this case, the effective field theory will be invariant under some non-maximal multipole symmetry.
+% We say that $D$ is order-$s$ if the maximum number of derivatives appearing in $D$ is $s$.
 
-Generically, we cannot find any $D$ with $s\le a_\text{max}$ such that $DP_a^{I_a} = 0$ for all the polynomials we want. In that case we can still write down a theory with derivatives 
+Writing the invariant derivative operators as $D_\alpha$, the most general kinetic term is ~\cite{Gromov2019} 
 \begin{align}
-D = q^{i_1\dots i_n}\partial_{i_1}\dots \partial_{i_n},
+K[\phi(x)] &= g_{\alpha\beta} (D_\alpha \phi) (D_\beta \phi)
 \end{align}
-with $n=a_\text{max}+1$. Then, the field theory will be invariant under the maximal multipole group of degree $a_\text{max}$.
-
-Armed with the invariant derivatives $D_\alpha$ ($\alpha$ is another abstract index), we can start to construct a kinetic term.
-The most general term, at lowest order in derivatives, would be~\cite{Gromov2019}
-\begin{align}
-K[\phi(x)] &= g_{\alpha\beta} (D_\alpha \phi) (D_\beta \phi).
-\end{align}
+for some symmetric tensor $g_{\alpha\beta}$.
 We will often write the Fourier transform of the kinetic term as $\phi_{-k} K_k \phi_k$.
@@
 This already appears in Ref.~\cite{Griffin2015}, so we are simply reviewing it here in preparation for the more generic cases to follow.
-Here, we will discuss both classical and quantum systems.
+Here, we will discuss both thermal and $T=0$ systems.
 
-Consider a clean classical system with a spontaneously broken maximal multipole symmetry of degree $a=n-1$.
-Then the kinetic term is
+Consider first a clean classical system at $T>0$ with a spontaneously broken maximal multipole symmetry of degree $a$.
+Then the kinetic term is proportional to 
 \begin{align}
-K[\phi(x)] = (\partial_{i_1} \dots \partial_{i_n} \phi)^2, \label{eqn:adisp}
+K[\phi(x)] = (\partial_{i_1} \dots \partial_{i_{a+1}} \phi)^2. \label{eqn:adisp}
 \end{align} 
-where $\phi$ is the order parameter field. 
-Now assume we are in an ordered phase, but begin to nucleate domains of linear size $L$.
-Since the order parameter is continuous there will be a thick domain ``wall", in which the order parameter changes gradually from its value in the background phase to its value in the domain.
+ \ethan{really it's $\sim e^{i\phi}$ which is the order parameter}
+Now consider nucleating domains of linear size $L$ in which the expectation value of the order parameter --- namely $\langle e^{i\phi}\rangle$ --- varies by an amount of order 1.
+Since the order parameter is continuous there will be a thick domain ``wall" over which the order parameter changes gradually from its value in the background phase to its value in the domain.
 The thickness will be of order $L$.
@@
 As before, we interpret the divergence of the correlation function as a sign that the symmetry cannot be spontaneously broken.
-This allows us to reproduce our scaling argument that spontaneous symmetry breaking of the multipole group $\mathcal{M}^a_\max$ cannot occur for $d\le2(a+1)$.
+This allows us to reproduce our scaling argument that SSB of the multipole group $\mathcal{M}^a_\max$ cannot occur for $d\le2(a+1)$.
 
-If we look at the quantum (zero temperature) setting, the energy scaling no longer tells us where the critical dimension is.
+If we look at the zero temperature setting, the energy scaling no longer tells us where the critical dimension is.
 This is because even when there is no energy cost to forming domains, there is no entropy gain.
 Instead, quantum fluctuations must be the motivation for domain nucleation.
-To find the quantum critical dimension, we calculate the correlation function,
+To find the quantum critical dimension, we calculate the correlation function by way of the (imaginary-time) IR Lagrangian
+\begin{align}
+     \mathcal{L}[\phi] = (\partial_\tau\phi)^2 + K[\phi].
+\end{align} 
+This gives 
 \begin{align}
@@
 which now diverges at $d\le a+1$~\cite{Griffin2015}, where $d$ is the number of spatial dimensions.
-We have halved the critical dimension, so that spontaneous symmetry breaking cannot occur at $d\le a+1$.
-It is interesting to note that the dynamical critical exponent in these theories is $z = a+1$, so the classical and quantum critical dimensions are related by $d_\text{cl} = d_\text{q} + z$.
-
-\charlie{Is it more useful to say $d_\text{cl} = d_\text{q} + z$ or $d_\text{cl} = 2 d_\text{q}$, or not try to make any general statement at all?}
+We have halved the critical dimension, so that SSB cannot occur at $d\le a+1$.
+Note that the dynamical critical exponent in these theories is $z = a+1$, so that as expected, the classical and quantum critical dimensions are related by $d_\text{cl} = d_\text{q} + z$.
 
@@
 
-\subsection{Breaking to a subgroup} \label{sub:subgroup}
+In this section we examine what happens when a multipolar symmetry is spontaneously broken to a subgroup.
+Studying this problem will reveal a few ways in which our general expectations for symmetry breaking and universality need to be revised in the context of multipolar symmetries.
+
+\subsection{\note{Breaking to a subgroup}} \label{sub:subgroup}
 
@@
 Let us now consider the case where the order parameter breaks the symmetry from the original group $G$ to some subgroup $H$.
-For simplicity, we will let $G$ be the maximal multipole group $\mathcal{M}^{a}_\text{max}$ and $H$ be $\mathcal{M}^b_\text{max}$, with $b<a$.
+For simplicity, we will let $G$ be the maximal multipole group $\mathcal{M}^{a}_\text{max}$ and will let $H$ be $\mathcal{M}^b_\text{max}$, with $b<a$.
 
-The simplest case is when $a=1$ and $b=0$.
-In physical terms, we are taking a system that conserves total charge and total dipole moment, then condensing objects that carry a dipole moment but no total charge.
-\note{If $\phi(x)$ is charged under the monopole symmetry, then the operator $\partial_i \phi(x)$ is charged under the dipole part of the symmetry group.
-\charlie{Do we want correlation functions of $\partial_i\phi$ or $\exp(\partial_i\phi)$?} In a classical system at $T>0$ the correlation functions of this operator are
+As an illustrative example, consider the case when a dipolar symmetry is spontaneously broken to the monopolar subgroup, so that $a=1$ and $b=0$.
+In the spirit of Goldstone's theorem, the most natural thing to do would be to write down an action involving a set of fields $\theta_j$ which transform linearly under the dipolar part of the symmetry group, but which are invariant under the remaining monopole subgroup (i.e.
+under a transformation parameterized by $P(x) = \alpha + \beta_j x^j$, $\theta_j \mapsto \theta_j + \beta_j$).
+In this case, the minimal IR action for the putative symmetry-breaking phase would be
+\begin{align}
+    \mathcal{L}= \sum_j  (\partial_\tau \theta_j)^2 + \sum_{i,j} g_{ij} (\partial_i \theta_j)^2.
+\end{align}
+
+By calculating correlation functions of the $\theta_j$ fields using the above action, we would conclude that SSB would be possible only for $d>1$ at $T=0$, and $d>2$ at $T>0$.
+However, this argument does not actually give the correct lower critical dimension.
+To see this, consider instead the single scalar field theory
+\begin{align}
+    \mathcal{L} = (\partial_\tau\phi)^2 + g_{ij} (\partial_i\partial_j\phi)^2 \end{align}
+where $\phi$ transforms under the symmetry as $\phi \mapsto \phi + \alpha + \beta_jx^j$.  In a classical system at $T>0$ the correlation functions of this operator are
 \begin{align}
@@
 \end{align}
-so that the critical dimension is $d=2$, as in the standard Mermin-Wagner theorem. }
+so that the critical dimension is $d=2$, as in the standard Mermin-Wagner theorem for monopole symmetries.
+This allows us to interpret the dipole charges of the monopole symmetry as monopole charges of the dipole part of the symmetry group.
+
+However, this interpretation does not hold for  quantum systems. The correlation function scales as
+\begin{align}
+\left\langle \partial_i\phi(x) \partial_j\phi(0) \right\rangle &\sim \delta_{ij} \int \frac{d\omega\, d^dk\, k^2}{\omega^2 + k^4}\nn
+&= \delta_{ij} \int \frac{d^dk\, k^2}{k^2},
+\end{align}
+so that SSB is possible in any $d>0$. Thus, we cannot view this as analogous to ordinary monopole symmetry breaking.
 
@@
 
-\note{In more generality, for this pattern of symmetry breaking there will be a charged operator $\partial_{i_1} \dots \partial{i_m} \phi$, where $m=b+1$.
+In more generality, for $G = \mathcal{M}^{a}_\text{max}$ and $H = \mathcal{M}^b_\text{max}$ there will be a charged operator $\partial_{i_1} \dots \partial_{i_{b+1}}\phi$ with a dispersion as in Eqn.~\ref{eqn:adisp}.
 Because $G$ is a maximal multipole group, we can still write the polynomials as in Eqn.~\ref{eqn:basis}.
+Recall that $P_c^{I_c}$ is a degree-$c$ monomial.
 As long as $\phi$ transforms as
@@
 \end{align}
-then $\partial_{i_1} \dots \partial_{i_m} \phi$ is invariant under $H=\mathcal{M}^b_\max$. }
+then $\partial_{i_1} \dots \partial_{i_{b+1}} \phi$ is invariant under $H=\mathcal{M}^b_\max$.
 
-\note{Classical correlation functions for these operators take the form
+Classical correlation functions for these operators take the form
 \begin{align}
-C &= \left\langle \partial_{i_1} \dots \partial_{i_m} \phi(x) \partial_{j_1} \dots \partial_{j_m} \phi(0) \right\rangle \nn
-&\sim \int \frac{d^dk\, k^{2m}}{k^{2n}},
+C &= \left\langle \partial_{i_1} \dots \partial_{i_{b+1}} \phi(x) \partial_{j_1} \dots \partial_{j_{b+1}} \phi(0) \right\rangle \nn
+&\sim \int \frac{d^dk\, k^{2(b+1)}}{k^{2(a+1)}},
 \end{align}
-and diverge when $d \le 2(n-m) = 2(a-b)\equiv d_\text{cl}$. On the other hand, the $T=0$ correlation functions behave as
+and diverge when $d \le 2(a-b)\equiv d_\text{cl}$. On the other hand, the $T=0$ correlation functions behave as
 \begin{align}
-C &= \left\langle \partial_{i_1} \dots \partial_{i_m} \phi(x) \partial_{j_1} \dots \partial_{j_m} \phi(0) \right\rangle \nn
-&\sim \int \frac{d\omega\, d^dk\, k^{2m}}{\omega^2 + k^{2n}}\nn
-&\sim \int \frac{d^dk\, k^{2m}}{k^n},
+C &= \left\langle \partial_{i_1} \dots \partial_{i_{b+1}} \phi(x) \partial_{j_1} \dots \partial_{j_{b+1}} \phi(0) \right\rangle \nn
+&\sim \int \frac{d\omega\, d^dk\, k^{2({b+1})}}{\omega^2 + k^{2(a+1)}}\nn
+&\sim \int \frac{d^dk\, k^{2(b+1)}}{k^{a+1}},
 \end{align}
-so the quantum critical dimension is $d_\text{q} = n - 2m = a - 2b - 1$.
+so the quantum critical dimension is $d_\text{q} = a - 2b - 1$.
 We recover the maximal generalized Mermin-Wagner argument if we let $b=-1$ denote the trivial group ($b=0$ denotes the monopole group).
 Furthermore, the critical dimensions again satisfy $d_\cl = d_\text{q} + z$.
-\charlie{Distinguish $z$ and $z_\max$?}
-}
 
@@
 
-\subsection{No mediated condensation} \label{sub:mediated}
+\subsection{\note{Different phases for the same subgroup}} \label{sub:phases}
 
-We have seen that the generalized Mermin-Wagner argument provides critical dimensions for partial breaking that are lower than the critical dimensions for full symmetry breaking.
-\note{Can we use this behavior to try to evade the Mermin-Wagner argument for the full symmetry group?}
+The previous subsection shows that when the kinetic term in the effective action is $K\sim (\partial_i\partial_j\phi)^2$, there are intermediate dimensions ($d=2$ for $T>0$ or $d=1,2$ for $T=0$) where the symmetry breaking pattern preserves the monopole group.
+For smaller dimensions there is no symmetry breaking, while for larger dimensions (with this effective action) the group is fully broken.
+We can also write down theories where the kinetic term will be $(\partial_i\theta_j)^2$.
+The latter type of theory will preserve the monopole group in any large enough dimension.
+Since the latter theory is quadratic, it behaves like an ordinary theory, with a MW theorem prohibiting symmetry breaking for $d\le 2$ at $T>0$ or $d\le 1$ at zero temperature.
 
-%\remove{In fact, we can use this observation to evade the generalized Mermin-Wagner theorem, completely breaking any multipole symmetry for $d>2$ classically, or $d>1$ quantum mechanically}.
-This possibility has been pointed out in Refs.~\cite{Griffin2015Cascade, Griffin2015}, but we include a further interpretation in terms of what types of phase transitions are allowed in different dimensions.
-For simplicity, we will begin by considering the classical dipole-preserving theory with symmetry group $\mathcal{M}^2_\max$.
+We can organize these different theories that have the same SSB structure by their compressibilities.
+Since we are considering theories that conserve the dipole group, there will be dipole charges and monopole charges.
+Call the compressibility for the monopole charges $\mu_0$ and the compressibility for the dipole charges $\mu_1$.
+In any number of dimensions, there will be a Mott-insulator-like phase where $\mu_0=\mu_1  = 0$.
+The kinetic term $(\partial_i\theta_j)^2$ assumes $\mu_0=0$ and $\mu_1>0$, while the term $(\partial_i \partial_j \phi)^2$ assumes both compressibilies are nonzero.
 
-To be explicit, consider a monopole field $\phi(x)$ and a dipole field $\varphi_i(x)$. Under the dipole symmetry they transform as 
+The list of compressibilies only tells us which fields are available for writing down an effective action.
+As in the previous subsection, which operators develop long-range order depends on the dimension.
+The calculations are analogous to previous correlation functions, and the results are as follows.
+At nonzero temperature, the order $b$ of the preserved subgroup $\mmax{b}$ is
 \begin{align}
-\phi(x) &\goesto \phi(x) + a_ix^i + b,\nn
-\varphi_i(x) &\goesto \varphi_i(x) + a_i.
+\begin{tabular}{l|r|r|r}
+     & $d=1,2$ & $d=3,4$, & $d>4$ \\
+    $(\partial_i\partial_j\phi)^2$ &  1 &  0 & -1 \\
+    $(\partial_i\theta_j)^2$       &  1 &  0 &  0 \\
+    Mott                           &  1 &  1 &  1
+\end{tabular}
 \end{align}
-The leading terms in the Lagrangian will be
+where we are, as always, denoting the trivial group by $b=-1$.
+Because the correlation calculations do not have any $\omega$ integral, the critical dimensions for a given symmetry-breaking pattern are the same in different theories.
+
+The same is not true at $T=0$. Similar correlation function calculations show that the symmetry-breaking patterns are 
 \begin{align}
-\mathcal{L} &= (\partial_t \phi)^2 + (\partial_t \varphi_i)^2 + (\partial_i \partial_j \phi)^2 + (\partial_i \varphi_j)^2,
+\begin{tabular}{l|r|r|r}
+     & $d=1$ & $d=2$, & $d>2$ \\
+    $(\partial_i\partial_j\phi)^2$ &  0 &  0 & -1 \\
+    $(\partial_i\theta_j)^2$       &  1 &  0 &  0 \\
+    Mott                           &  1 &  1 &  1
+\end{tabular} \label{eqn:zerophases}
 \end{align}
-so that $\phi$ has a $k^4$ dispersion and $\varphi_i$ has a $k^2$ dispersion.
-We can also include an interaction term of the form $g(\partial_i \phi - \varphi_i)^2$, which does not break the symmetry.
-The $\varphi_i$ field is well-defined whenever the dipole symmetry is broken.
+The interesting difference from finite temperature is that the theory with a $\phi$ field can preserve $\mmax{0}$ at $d=1$, while the theory with a $\theta_j$ field cannot.
+Both the $T>0$ and the $T=0$ table can easily be generalized to higher values of $a$, the order of the preserved multipole group.
 
-\note{We can then redefine the $\varphi_i$ field to $\tilde{\varphi}_i = \varphi_i - \partial_i \phi$, so that it is neutral under the symmetry.
-The interaction term becomes $g\tilde{\varphi}_i^2$, so that the $\tilde{\varphi}_i$ field becomes massive.
-The only remaining massless field is $\phi$, which remains charged under the full dipole group and keeps its $k^4$ dispersion.
-As a result, the critical dimension for breaking the full dipole group is not affected by the possibility of partially breaking.
-}
-\charlie{What else to say here? Compare to Griffin/Grosvenor?}
+Whenever there are distinct theories with the same symmetry-breaking pattern, some of the theories have operators with quasi-long-range order.
+\todo{QLRO vs SRO} For example, the $(\partial_i\theta_j)^2$  theory at $d=1$ acts like a Luttinger liquid of dipoles.
+A full analysis of quasi-long-range ordered fields is beyond the scope of this paper.
+Furthermore, we have only considered free fixed points.
+There certainly may be interacting theories with distinct phases and symmetry-breaking patterns~\cite{TonerRadzihovsky}.
 
-%As long as $\phi_i$ is well-defined, integrating out fluctuations in $\phi_i$ endows the monopole field $\phi$ with a $k^2$ dispersion.
-The coupling $g$ is set by how far the theory sits from the dipole critical point.
+% ------------------------
 
-%With a $k^2$ dispersion for the monopole field, we now have a route to evade the (generalized) Mermin-Wagner argument and fully break the dipole symmetry in dimensions $d> 2$.
-First, partially break the symmetry so that $\phi_i$ is well defined, which is allowed for any $d>2$.
-Then, we can use the effective dispersion of the $\phi$ field to break the remaining part of the symmetry, also in any $d>2$.
-This procedure can be generalized, so that any multipole symmetry can be fully broken in $d>2$, provided we have enough symmetry-breaking order parameters.
-The discussion also translates to the quantum setting, so that the ultimate critical (space) dimension is always $d=1$.
+% Once we have identified a symmetry group $G$ and its preserved subgroup $H$ we still have not fully specified a phase of matter.
+As an example, consider the previous case of partially breaking the maximal dipole group $\mathcal{M}^1_\max$ to the monopole group $\mathcal{M}^0_\max$.
+There is one theory with Lagrangian
+% \begin{align}
+% \mathcal{L}_2 = (\partial_t \phi)^2 + \sum_{ij} (\partial_i \partial_j \phi)^2,
+% \end{align}
+% which has critical dynamical exponent $z=2$. There is another theory with Lagrangian
+% \begin{align}
+% \mathcal{L}_1 &= (\partial_t \varphi_i)^2  + \sum_{ij} (\partial_i \varphi_j)^2,
+% \end{align}
+% and $z=1$. The two theories differ in that, if $\phi$ transforms as $\phi \goesto \phi + a_ix^i + b$, then $\varphi_i$ transforms as $\varphi_i \goesto \varphi_i + a_i$
 
-%The preceding procedure gives us insight into what the generalized Mermin-Wagner arguments were telling us.
-When we wrote down those arguments, we did not include the intermediate fields.
-Thus, the generalized Mermin-Wagner arguments tell us when a direct transition between two phases is or is not allowed.
-Consider again the dipole group $\mathcal{M}^2_\max$.
-The phase in which the symmetry is fully broken can exist in any dimension $d>2$ ($d>1$ for quantum systems), but there can only be direct phase transitions between this phase and the wholly symmetry unbroken phase in $d>4$ ($d>2$).
-In dimensions $2<d<4$ ($1<d<2$) the fully symmetry broken and the fully symmetry unbroken phase must be separated by an intermediate phase that breaks dipole but not monopole symmetry.
+% \charlie{Does this mean the phase transition (if it exists) has to be topological,  \ethan{in what sense?} in the sense that the symmetry group and preserved subgroups are the same/ beyond Landau?}
+
+% For both theories, the classical critical dimension is $d_\cl = 2$.
+However, they have different quantum critical dimensions, $d_{\q, 1} = 1$ and $d_{\q,2}=0$.
+Both theories obey $d_\cl = d_\q + z$.
+Of course, for $d>2$ the $z=2$ theory develops long-range order for $e^{i\phi(x)}$, breaking the monopole symmetry in addition to the dipole symmetry.
+
+% We can now generalize to arbitrary $a$ and $b$. To do so, we define a field $\phi_{i_1\dots i_k}$ which transforms under $\mathcal{M}^a_\max$ as 
+% \begin{align}
+% \delta \phi_{i_1\dots i_k} = \lambda_{I_c} \partial_{i_1} \dots \partial_{i_k} P_c^{I_c}.
+% \end{align}
+% The kinetic term for this field will be
+% \begin{align}
+% K \sim (\partial_{i_{k+1}} \dots \partial_{i_{a+1}} \phi_{i_1\dots i_k})^2, \label{eqn:atok}
+% \end{align}
+% so that $z = 1+a-k$.
+Although $\phi_{i_1\dots i_k}$ is invariant under $\mathcal{M}^k_\max$, that field may not be able to develop long-range order.
+The fields that do develop LRO may be invariant under larger groups.
+In particular, $\partial_{i_{k+1}} \dots \partial_{i_{b+1}} \phi_{i_1\dots i_k} $ is invariant under $\mathcal{M}^b_\max$
+
+% At finite temperature, the field $\phi_{i_1\dots i_k}$ develops long-range order when $d>2(a-k)$, so that the preserved subgroup is $\mmax{k}$.
+When $2(a-b)<d\le 2(a-b+1)$, the preserved subgroup is $\mmax{b}$.
+The upshot is that there are $N^d_{ab}$ to spontaneously break the symmetry from $\mmax{a}$ to $\mmax{b}$ in $d$ spacial dimensions.
+The function is
+% \begin{align}
+% N^d_{ab} = \begin{cases} 0, \quad d\le 2(a-b) \\ 2 \\ 
+% \end{cases}
+% \end{align}
+
+% All of these theories will have $d_\cl = 2(a-b)$ and $d_\q = 2(a-b) - z$.
+
+% In addition to having different critical exponents $z$, these different theories with the same symmetry breaking patterns have different numbers of Goldstone modes.
+For the dipole examples considered above, the $z=1$ theory has $d$ massless modes (the components of $\varphi_i$) while the $z=2$ theory only has a single massless mode, $\phi$.
+\charlie{What else should we say about this?}
+
+% \todo{How should we address the question of stability?}
+% \ethan{I think my comments about stability are actually not relevant here --- they only imply that the transition between the ordered and disordered phases cannot be studied using by doing a perturbative analysis of the effects of vortex operators in the ordered phase}
 
@@
 The term $H_\text{int}$ is allowed by the symmetry because it simultaneously removes an edge boson and creates a dipole of site bosons.
-When the edge bosons are condensed, $H_\text{int}$ gives the site bosons an effective single particle hopping, and hence an effective $k^2$ dispersion, which would otherwise be disallowed by symmetry.
-\todo{Yes to effective hopping, no to $k^2$ dispersion.
-Discuss SF/SC?}
+When the edge bosons are condensed, $H_\text{int}$ gives the site bosons an effective single particle hopping.
+%and hence an effective $k^2$ dispersion, which would otherwise be disallowed by symmetry.
+\charlie{What would the effective dispersion actually be?
+Does it make sense to say that it is $k^2$ at short distances and $k^4$ at long distance?}
 
-\todo{Redo discussion of phases. Reframe in terms of symmetry-breaking patterns, not phases.}
+\charlie{Relevance of cosine terms and stability to crystal phase?}
 
-\todo{Define multiple dipole-breaking patterns at the effective level. Relevance of cosine terms and stability to crystal phase?}
-At sufficiently high dimension, this system will support three phases, which may be loosely described as the Mott insulator, the monopole superfluid where site bosons are condensed and the dipole superfluid where edge bosons (but not site bosons) are condensed.
-Note that the dipole superfluid does preserve lattice rotations.
-In the monopole superfluid phase, the effective Hamiltonian contains bare creation and annihilation terms for the edge bosons, which automatically breaks number conservation of the edge bosons, and effectively induces a dipole superfluid as well.
+The phases of this model will be those in Eqn.~\ref{eqn:zerophases}.
+The values of $t_b$ and $t_d$ will determine which fields are available for the IR action.
+When the site bosons are condensed the $\phi$ field exists but might not be long-range ordered.
+Similarly when the edge bosons are condensed the $\theta_j$ field exists but might not be long-range ordered.
+For $d<2$ any symmetry-breaking pattern is possible.
+For $0<d\le 2$ the symmetry can not be completely broken, but the dipole part can be broken.
+\charlie{What is the effect of $H_\text{int}$?}
+
+%At sufficiently high dimension, this system will support three phases, which may be loosely described as the Mott insulator, the monopole superfluid where site bosons are condensed and the dipole superfluid where edge bosons (but not site bosons) are condensed.
+%Note that the dipole superfluid does preserve lattice rotations.
+%In the monopole superfluid phase, the effective Hamiltonian contains bare creation and annihilation terms for the edge bosons, which automatically breaks number conservation of the edge bosons, and effectively induces a dipole superfluid as well.
 We can identify these phases with symmetry breaking patterns of the multipole group.
@@
 
-At first, we will set $h_d=0$ but turn on $h_b$.
+%At first, we will set $h_d=0$ but turn on $h_b$.
 This corresponds to the main considerations from Sec.~\ref{sec:disord}, where the disorder couples to the order parameter that fully breaks the symmetry.
-This raises the critical dimension for the superfluidity of the site bosons.
+This raises the critical dimension for the long-range order of the site bosons.
 The edge bosons may still enter a superfluid phase for $d>1$, but the site bosons may not condense with long range phase order until $d>4$.
@@
 
-It is difficult to justify including including one disorder field but not the other.
+%It is difficult to justify including including one disorder field but not the other.
 The standard reason to forbid disorder is by imposing translation invariance, but if we are already including one disorder field we are necessarily breaking translation symmetry.
 For $h_b,h_d\ne 0$, we can fully rely on the Imry-Ma argument.
-No symmetry breaking can occur for $d\le 4$, the monopole superfluid must be separated from any symmetry-unbroken phase by a dipole superfluid for $4<d\le 8$, and any symmetry-breaking transition can occur for $d>8$.
+No symmetry breaking can occur for $d\le 4$, the dipole part of the symmetry may be broken for $4<d\le 8$, and any symmetry-breaking phase can occur for $d>8$.
 
 \section{Discussion} \label{sec:disc}
+
+\charlie{Should we say anything about Ref.~\cite{Argurio2021}, which doesn't consider multipole symmetry, but does look at restricted mobility and crystal-like phases from abnormal kinetic terms?}
 
@@
 
+\appendix
+
+\section{No mediated condensation} \label{app:mediated}
+
+In Sec.~\ref{sec:partial} we saw that the generalized Mermin-Wagner argument provides critical dimensions for partial breaking that are lower than the critical dimensions for full symmetry breaking.
+\note{Can we use this behavior to try to evade the Mermin-Wagner argument for the full symmetry group?}
+
+%\remove{In fact, we can use this observation to evade the generalized Mermin-Wagner theorem, completely breaking any multipole symmetry for $d>2$ classically, or $d>1$ quantum mechanically}.
+This possibility has been pointed out in Refs.~\cite{Griffin2015Cascade, Griffin2015}, but we include a further interpretation in terms of what types of phase transitions are allowed in different dimensions.
+For simplicity, we will begin by considering the classical dipole-preserving theory with symmetry group $\mathcal{M}^2_\max$.
+
+To be explicit, consider a monopole field $\phi(x)$ and a dipole field $\varphi_i(x)$. Under the dipole symmetry they transform as 
+\begin{align}
+\phi(x) &\goesto \phi(x) + a_ix^i + b,\nn
+\varphi_i(x) &\goesto \varphi_i(x) + a_i.
+\end{align}
+The leading terms in the Lagrangian will be
+\begin{align}
+\mathcal{L} &= (\partial_t \phi)^2 + (\partial_t \varphi_i)^2 + (\partial_i \partial_j \phi)^2 + (\partial_i \varphi_j)^2,
+\end{align}
+so that $\phi$ has a $k^4$ dispersion and $\varphi_i$ has a $k^2$ dispersion.
+We can also include an interaction term of the form $g(\partial_i \phi - \varphi_i)^2$, which does not break the symmetry.
+The $\varphi_i$ field is well-defined whenever the dipole symmetry is broken.
+
+\note{We can then redefine the $\varphi_i$ field to $\tilde{\varphi}_i = \varphi_i - \partial_i \phi$, so that it is neutral under the symmetry.
+The interaction term becomes $g\tilde{\varphi}_i^2$, so that the $\tilde{\varphi}_i$ field becomes massive.
+The only remaining massless field is $\phi$, which remains charged under the full dipole group and keeps its $k^4$ dispersion.
+As a result, the critical dimension for breaking the full dipole group is not affected by the possibility of partially breaking.
+}
+\charlie{What else to say here? Compare to Griffin/Grosvenor?}
+\todo{Compare to fracton superfluids}
+
+%As long as $\phi_i$ is well-defined, integrating out fluctuations in $\phi_i$ endows the monopole field $\phi$ with a $k^2$ dispersion.
+The coupling $g$ is set by how far the theory sits from the dipole critical point.
+
+%With a $k^2$ dispersion for the monopole field, we now have a route to evade the (generalized) Mermin-Wagner argument and fully break the dipole symmetry in dimensions $d> 2$.
+First, partially break the symmetry so that $\phi_i$ is well defined, which is allowed for any $d>2$.
+Then, we can use the effective dispersion of the $\phi$ field to break the remaining part of the symmetry, also in any $d>2$.
+This procedure can be generalized, so that any multipole symmetry can be fully broken in $d>2$, provided we have enough symmetry-breaking order parameters.
+The discussion also translates to the quantum setting, so that the ultimate critical (space) dimension is always $d=1$.
+
+%The preceding procedure gives us insight into what the generalized Mermin-Wagner arguments were telling us.
+When we wrote down those arguments, we did not include the intermediate fields.
+Thus, the generalized Mermin-Wagner arguments tell us when a direct transition between two phases is or is not allowed.
+Consider again the dipole group $\mathcal{M}^2_\max$.
+The phase in which the symmetry is fully broken can exist in any dimension $d>2$ ($d>1$ for quantum systems), but there can only be direct phase transitions between this phase and the wholly symmetry unbroken phase in $d>4$ ($d>2$).
+In dimensions $2<d<4$ ($1<d<2$) the fully symmetry broken and the fully symmetry unbroken phase must be separated by an intermediate phase that breaks dipole but not monopole symmetry.
+
 \end{document}

### 2021-12-13 multipoleSSB/manuscript_Dec10.tex
comments from Rahul

--- 2021-12-10 multipoleSSB/charlies_edits.tex
@@
 We will always imagine $\phi$ as constituting the phase mode for some microscopic order parameter $\psi \sim e^{i\phi}$, with $\psi$ transforming by various $U(1)$ phases under the multipole group.
-\ethan{are there any examples showing that this interpretation is too restrictive?}
+%\ethan{are there any examples showing that this interpretation is too restrictive?} RN: Not that I know of
 
- The multipole group generalizes the internal shift symmetry $\phi(x) \goesto\phi (x) +c$ by allowing a shift by some set of polynomials involving the spatial coordinates, viz.
-$\phi (x) \goesto \phi (x) + \lambda_\alpha P^\alpha(x)$.
+ The multipole group generalizes the internal shift symmetry $\phi(x) \mapsto \phi (x) +c$ by allowing a shift by some set of polynomials involving the spatial coordinates, viz.
+$\phi (x) \mapsto \phi (x) + \lambda_\alpha P^\alpha(x)$.
 The variables $\lambda_\alpha$ are symmetry parameters, while $\alpha$ labels the set of polynomials $P^\alpha(x)$.
@@
 It is helpful to limit ourselves to homogeneous polynomials.
-We can label these as $P_a^{I_a}$, where $a$ is the degree of the polynomial and $I_a$ is an abstract index that runs over the polynomials of degree $a$.
+We can label these as $P_a^{\alpha}$, where $a$ is the degree of the polynomial and $\alpha$ is now an index that runs over the homogeneous of degree $a$.
 
@@
 The simplest case is the maximal multipole group $\mathcal{M}^a_\text{max}$, which includes all shifts by polynomials of degree $a$ or less.
-Individual monomials can be written as
+Individual polynomials can be written as
 \begin{align}
@@
 \end{align}
-where $\gamma$ is an index that runs over monomials of degree $c$ and $c\le a$.
+where $\gamma$ is an index that runs over polynomials of degree $c$ and $c\le a$.
 The composite index $I_c = \{i_1,\dots,i_c\}$  expands as $x^{I_c}=x^{i_1}\cdots x^{i_c}$, and similarly for $\partial_{I_c}$.
@@
 Reference~\cite{Gromov2019} describes the process in detail, so in the following we will be somewhat succinct.
-We will continue to write things down in terms of the compact scalar $\phi$, which (at in the cases for which SSB is not preempted by strong fluctuations) is to be viewed as the Goldstone for the sponatneously broken symmetry.
+We will continue to write things down in terms of the compact scalar $\phi$, which (at least in the cases for which SSB is not preempted by strong fluctuations) is to be viewed as the Goldstone for the sponatneously broken symmetry.
 
@@
 \end{align}
-suggests that, at least in some broad class of theories, the quantum critical dimension will be half the classical critical dimension.
-This is not true in general, though, as we will show for theories with quenched disorder.
+suggests that, at least in some broad class of theories, the quantum critical dimension will be half the classical critical dimension. This is not true in general, though, as we will show.
 
@@
 
-Consider a situation in which a dipolar symmetry $\mcm^1_{\rm max}$ is spontaneously broken to its monopolar subgroup $\mcm^0$, and momentarily specialize to the case of $d=3$ and $T>0$.
+Consider a situation in which a dipolar symmetry $\mcm^1_{\rm max}$ is spontaneously broken to its monopolar subgroup $\mcm^0$, and specialize to the case of $d=3$ and $T>0$.
 In the spirit of Goldstone's theorem, the most naive thing to do when analyzing the symmetry broken phase would be to write down an action involving a set of $d=3$ Goldstone fields $\theta_j$, with $\t_j$ transforming linearly under the dipolar part of the symmetry group, but remaining invariant under the unbroken monopole subgroup, so that $\theta_j \mapsto \theta_j + \beta_j$ under a transformation parameterized by $P(x) = \alpha + \beta_j x^j$.
@@
 Let $b\leq a$ denote the smallest degree of multipole charge with nonzero compressibility, i.e.
-consider a situation in which $\kappa_d>0$ for all $b\leq d\leq a$, while $\kappa_d=0$ for all $d<b$.
+consider a situation in which $\kappa_c>0$ for all $b\leq c\leq a$, while $\kappa_c=0$ for all $c<b$.
 Then the minimal IR Lagrangian for the symmetry broken phase looks schematically like
@@
 Indeed, if $\mmax b$ is spontaneously broken then $n_b$ creates Goldstones when acting on the ground state, thereby ensuring that the density-density correlator appearing in the calculation of \eqref{kappadef} must be nonzero at long wavelengths.
-\ethan{is this too laconic?}
+% \ethan{is this too laconic?} RN Seems fine to me
 
@@
 In the previous subsection, we saw that the number of Goldstone modes in the symmetry-breaking theory is determined by the smallest degree of multipole charge $b$ such that $\kappa_b\neq0$.
-We now turn to determining the symmetry breaking pattern that occurs for a given choice of $b,d,$ and $T$, thereby allowing us to formulate a generalized version of the Mermin-Wagner theorem for the case of partial multipole breaking.
+We now turn to determining the symmetry breaking pattern that occurs for a given choice of $b$, spatial dimension $d,$ and temperature $T$, thereby allowing us to formulate a generalized version of the Mermin-Wagner theorem for the case of partial multipole breaking.
 
-Since $\kappa_d=0$ if $d<b$, none of the subgroups $\mmax {d<b}$ can be spontaneously broken.
+Since $\kappa_c=0$ if $c<b$, none of the subgroups $\mmax {c<b}$ can be spontaneously broken.
 Suppose then that the symmetry is spontaneously broken down to $\mmax {c}$, where $c\geq b$.
@@
 In a disordered classical system, the energy scaling argument gives a nice explanation for why the critical dimension changes.
-Consider the formation of a domain in an otherwise ordered phase fully breaking the maximal multipole group $\mathcal{M}^a_\max$.
+Consider the formation of a domain of linear size $L$ in an otherwise ordered phase fully breaking the maximal multipole group $\mathcal{M}^a_\max$.
 There are $\sim L^d$ disorder samples in the new domain.
@@
 Comparing this to the cost of domain formation we calculated previously, the ordered phase will be unstable to domain nucleation when $d\le 4(a+1)$.
+We emphasize that we are here considering disorder that couples linearly to the order parameter.
 
@@
 The Imry-Ma argument only appeals to energy considerations, so the quantum critical dimension in the presence of disorder remains the same as the classical critical dimension~\cite{Vojta2013}.
-In disordered quantum systems, SSB is impossible for $d\le 4(a+1)$.
+In disordered quantum systems, full SSB is impossible for $d\le 4(a+1)$.
 
@@
 We are not aware of a way to condense objects with dipolar charge while preserving {\it continuous} rotation symmetry.
-\ethan{when doing mean field for the dipoles the potential may be such as to favor a case where only one species condenses.
-The quartic part of the potential is a complicated function that depends on the dimension and parent Mott insulator, so it's better to just discuss both possibilities}
+%\ethan{when doing mean field for the dipoles the potential may be such as to favor a case where only one species condenses.
+The quartic part of the potential is a complicated function that depends on the dimension and parent Mott insulator, so it's better to just discuss both possibilities} RN Makes sense, and ok
 
@@
 
-\charlie{Should we say anything about Ref.~\cite{Argurio2021}, which doesn't consider multipole symmetry, but does look at restricted mobility and crystal-like phases from abnormal kinetic terms?}
-
-In this paper we analyzed the spontaneous symmetry breaking of various multipole groups and discussed generalized Mermin-Wagner theorems, which we argued should be best understood as constraints on when a direct transition fully breaking a multipole group can occur, and when a symmetry unbroken phase and a (monopole) symmetry breaking phase must be separated by intermediate phases breaking higher multipole symmetries \ethan{is this still the best way of thinking about things?
-(also by ``direct transition'' I am assuming you mean ``direct continuous transition''?)}.
+In this paper we analyzed the spontaneous symmetry breaking of various multipole groups and discussed generalized Mermin-Wagner theorems for breaking a maximal multipole group, either fully or to a subgroup.
 We also considered multipole groups that are not the maximal multipole group, and the effect of quenched disorder.
@@
 We should be clear that the arguments in this paper are Mermin-Wagner or Imry-Ma arguments.
-These essentially amount to stability analyses about the Gaussian symmetry breaking fixed point.
-\ethan{I don't really understand how these are ``stability'' arguments --- stability with respect to which perturbations?
+%These essentially amount to stability analyses about the Gaussian symmetry breaking fixed point.
+%\ethan{I don't really understand how these are ``stability'' arguments --- stability with respect to which perturbations?
 The free theories we wrote down are at least all stable in the RG sense (the core energies of any vortex configurations can be made arbitrarily large, which means that the Gaussian actions can describe the IR physics even when the fluctuations are strong enough to prevent the symmetries from being spontaneously broken).
-Maybe there's something with disorder that I'm missing?}  In principle, even when the Gaussian symmetry breaking fixed point is unstable, a non-trivial fixed point with long range order could arise (see e.g.
+Maybe there's something with disorder that I'm missing?} RN: Ethan, if stuff is stable, there is no problem.
+What this discussion was intended to say, is that even if the Gaussian symmetry breaking fixed point is unstable to fluctuations, you could have a non-trivial but still symmetry breaking fixed point, and that is beyond the analysis.
+is the new discussion clearer?
+In principle, even when the Gaussian symmetry breaking fixed point is unstable to fluctuations, a non-trivial fixed point with long range order could arise (see e.g.
 \cite{TonerRadzihovsky}).
@@
 For $n=2$, there can in addition be quasi-long-range-ordered phases.
-Determining what kind of phase {\it can} obtain in the absence of long range ordered symmetry breaking is beyond the scope of this work, and would at a minimum require understanding the nature and role of topological defects in the symmetry breaking order parameter, akin to vortices in the XY model.
-\ethan{I think that since all the symmetries we are considering are Abelian, we can always have QLRO in the dimensions where the Gaussian action allows it (we can always work in a limit where the vortex core energy is infinitely large, meaning that we can always find a regime in which the description in terms of a phase mode is legitimate).
+Determining what kind of phase {\it can} be obtained in the absence of long range ordered symmetry breaking would at a minimum require understanding the nature and role of topological defects in the symmetry breaking order parameter, akin to vortices in the XY model.
+We expect that if we stick with an $O(2)$ internal group, then quasi-long range order should always be a possibility at the marginal spatial dimension, at least in the limit where the vortex core energy is infinitely large.
+The extension to non-Abelian internal groups is beyond the scope of this work.
+%\ethan{I think that since all the symmetries we are considering are Abelian, we can always have QLRO in the dimensions where the Gaussian action allows it (we can always work in a limit where the vortex core energy is infinitely large, meaning that we can always find a regime in which the description in terms of a phase mode is legitimate).
 The case of $O(n)$ models with $n>2$ is different because the sigma model for the putative broken phase is non-Abelian, with the stiffness flowing to zero.
 In the present Abelian case there is no such flow, and vortices are the only thing standing in the way of a Gaussian description.
-But as mentioned above, they can be gotten rid of (in any dimension) by making their core energies large.} The range of possible symmetry unbroken phases could be even richer in the presence of disorder, where various glassy phases could also come into play \cite{Fisheretal}.
+But as mentioned above, they can be gotten rid of (in any dimension) by making their core energies large.} RN: Accepted Ethan's comment and incorporated it into the text
+The range of possible symmetry unbroken phases could also be even richer in the presence of disorder, where various glassy phases could also come into play \cite{Fisheretal}.
 Our discussion of disorder physics was also limited to quenched short-range correlated disorder.
@@
 the U.S. Department of Energy, Office of Science, Basic
-Energy Sciences, under Award \# DE-SC0021346. EL was supported by the Hertz Fellowship.
+Energy Sciences, under Award \# DE-SC0021346. E.L. was supported by a Hertz Fellowship.
 
@@
 
-\appendix
-
-\section{No mediated condensation} \label{app:mediated}
-
-\ethan{not sure if this appendix is necessary, given that the paper is about MW theorems?}
-
-In Sec.~\ref{sec:partial} we saw that the generalized Mermin-Wagner argument provides critical dimensions for partial breaking that are lower than the critical dimensions for full symmetry breaking.
-\note{Can we use this behavior to try to evade the Mermin-Wagner argument for the full symmetry group?}
-
-%\remove{In fact, we can use this observation to evade the generalized Mermin-Wagner theorem, completely breaking any multipole symmetry for $d>2$ classically, or $d>1$ quantum mechanically}.
-This possibility has been pointed out in Refs.~\cite{Griffin2015Cascade, Griffin2015}, but we include a further interpretation in terms of what types of phase transitions are allowed in different dimensions.
-For simplicity, we will begin by considering the classical dipole-preserving theory with symmetry group $\mathcal{M}^2_\max$.
-
-To be explicit, consider a monopole field $\phi(x)$ and a dipole field $\varphi_i(x)$. Under the dipole symmetry they transform as 
-\begin{align}
-\phi(x) &\goesto \phi(x) + a_ix^i + b,\nn
-\varphi_i(x) &\goesto \varphi_i(x) + a_i.
-\end{align}
-The leading terms in the Lagrangian will be
-\begin{align}
-\mathcal{L} &= (\partial_t \phi)^2 + (\partial_t \varphi_i)^2 + (\partial_i \partial_j \phi)^2 + (\partial_i \varphi_j)^2,
-\end{align}
-so that $\phi$ has a $k^4$ dispersion and $\varphi_i$ has a $k^2$ dispersion.
-We can also include an interaction term of the form $g(\partial_i \phi - \varphi_i)^2$, which does not break the symmetry.
-The $\varphi_i$ field is well-defined whenever the dipole symmetry is broken.
-
-\note{We can then redefine the $\varphi_i$ field to $\tilde{\varphi}_i = \varphi_i - \partial_i \phi$, so that it is neutral under the symmetry.
-The interaction term becomes $g\tilde{\varphi}_i^2$, so that the $\tilde{\varphi}_i$ field becomes massive.
-The only remaining massless field is $\phi$, which remains charged under the full dipole group and keeps its $k^4$ dispersion.
-As a result, the critical dimension for breaking the full dipole group is not affected by the possibility of partially breaking.
-}
-\charlie{What else to say here? Compare to Griffin/Grosvenor?}
-\todo{Compare to fracton superfluids}
-
-%As long as $\phi_i$ is well-defined, integrating out fluctuations in $\phi_i$ endows the monopole field $\phi$ with a $k^2$ dispersion.
-The coupling $g$ is set by how far the theory sits from the dipole critical point.
-
-%With a $k^2$ dispersion for the monopole field, we now have a route to evade the (generalized) Mermin-Wagner argument and fully break the dipole symmetry in dimensions $d> 2$.
-First, partially break the symmetry so that $\phi_i$ is well defined, which is allowed for any $d>2$.
-Then, we can use the effective dispersion of the $\phi$ field to break the remaining part of the symmetry, also in any $d>2$.
-This procedure can be generalized, so that any multipole symmetry can be fully broken in $d>2$, provided we have enough symmetry-breaking order parameters.
-The discussion also translates to the quantum setting, so that the ultimate critical (space) dimension is always $d=1$.
-
-%The preceding procedure gives us insight into what the generalized Mermin-Wagner arguments were telling us.
-When we wrote down those arguments, we did not include the intermediate fields.
-Thus, the generalized Mermin-Wagner arguments tell us when a direct transition between two phases is or is not allowed.
-Consider again the dipole group $\mathcal{M}^2_\max$.
-The phase in which the symmetry is fully broken can exist in any dimension $d>2$ ($d>1$ for quantum systems), but there can only be direct phase transitions between this phase and the wholly symmetry unbroken phase in $d>4$ ($d>2$).
-In dimensions $2<d<4$ ($1<d<2$) the fully symmetry broken and the fully symmetry unbroken phase must be separated by an intermediate phase that breaks dipole but not monopole symmetry.
-
 \end{document}

### 2021-12-13 multipoleSSB/manuscript.tex
arxiv submission

--- 2021-11-14 multipoleSSB/manuscript.tex
@@
-\documentclass[twocolumn, longbibliography]{revtex4-2}
+\documentclass[pra,aps,twocolumn, amsfonts,amsmath,amssymb,nofootinbib,superscriptaddress]{revtex4-2}
 
@@
 This paper is organized as follows.
-In Section \ref{multipolegroup} we introduce the multipole group, how to build multipole-invariant field theories, and the generalized Mermin-Wagner argument for total breaking of the maximal multipole symmetry.
-In Section~\ref{sec:evade} we show how to evade the generalized Mermin-Wagner argument by considering phases in which multipole symmetries are partially broken.
-We discuss generalized Imry-Ma arguments in the presence of quenched disorder in Section~\ref{sec:disord}.
-Finally, we consider an explicit lattice model that illustrates some of these ideas in Section~\ref{sec:example} before concluding with a discussion of open questions in \ref{sec:disc}.
+In Section \ref{multipolegroup} we introduce the multipole group, and how to build field theories invariant under it.
+In section \ref{sec:full_breaking} we discuss generalized Mermin-Wagner arguments for situations in which a multipole group is spontaneously broken down to the trivial group, while in Section~\ref{sec:partial} we explore the more subtle case where a nontrivial subgroup is preserved.
+This analysis allows us to construct examples where a single symmetry-breaking pattern can be described by several distinct types of Goldstone modes, as well as an example where a continuous symmetry is spontaneously broken in one dimension at $T=0$.
+We then discuss generalized Imry-Ma arguments in the presence of quenched disorder in Section~\ref{sec:disord}.
+Finally, in Section~\ref{sec:example} we consider an explicit lattice model that illustrates some of the ideas of the previous sections.
+We conclude with a discussion of open questions in \ref{sec:disc}.
 
@@
 The multipole group is well-explained in Ref.~\cite{Gromov2019}.
-It generalizes abelian global symmetries such as $\phi(x) \goesto\phi (x) +c$ by allowing a shift by some set of polynomials, $\phi (x) \goesto \phi (x) + \lambda_\alpha P^\alpha(x)$.
+For concreteness, and since we will be interested in situations where a multipolar symmetry group is spontaneously broken, we will describe the multipole group in terms of its action on a compact scalar field $\phi$, which will play the role of a Goldstone boson for the broken multipolar symmetry.
+We will always imagine $\phi$ as constituting the phase mode for some microscopic order parameter $\psi \sim e^{i\phi}$, with $\psi$ transforming by various $U(1)$ phases under the multipole group.
+
+ The multipole group generalizes the internal shift symmetry $\phi(x) \mapsto \phi (x) +c$ by allowing a shift by some set of polynomials involving the spatial coordinates, viz.
+$\phi (x) \mapsto \phi (x) + \lambda_\alpha P^\alpha(x)$.
 The variables $\lambda_\alpha$ are symmetry parameters, while $\alpha$ labels the set of polynomials $P^\alpha(x)$.
@@
 \begin{align}
-P_c^{I_c} = \mu^{I_c}_{i_1\dots i_c}x^{i_1}\dots x^{i_c}, \label{eqn:basis}
+P_c^{\gamma} = \mu^{\gamma}_{I_c}x^{I_c}, \label{eqn:basis}
 \end{align}
-where each  matrix  $\mu^{I_c}_{i_1\dots i_c}$ is fully symmetric and $c\le a$. This group also includes all translations and rotations.
+where $\gamma$ is an index that runs over polynomials of degree $c$ and $c\le a$.
+The composite index $I_c = \{i_1,\dots,i_c\}$  expands as $x^{I_c}=x^{i_1}\cdots x^{i_c}$, and similarly for $\partial_{I_c}$.
+Each  matrix  $\mu^{\gamma}$ is fully symmetric on its $I_c$ indices.
+This group also includes all translations and rotations.
 
-An example of a multipole group that contains all translations and rotations but is not the maximal multipole group is the group generated by shifts of the form
+An example of a multipole group that contains all translations and rotations but is \emph{not} the maximal multipole group is the group generated by shifts of the form
 \begin{align}
-\phi \goesto \phi + \lambda_0 P_0^0 + \lambda_{i} P^{i}_1 + \lambda_{I_2} P^{I_2}_2,
+\phi \goesto \phi + \lambda_0 P_0^0 + \lambda_{i} P^{i}_1 + \lambda_{\beta} P^{\beta}_2,
 \end{align}
@@
 
-We can now consider building field theories that are invariant under various multipole groups.
-Reference~\cite{Gromov2019} describes the process in detail, so we will only sketch the outline.
-References~\cite{Griffin2015, Gromov2019} discuss quantum field theories where configurations are weighted by an action.
-We will also want to consider thermal field theories with Hamiltonian descriptions, so we will abstractly discuss ``kinetic terms" in this section.
+We can now consider building IR field theories for phases in which a multipolar symmetry is completely spontaneously broken, with no residual unbroken subgroup.
+Reference~\cite{Gromov2019} describes the process in detail, so in the following we will be somewhat succinct.
+We will continue to write things down in terms of the compact scalar $\phi$, which (at least in the cases for which SSB is not preempted by strong fluctuations) is to be viewed as the Goldstone for the sponatneously broken symmetry.
 
-The key step is to be able to find spatial derivatives $D$ that annihilate all the polynomials $P_a^{I_a}$ that we want to include in our multipole group.
-We say that $D$ is order-$s$ if it takes the form
+Since we are only interested in the low-energy physics of the putative symmetry-broken phase, constructing an appropriate IR field theory amounts to nothing more than constructing an appropriate kinetic term for $\phi$.
+To find a kinetic term invariant under a given multipole group $\mathcal{M}$, we need to find operators $D$ built out of spatial derivatives that annihilate all the polynomials in $\mathcal{M}$.
+If $a_\text{max}$ is the highest degree of the polynomials in $\mathcal{M}$, the simplest derivative operators which generically do the job are of the form
 \begin{align}
-D = q^i\partial_i + q^{ij}\partial_1\partial_j + \cdots + q^{i_1\dots i_s}\partial_{i_1}\dots\partial_{i_s},
+D = q^{I_{a_{\rm max}+1}} \p_{I_{a_{\rm max}+1}},
 \end{align}
-where each $q$ is a symmetric tensor.
-Although it is not generically possible~\cite{Gromov2019}, we can sometimes find a set of $D_\alpha$ ($\alpha$ is an abstract index) with $s\le a_{\text{max}}$, where $a_\text{max}$ is the highest degree of the polynomials.
+for some symmetric tensor $q^{I_{a_{\rm max}+1}}$.
+Although it is not generically possible to do so~\cite{Gromov2019}, we can sometimes find a set of $D_\alpha$ ($\alpha$ is an abstract index) with $s\le a_{\text{max}}$, where $a_\text{max}$ is the highest degree of the polynomials.
 In this case, the effective field theory will be invariant under some non-maximal multipole symmetry.
 
-Generically, we cannot find any $D$ with $s\le a_\text{max}$ such that $DP_a^{I_a} = 0$ for all the polynomials we want. In that case we can still write down a theory with derivatives 
+Writing the invariant derivative operators as $D_\alpha$, the most general kinetic term is ~\cite{Gromov2019} 
 \begin{align}
-D = q^{i_1\dots i_n}\partial_{i_1}\dots \partial_{i_n},
+K[\phi(x)] &= g_{\alpha\beta} (D_\alpha \phi) (D_\beta \phi)
 \end{align}
-with $n=a_\text{max}+1$. Then, the field theory will be invariant under the maximal multipole group of degree $a_\text{max}$.
-
-Armed with the invariant derivatives $D_\alpha$ ($\alpha$ is another abstract index), we can start to construct a kinetic term.
-The most general term, at lowest order in derivatives, would be~\cite{Gromov2019}
-\begin{align}
-K[\phi(x)] &= g_{\alpha\beta} (D_\alpha \phi) (D_\beta \phi).
-\end{align}
+for some symmetric tensor $g_{\alpha\beta}$.
 We will often write the Fourier transform of the kinetic term as $\phi_{-k} K_k \phi_k$.
@@
 This is the sense in which the Haah group mentioned earlier corresponds to a field theory for Haah's code~\cite{BB}.
-See Ref.~\cite{Gromov2019} for the full story.
+See Ref.~\cite{BB, Gromov2019} for the full story.
 
-\subsection{Generalized Mermin-Wagner argument}
+\section{Generalized Mermin-Wagner: full multipole breaking}\label{sec:full_breaking}
 
-We are now ready to describe the generalized Mermin-Wagner argument for an arbitrary maximal multipole group.
+We are now ready to describe the generalized Mermin-Wagner argument in the case where an arbitrary maximal multipole group is spontaneously broken down to the trivial subgroup.
 This already appears in Ref.~\cite{Griffin2015}, so we are simply reviewing it here in preparation for the more generic cases to follow.
-Here, we will discuss both classical and quantum systems.
+Here, we will discuss both thermal and $T=0$ systems.
 
-Consider a clean classical system with a spontaneously broken maximal multipole symmetry of degree $a=n-1$.
-Then the kinetic term is
+Consider first a clean system at $T>0$ with a spontaneously broken maximal multipole symmetry of degree $a$.
+Then the kinetic term is proportional to 
 \begin{align}
-K[\phi(x)] = (\partial_{i_1} \dots \partial_{i_n} \phi)^2, \label{eqn:adisp}
+K[\phi(x)] = (\partial_{I_{a+1}} \phi)^2. \label{eqn:adisp}
 \end{align} 
-where $\phi$ is the order parameter field. 
-Now assume we are in an ordered phase, but begin to nucleate domains of linear size $L$.
-Since the order parameter is continuous there will be a thick domain ``wall", in which the order parameter changes gradually from its value in the background phase to its value in the domain.
+Now consider nucleating domains of linear size $L$ in which the expectation value of the order parameter --- namely $\langle e^{i\phi}\rangle$ --- varies by an amount of order 1.
+Since the order parameter is continuous there will be a thick domain ``wall" over which the order parameter changes gradually from its value in the background phase to its value in the domain.
 The thickness will be of order $L$.
@@
 As before, we interpret the divergence of the correlation function as a sign that the symmetry cannot be spontaneously broken.
-This allows us to reproduce our scaling argument that spontaneous symmetry breaking of the multipole group $\mathcal{M}^a_\max$ cannot occur for $d\le2(a+1)$.
+This allows us to reproduce our scaling argument that SSB of the multipole group $\mathcal{M}^a_\max$ cannot occur for $d\le2(a+1)$.
 
-If we look at the quantum (zero temperature) setting, the energy scaling no longer tells us where the critical dimension is.
+If we look at the zero temperature setting, the energy scaling no longer tells us where the critical dimension is.
 This is because even when there is no energy cost to forming domains, there is no entropy gain.
 Instead, quantum fluctuations must be the motivation for domain nucleation.
-To find the quantum critical dimension, we calculate the correlation function,
+To find the quantum critical dimension, we calculate the correlation function by way of the (imaginary-time) IR Lagrangian
 \begin{align}
-C(x) = \left\langle \phi(x) \phi(0) \right\rangle &\sim \int \frac{d\omega d^dk}{\omega^2 + k^{2( a + 1 )}},
+     \mathcal{L}[\phi] = (\partial_\tau\phi)^2 + K[\phi].
+\end{align} 
+This gives 
+\begin{align}
+\label{tzerocorr}C(x) = \left\langle \phi(x) \phi(0) \right\rangle &\sim \int \frac{d\omega d^dk}{\omega^2 + k^{2( a + 1 )}},
 \end{align}
 which now diverges at $d\le a+1$~\cite{Griffin2015}, where $d$ is the number of spatial dimensions.
-We have halved the critical dimension, so that spontaneous symmetry breaking cannot occur at $d\le a+1$.
-It is interesting to note that the dynamical critical exponent in these theories is $z = a+1$, so the classical and quantum critical dimensions are related by $d_\text{cl} = d_\text{q} + z$.
+We have halved the critical dimension, so that SSB cannot occur at $d\le a+1$.
+Note that the dynamical critical exponent in these theories is $z = a+1$, so that as expected, the classical and quantum critical dimensions are related by $d_\text{cl} = d_\text{q} + z$.
 
@@
 \end{align}
-In a classical system, spontaneous symmetry breaking cannot occur if $d_\eff\le 2$, while in a quantum system the critical dimension is $d_\eff = 1$.
+For $T>0$, we then conclude that SSB cannot occur if $d_\eff\le 2$, while at $T=0$ the critical dimension is $d_\eff = 1$.
 
-We should note that the statements in this section were framed in terms of the dispersion of the Goldstone modes rather than the structure of the symmetry group.
+We should emphasize that the statements in this subsection were framed in terms of the dispersion of the Goldstone modes rather than the structure of the symmetry group.
 It is always possible to find the dispersion given the multipole symmetry group, but it may require some basis changes (as in the Haah multipole group).
 
-\section{Evading the generalized Mermin-Wagner argument} \label{sec:evade}
+\section{Partial breaking of multipole symmetries} \label{sec:partial}
 
-\subsection{Breaking to a subgroup} \label{sub:subgroup}
+In the previous section we studied situations in which a multipole group was fully broken.
+In this section we turn our attention to examining what happens when the symmetry breaking is incomplete, with a nontrivial subgroup remaining unbroken.
+A simple example that illustrates why this problem is nontrivial is the following.
 
-In the previous section we assumed we had an order parameter field that fully broke the multipole symmetry.
-Let us now consider the case where the order parameter breaks the symmetry from the original group $G$ to some subgroup $H$.
-For simplicity, we will let $G$ be the maximal multipole group $\mathcal{M}^{a}_\text{max}$ and $H$ be $\mathcal{M}^b_\text{max}$, with $b<a$.
+Consider a situation in which a dipolar symmetry $\mcm^1_{\rm max}$ is spontaneously broken to its monopolar subgroup $\mcm^0$, and specialize to the case of $d=3$ and $T>0$.
+In the spirit of Goldstone's theorem, the most naive thing to do when analyzing the symmetry broken phase would be to write down an action involving a set of $d=3$ Goldstone fields $\theta_j$, with $\t_j$ transforming linearly under the dipolar part of the symmetry group, but remaining invariant under the unbroken monopole subgroup, so that $\theta_j \mapsto \theta_j + \beta_j$ under a transformation parameterized by $P(x) = \alpha + \beta_j x^j$.
+In this case, the IR field theory for the putative symmetry-breaking phase would be of the form
+\begin{align}
+\label{vector} \mathcal{L}= \sum_j  (\partial_\tau \theta_j)^2 + \sum_{i,j} g_{ij} (\partial_i \theta_j)^2.
+\end{align}
+As we are working in $d=3$ and at $T>0$, the theory \eqref{vector} indeed has SSB for the dipole symmetry, with $e^{i\t_j}$ developing long-range order.
 
-The simplest case is when $a=1$ and $b=0$.
-In physical terms, we are taking a system that conserves total charge and total dipole moment, then condensing objects that carry a dipole moment but no total charge.
-We then recognize that these fundamental dipoles, which are of course the dipole charges of the monopole symmetry, are also the monopole charges of the dipole symmetry.
-We should then expect symmetry breaking cannot occur for $d\le 2$ classically or $d\le 1$ quantum mechanically.
+The theory \eqref{vector} is not the only possibility consistent with this pattern of symmetry breaking, however. Indeed, consider instead the scalar field theory 
+\begin{align}
+\label{singlescalar}    \mathcal{L} = (\partial_\tau\phi)^2 + g_{ij} (\partial_i\partial_j\phi)^2, \end{align}
+where $\phi$ transforms under the symmetry as $\phi \mapsto \phi + \alpha + \beta_jx^j$. 
+ In the scalar theory \eqref{singlescalar}, the analysis of section \ref{sec:full_breaking} shows that the monopole symmetry is {\it not} spontaneously broken, with $e^{i\phi}$ possessing short-ranged correlation functions.
+However, the dipolar symmetry does indeed have SSB, and it is easy to check that the operators $e^{i\p_j\phi}$ --- which transform under the symmetry in the same way as $e^{i\t_j}$ --- have long-range order.
+Thus in this case, the two theories \eqref{vector} and \eqref{singlescalar} exhibit the same pattern of symmetry breaking, despite having a different number of Goldstone modes.\footnote{We will refer to the mode $\phi$ in the theory \eqref{singlescalar} as the ``Goldstone mode'' of the spontaneously broken dipole symmetry, since it is the only mode in the theory.
+Note however that it is $\p_j\phi$ (and not $\phi$ itself) which transforms under the symmetry action in the way expected of Goldstone bosons.} In order to formulate the correct version of the Mermin-Wagner theorem in the partial breaking case, it will be necessary to understand why this is so.
 
-It is interesting to note that condensing an object that carries dipole charge must break rotation symmetry.
-In Sec.~\ref{sec:example} we show how to break the dipole symmetry while preserving a lattice rotation symmetry by considering multiple species of dipole.
-We are not aware of a way to condense a dipole while preserving continuous rotation symmetry.
+\subsection{Goldstone counting} \label{sub:subgroup}
 
-In more generality, the order parameter for this pattern of symmetry breaking will be an object $\phi_{i_1\dots i_m}$, where $m=b+1$.
-Because $G$ is a maximal multipole group, we can still write the polynomials as in Eqn.~\ref{eqn:basis}.
-For our new order parameter, they act as
-\begin{align}
-\delta \phi_{i_1 \dots i_m} = \lambda_{I_c} \partial_{i_1} \dots \partial_{i_m} P_c^{I_c}, \label{eqn:partial}
-\end{align}
-so that $\phi_{i_1 \dots i_m}$ is invariant under $H=\mathcal{M}^b_\max$.
+That the number of Goldstone modes is not uniquely determined by the pattern of symmetry breaking is of course not in contradiction with Goldstone's theorem, as illustrated e.g.
+by the example of the ferromagnet vs the antiferromagnet.
+Indeed, correctly determining the number of Goldstone modes realized in a given situation depends not just on the symmetry breaking pattern, but also on information relating to how the various charge generators act on the ground state, as nicely described in \cite{watanabe2013redundancies}.
 
-The kinetic term should take the form
-\begin{align}
-K[\phi(x)] &= (\partial_{i_{m+1}}\dots \partial_{i_n} \phi_{i_1 \dots i_m})^2,
-\end{align}
-which is invariant under shifts of the form of Eq.~\ref{eqn:partial}.
-The dispersion for the Goldstone modes is $K_k \sim k^{2(n-m)} = k^{2(a-b)}$.
-Following the correlation function calculation above, we find a classical critical dimension of $d = 2(a-b)$ and a quantum critical dimension of $d = a-b$.
-We recover the maximal generalized Mermin-Wagner argument if we let $b=-1$ denote the trivial group ($b=0$ denotes the monopole group).
+A general formal framework for understanding the counting of Goldstone modes in the present context can be developed by following the analysis of \cite{watanabe2013redundancies}, but here we will simply content ourselves with a heuristic physical argument.
+For simplicity, we will focus on systems in which the total symmetry group $\mcm_{\rm max}^a$ is spontaneously broken down to some nontrivial subgroup.
 
-\subsection{Mediated condensation} \label{sub:mediated}
+Let us define the compressibilities $\kappa_c$ by 
+\be \label{kappadef} \kappa_c^{I_c} = \frac{d\langle n_c^{I_c}\rangle}{d\mu_c^{I_c}},\ee 
+where $n_c$ is the number density of $c$-pole charges, $\mu_c$ is a conjugate chemical potential, and $I_c = \{i_1,\dots,i_c\}$ is a composite index as before.
 
-We have seen that the generalized Mermin-Wagner argument provides critical dimensions for partial breaking that are lower than the critical dimensions for full symmetry breaking.
-In fact, we can use this observation to evade the generalized Mermin-Wagner theorem, completely breaking any multipole symmetry for $d>2$ classically, or $d>1$ quantum mechanically.
-This possibility has been pointed out in Refs.~\cite{Griffin2015Cascade, Griffin2015}, but we include a further interpretation in terms of what types of phase transitions are allowed in different dimensions.
-For simplicity, we will begin by considering the classical dipole-preserving theory with symmetry group $\mathcal{M}^2_\max$.
+Let $b\leq a$ denote the smallest degree of multipole charge with nonzero compressibility, i.e.
+consider a situation in which $\kappa_c>0$ for all $b\leq c\leq a$, while $\kappa_c=0$ for all $c<b$.
+Then the minimal IR Lagrangian for the symmetry broken phase looks schematically like
+\be \label{generall} \mathcal{L} = \kappa_b (\partial_\tau \phi_{I_b})^2 + g_{J_{a-b+1}}(\partial_{J_{a-b+1}}\phi_{I_b})^2,\ee 
+where 
+$\p_{J_{a-b+1}} \equiv \p_{j_1} \cdots \p_{j_{a-b+1}}$, and where $\phi_{I_b}$ is a field that shifts by a polynomial of degree $a-b$ under the action of $\mmax a$.
+The compressibility appears as the coefficient of the time derivative term since $\p_\tau \phi_{I_b}$ is the momentum conjugate to $\phi_{I_b}$ in the quantum theory, and hence correlation functions of fluctuations in $n_b$ are determined by those of $\p_\tau \phi_{I_b}$.
 
-To be explicit, consider a monopole field $\phi(x)$ and a dipole field $\phi_i(x)$. Under the dipole symmetry they transform as 
-\begin{align}
-\phi(x) &\goesto \phi(x) + a_ix^i + b,\nn
-\phi_i(x) &\goesto \phi_i(x) + a_i,
-\end{align}
-where we can more explicitly see that $\phi(x)$ transforms like a monopole field under the dipole part of the symmetry. The kinetic term will look like
-\begin{align}
-K[\phi,\phi_i] &= (\partial_i \partial_j \phi)^2 + (\partial_i \phi_j)^2,
-\end{align}
-so that $\phi$ has a $k^4$ dispersion and $\phi_i$ has a $k^2$ dispersion.
-We can also include an interaction term of the form $g(\partial_i \phi - \phi_i)^2$, which does not break the symmetry.
-The $\phi_i$ field is well-defined whenever the dipole symmetry is broken.
-As long as $\phi_i$ is well-defined, integrating out fluctuations in $\phi_i$ endows the monopole field $\phi$ with a $k^2$ dispersion.
-The coupling $g$ is set by how far the theory sits from the dipole critical point.
+A necessary condition for $\mmax b$ to be spontaneously broken is that $\kappa_b \neq 0$.
+Indeed, if $\mmax b$ is spontaneously broken then $n_b$ creates Goldstones when acting on the ground state, thereby ensuring that the density-density correlator appearing in the calculation of \eqref{kappadef} must be nonzero at long wavelengths.
 
-With a $k^2$ dispersion for the monopole field, we now have a route to evade the (generalized) Mermin-Wagner argument and fully break the dipole symmetry in dimensions $d> 2$.
-First, partially break the symmetry so that $\phi_i$ is well defined, which is allowed for any $d>2$.
-Then, we can use the effective dispersion of the $\phi$ field to break the remaining part of the symmetry, also in any $d>2$.
-This procedure can be generalized, so that any multipole symmetry can be fully broken in $d>2$, provided we have enough symmetry-breaking order parameters.
-The discussion also translates to the quantum setting, so that the ultimate critical (space) dimension is always $d=1$.
+However, a nonzero $\kappa_b$ does not {\it necessarily} imply that $\mmax b$ is spontaneously broken.
+$\kappa_b\neq 0$ means that the system possesses gapless modes created by $n_b$, but it is possible for fluctuations to be strong enough so that these modes cannot be associated with the Goldstone modes of a spontaneously broken $\mmax b$.
+Thus the theory \eqref{generall} does not necessarily describe a phase where $\mmax b$ is spontaneously broken, and the exact pattern of symmetry breaking depends on the spatial dimension and whether or not $T$ is nonzero.
+For example, we can again consider the theory in \eqref{singlescalar} at $d=3, T>0$: even though the charge compressibility $\kappa_0$ (the coefficient of $(\p_\tau\phi)^2$ in \eqref{singlescalar}) is nonzero, $\phi$ fluctuates strongly enough that the monopole symmetry is not spontaneously broken.
 
-The preceding procedure gives us insight into what the generalized Mermin-Wagner arguments were telling us.
-When we wrote down those arguments, we did not include the intermediate fields.
-Thus, the generalized Mermin-Wagner arguments tell us when a direct transition between two phases is or is not allowed.
-Consider again the dipole group $\mathcal{M}^2_\max$.
-The phase in which the symmetry is fully broken can exist in any dimension $d>2$ ($d>1$ for quantum systems), but there can only be direct phase transitions between this phase and the wholly symmetry unbroken phase in $d>4$ ($d>2$).
-In dimensions $2<d<4$ ($1<d<2$) the fully symmetry broken and the fully symmetry unbroken phase must be separated by an intermediate phase that breaks dipole but not monopole symmetry.
+\subsection{Generalized Mermin-Wagner: partial multipole breaking}
+
+In the previous subsection, we saw that the number of Goldstone modes in the symmetry-breaking theory is determined by the smallest degree of multipole charge $b$ such that $\kappa_b\neq0$.
+We now turn to determining the symmetry breaking pattern that occurs for a given choice of $b$, spatial dimension $d,$ and temperature $T$, thereby allowing us to formulate a generalized version of the Mermin-Wagner theorem for the case of partial multipole breaking.
+
+Since $\kappa_c=0$ if $c<b$, none of the subgroups $\mmax {c<b}$ can be spontaneously broken.
+Suppose then that the symmetry is spontaneously broken down to $\mmax {c}$, where $c\geq b$.
+To find the allowed values of $c$ consistent with this assumption, we need to calculate correlation functions of operators of the form $\exp(i\p_{J_{c-b}}\phi_{I_b})$, these being the operators transforming nontrivially under $\mmax c$ with the longest-range correlation functions.
+We have
+\be\ba \langle \p_{J_{c-b}}\phi_{I_b}(x) \p_{J_{c-b}}\phi_{I_b}(0)\rangle & \sim \int \frac{d\omega \, dk\, k^{d-1+2(c-b)}}{\omega^2 + k^{2(a-b+1)}} \\ 
+& \sim \int dk\, k^{d-1+2(c-b) - \varsigma_T(a-b+1)}, \ea\ee 
+where we have defined the function $\varsigma_T$ such that $\varsigma_T = 1$ if $T=0$ and $\varsigma_T = 2$ if $T>0$.
+Consequently, we find that $\mmax c$ can be spontaneously broken only in dimensions $d$ such that
+\be \label{pbreak_mw} d>2(b-c) + \varsigma_T(a-b+1).\ee 
+For a given choice of symmetry breaking pattern (captured by $a$ and $c$) and a given choice of which compressibilities are nonzero (captured by $b$), the quantity on the RHS determines the lower critical dimension for symmetry breaking, thereby generalizing the Mermin-Wagner theorem to the partial breaking case.
+As a sanity check on \eqref{pbreak_mw}, note that when we set $c=b=0$ (full multipole breaking) we recover the lower critical dimension of $\varsigma_T(a+1)$ obtained in the previous section.
+Also note that as expected, the lower critical dimensions in the quantum $(\varsigma_T = 1)$ and classical $(\varsigma_T = 2)$ cases are related by $d_{\rm cl} = d_{\rm q} + z$, where $z=a-b+1$ is the dynamical exponent of the Goldstone theory \eqref{generall}.
+
+Finally, as an interesting example, consider the case where the dipolar group $(a=1)$ is broken down to the monopole group $(b=0,c=1)$.
+At $T=0$, we see from \eqref{pbreak_mw} that this is possible provided that $d>0$.
+Thus the partial breaking of dipole symmetry provides us with an example where a continuous symmetry can be spontaneously broken in $d=1$ at zero temperature.
 
@@
 The Imry-Ma argument only appeals to energy considerations, so the quantum critical dimension in the presence of disorder remains the same as the classical critical dimension~\cite{Vojta2013}.
-In disordered quantum systems, spontaneous symmetry breaking is impossible for $d\le 4(a+1)$.
+In disordered quantum systems, full SSB is impossible for $d\le 4(a+1)$.
 
 To reproduce this argument using a correlation function calculation~\cite{ImryMa} we need to calculate the response to disorder.
-Call the disorder field $\theta(x)$ and redefine $\phi(x)$ to be the variation away from the average value $\phi_0$.
+Call the disorder field $h(x)$ and redefine $\phi(x)$ to be the variation away from the average value $\phi_0$.
 The relevant part of the Hamiltonian is
 \begin{align}
-H &\sim \int d^dx \left[ \half K[ \phi(x)] - h \phi(x) \theta(x) \right] \nn
-&\sim \int d^dk \left[ \half \phi_{-k} K_k \phi_k -h \phi_{-k} \theta_k \right] \nn
-&\sim \int d^dk \left[ \half \phi_{-k}' K_k \phi_k' - \half h^2 \theta_{-k} K^{-1}_k \theta_k \right],
+H &\sim \int d^dx \left[ \half K[ \phi(x)] - \phi(x) h(x) \right] \nn
+&\sim \int d^dk \left[ \half \phi_{-k} K_k \phi_k - \phi_{-k} h_k \right] \nn
+&\sim \int d^dk \left[ \half \phi_{-k}' K_k \phi_k' - \half h_{-k} K^{-1}_k h_k \right],
 \end{align}
-where $h$ sets the strength of the disorder, and we have written $\phi_k' = \phi_{-k} - h K^{-1}_k \theta_{-k}$. Then, from the expectation of $\phi_k$,
+where we have written $\phi_k' = \phi_{-k} - K^{-1}_k h_{-k}$. Then, from the expectation of $\phi_k$,
 \begin{align}
-\langle \phi_k \rangle &\sim \frac{1}{h} \frac{\delta Z}{\delta \theta_{-k} } \nn
-&\sim h K_k^{-1} \theta_{k},
+\langle \phi_k \rangle &\sim \frac{\delta Z}{\delta h_{-k} } \nn
+&\sim K_k^{-1} h_{k},
 \end{align}
@@
 A more general case is a system with symmetry $G = \mathcal{M}_\max^a$ and disorder that breaks the symmetry to $\mathcal{M}_\max^b$ but preserves $G$ on average.
-In this system the critical dimension should be $d=4(a-b)$ for both classical and quantum systems.
+In this system the critical dimension for long-range order of a field that breaks $\mmax{a}$ but preserves $\mmax{b}$ should be $d=4(a-b)$ for both classical and quantum systems.
 Again, let $b=-1$ denote the trivial group in order to recover our previous results.
-
-As in the clean systems, we can have a sequence of phase transitions separated by intermediate phases in which the multipole symmetry is broken to a progressively smaller subgroup.
-With enough order parameter fields, any multipole symmetry can be fully broken in dimension $d>4$, regardless of which order parameter the disorder couples to.
-The same is true in quantum systems.
-Whether a certain phase transition is allowed or whether there needs to be an intermediate phase depends on the spatial dimension and the nature of the disorder.
 
@@
 &\qquad - \sum_{j,\mu,\nu} t_b \left[  b_j b^{\dag}_{j+\mu} b^{\dag}_{j+\nu} b_{j+\mu+\nu} + H.C. \right]\nn
-H_d &= \sum_{j,\mu} \left[ \frac{U_d}{2} n'_{j,j+\mu} (n'_{j,j+\mu} - 1) -\mu \, n'_{j, j + \mu} \right]\nn
+H_d &= \sum_{j,\mu} \left[ \frac{U_d}{2} n^d_{j,j+\mu} (n^d_{j,j+\mu} - 1) -\mu_d \, n^d_{j, j + \mu} \right]\nn
 &\qquad  - \sum_{j,\mu,\nu} t_d \left[d^{\dag}_{j, j + \mu} d_{j + \nu, j+\mu + \nu} +H.C.\right]\nn
@@
 where sums are taken over sites $j$ and lattice directions $\mu$ and $\nu$.  
-The operator $b^\dagger_j$ creates a boson on site $j$ while $d^\dagger_{j,j+\mu}$ creates an boson on the edge between $j$ and $j+\mu$.
 For simplicity, let us set 
@@
 so there is a robust insulating phase at weak hopping $t_b, t_d$~\cite{Fisheretal}.
+A detailed discussion of the phase diagram and phenomenology of this model will be given in~\cite{dbhm}; in the following we will simply discuss the symmetry breaking patterns realized in various different limits.
 
 The Hamiltonian $H_b$ controls condensation of the site bosons.
-When $t_b$ is small, the site bosons are in number eigenstates while when $t_b$ is large the site bosons condense.
+When $t_b$ is small, the site bosons are in number eigenstates while when $t_b$ is large the site bosons condense (although note that since the term proportional to $t_b$ is quartic in the site boson operators, the condensation transition may be nonstandard).
 Similarly, the edge bosons condense when $t_d$ is large.
-Note that all $d$ species of edge boson will condense equally, so that lattice rotation symmetry is not broken.
+Note that lattice rotation symmetry will be broken if only a subset of the edge bosons condense, while if all $d$ species of condense equally then lattice rotation symmetry is preserved.
+We are not aware of a way to condense objects with dipolar charge while preserving {\it continuous} rotation symmetry.
+
 The term $H_\text{int}$ is allowed by the symmetry because it simultaneously removes an edge boson and creates a dipole of site bosons.
-When the edge bosons are condensed, $H_\text{int}$ gives the site bosons an effective single particle hopping, and hence an effective $k^2$ dispersion, which would otherwise be disallowed by symmetry.
+Note that when the edge bosons are condensed, $H_\text{int}$ gives the site bosons an effective single particle hopping: thus condensation of single bosons can either proceed directly (by making $t_b$ large) or indirectly (by first condensing dipoles, and making the effective single-particle hopping term large).
 
-At sufficiently high dimension, this system will support three phases, which may be loosely described as the Mott insulator, the monopole superfluid where site bosons are condensed and the dipole superfluid where edge bosons (but not site bosons) are condensed.
-Note that the dipole superfluid does preserve lattice rotations.
-In the monopole superfluid phase, the effective Hamiltonian contains bare creation and annihilation terms for the edge bosons, which automatically breaks number conservation of the edge bosons, and effectively induces a dipole superfluid as well.
-We can identify these phases with symmetry breaking patterns of the multipole group.
-In the trivial phase the group is unbroken.
-In the dipole superfluid, the group is broken to its monopole subgroup.
-The monopole superfluid then corresponds to the fully broken symmetry.
+Consider first the phase in which only the edge bosons have condensed, with the site bosons remaining gapped.
+In this case the dipole compressibility $\kappa_1$ is nonzero, while the charge compressibility vanishes, $\kappa_0=0$.
+The condensed phase is therefore described in the IR by the $z=1$ theory \eqref{vector}, and the dipolar symmetry is spontaneously broken down to the monopole subgroup provided that $d>1$ (for $T=0$) or $d>2$ (for $T>0$).
 
-In the absence of disorder, neither symmetry breaking pattern can occur for $d\le 1$. There may still be interesting phases, such as a Luttinger liquid.
-For $d>2$ dipole and monopole symmetry breaking may occur, so all three phases are intact. Furthermore, transitions between any two phases are allowed.
-This leaves the intermediate case of $d=2$, in which all three phases can exist, but there can be no transitions between any symmetry-unbroken phase and the monopole superfluid.
-Instead, there must be an intermediate phase with dipole but not monopole superfluidity.
-Representative single-parameter phase diagrams are shown in Fig.~\ref{fig:phases}.
-
-\begin{figure}
-\centering
-
-\begin{tikzpicture}[scale = .9]
-\draw[<->] (0,0) node[left]{(a)}  -- (8,0) node[right]{$s$};
-\draw (4,-.1) -- (4,.1);
-
-\node [align=center] (A) at (2, -0.5) {unbroken};
-\node [align=center] (C) at (6, -0.5) {fully\\broken};
-
-\end{tikzpicture}
-\begin{tikzpicture}[scale = .9]
-
-\draw[<->] (0,0) node[left]{(b)}  -- (8,0) node[right]{$s$};
-\draw (2.667,-.1) -- (2.667,.1);
-\draw (5.333,-.1) -- (5.333,.1);
-
-\node [align=center] (A) at (1.333, -0.5) {unbroken};
-\node [align=center] (B) at (4, -0.5) {dipole\\broken};
-\node [align=center] (C) at (6.667, -0.5) {fully\\broken};
-\end{tikzpicture}
-
-\caption{(a) For  $d>2$, there can be direct transitions between symmetry-unbroken and the fully-broken phase.
-Here $s$ is intended to represent a parameter along an arbitrary line in the phase diagram that crosses this transition.
-(b) At $d=2$, although the symmetry can be fully broken, the fully broken phase cannot touch the unbroken phase.
-Any line through the phase diagram that connects the two phases must pass though the dipole-broken phase.
-These phase diagrams are for translation invariant systems at zero temperature.}
-\label{fig:phases}
-\end{figure}
+Next, consider the phase where the site bosons have condensed.
+In this case both $\kappa_1,\kappa_0$ are nonzero, and deep in the condensed phase the IR physics is described by the $z=2$ theory \eqref{singlescalar}.
+The dipolar symmetry is spontaneously broken in $d>0$ if $T=0$ and in $d>2$ at $T>0$, while the monopole subgroup is spontaneously broken in $d>2$ at $T=0$ and $d>4$ at $T>0$.
+This model therefore provides us with a way of realizing symmetry-broken phases described by both of our earlier example theories \eqref{vector} and \eqref{singlescalar}.
 
@@
 \end{align}
-where $h_b$ and $h_d$ control the magnitude of the disorder and each instance of $\sigma$ is a random phase. We will always consider $h_b,h_d<<1$.
+where $h_b$ and $h_d$ control the magnitude of the disorder and each instance of $\sigma$ is a random phase. We will always consider $h_b,h_d\ll1$.
 
-At first, we will set $h_d=0$ but turn on $h_b$.
-This corresponds to the main considerations from Sec.~\ref{sec:disord}, where the disorder couples to the order parameter that fully breaks the symmetry.
-This raises the critical dimension for the superfluidity of the site bosons.
-The edge bosons may still enter a superfluid phase for $d>1$, but the site bosons may not condense with long range phase order until $d>4$.
-There cannot be a direct transition from the insulator to the monopole superfluid until $d>8$.
-A phase akin to a Bose glass \cite{Fisheretal} of monopoles, where site bosons condense but without long-range order, can still exist in $1\le d<4$.
-
-Now, let $h_d\ne 0$ while keeping $h_b=0$.
-The Imry-Ma argument for the edge bosons says that they should not be able to enter a superfluid phase when $d\le 4$.
-Once again, we can enter a Bose glass phase where $\langle d_{j, j+\mu} \rangle \ne 0$.
-In this case, the site bosons acquire a random single particle {\it hopping}.
-This may be expected to give rise to a Bose glass phase also for the site bosons.
-This implies that the monopoles cannot even undergo mediated condensation to superfluidity for $d\le 4$.
-For $d>4$, the dipoles may enter a superfluid phase, and the monopoles can enter a superfluid phase without mediation.
-
-It is difficult to justify including including one disorder field but not the other.
-The standard reason to forbid disorder is by imposing translation invariance, but if we are already including one disorder field we are necessarily breaking translation symmetry.
 For $h_b,h_d\ne 0$, we can fully rely on the Imry-Ma argument.
-No symmetry breaking can occur for $d\le 4$, the monopole superfluid must be separated from any symmetry-unbroken phase by a dipole superfluid for $4<d\le 8$, and any symmetry-breaking transition can occur for $d>8$.
+No symmetry breaking can occur for $d\le 4$, the dipole part of the symmetry may be broken for $4<d\le 8$, and any symmetry-breaking phase can occur for $d>8$.
 
@@
 
-In this paper we analyzed the spontaneous symmetry breaking of various multipole groups, and discussed generalized Mermin-Wagner theorems, which (we argued) should be best understood as constraints on when a direct transition fully breaking a multipole group can occur, and when a symmetry unbroken phase and a (monopole) symmetry breaking phase must be separated by intermediate phases breaking higher multipole symmetries.
+In this paper we analyzed the spontaneous symmetry breaking of various multipole groups and discussed generalized Mermin-Wagner theorems for breaking a maximal multipole group, either fully or to a subgroup.
 We also considered multipole groups that are not the maximal multipole group, and the effect of quenched disorder.
@@
 Of course, we could consider further combinations of effects.
-We could spontaneously break a symmetry from a group $G$ to a subgroup $H$ where either $G$ or $H$ are non-maximal.
-Or we could consider non-maximal groups with disorder.
-While the number of potential examples to consider is unlimited, they should all be analyzable using the ideas introduced herein.
+For example, we could spontaneously break a symmetry from a group $G$ to a subgroup $H$, where one or both of $G,H$ are non-maximal; we could also consider non-maximal groups with disorder.
+While the number of potential examples to consider is large, they should all be analyzable using the ideas introduced herein.
 
 We should be clear that the arguments in this paper are Mermin-Wagner or Imry-Ma arguments. 
-These essentially amount to stability analyses about the Gaussian symmetry breaking fixed point.
-In principle, even when the Gaussian symmetry breaking fixed point is unstable, a non-trivial fixed point with long range order could arise (see e.g.
+In principle, even when the Gaussian symmetry breaking fixed point is unstable to fluctuations, a non-trivial fixed point with long range order could arise (see e.g.
 \cite{TonerRadzihovsky}).
@@
 For $n=2$, there can in addition be quasi-long-range-ordered phases.
-Determining what kind of phase {\it can} obtain in the absence of long range ordered symmetry breaking is beyond the scope of this work, and would at a minimum require understanding the nature and role of topological defects in the symmetry breaking order parameter, akin to vortices in the XY model.
-The range of possible symmetry unbroken phases could be even richer in the presence of disorder, where various glassy phases could also come into play \cite{Fisheretal}.
+Determining what kind of phase {\it can} be obtained in the absence of long range ordered symmetry breaking would at a minimum require understanding the nature and role of topological defects in the symmetry breaking order parameter, akin to vortices in the XY model.
+We expect that if we stick with an $O(2)$ internal group, then quasi-long range order should always be a possibility at the marginal spatial dimension, at least in the limit where the vortex core energy is infinitely large.
+The extension to non-Abelian internal groups is beyond the scope of this work.
+The range of possible symmetry unbroken phases could also be even richer in the presence of disorder, where various glassy phases could also come into play \cite{Fisheretal}.
 Our discussion of disorder physics was also limited to quenched short-range correlated disorder.
@@
 the U.S. Department of Energy, Office of Science, Basic
-Energy Sciences, under Award \# DE-SC0021346
+Energy Sciences, under Award \# DE-SC0021346. E.L. was supported by a Hertz Fellowship.
 

### 2022-03-08 Presentation/MultipoleMW.tex
presentation for andys group meeting

\documentclass{beamer}

\setbeamercolor{palette primary}{bg=black,fg=white}
\setbeamercolor{palette secondary}{bg=black,fg=white}
\setbeamercolor{palette tertiary}{bg=black,fg=black}
\setbeamercolor{palette quaternary}{bg=black,fg=white}
\setbeamercolor{structure}{fg=black} % itemize, enumerate, etc
\setbeamercolor{section in toc}{fg=black} % TOC sections

\beamertemplatenavigationsymbolsempty
\usefonttheme[onlymath]{serif}

\usepackage{graphicx}

\newcommand{\nn}{\nonumber\\}
\renewcommand{\P}{\mathcal{P}}
\renewcommand{\L}{\mathcal{L}}
\newcommand{\M}[1]{\mathcal{M}_{\rm max}^{#1}}
\newcommand{\I}[1]{{I_{#1}}}
\newcommand{\J}[1]{{J_{#1}}}

\title{Multipole Mermin-Wagner Constraints}
\subtitle{}
\author{Charles Stahl}
\institute{University of Colorado \\ Department of Physics}
\date{February 18, 2022}

\AtBeginSection
{
  \begin{frame}
    \frametitle{Outline}
    \tableofcontents[currentsection]
  \end{frame}
}

\begin{document}

\frame{\titlepage}

%\begin{frame}{Outline}
%Ordinary Mermin-Wagner theorem\\\
%
%Multipole symmetries\\\
%
%Multipole field theories and Mermin-Wagner\\\
%
%\hrule
%\vspace{6pt}
%
%Mermin-Wagner for partial breaking of multipole symmetries
%\end{frame}

\section{Ordinary Mermin-Wagner theorem}

\begin{frame}{Mermin-Wagner theorem}
Consider a system with an internal $U(1)$ symmetry\\\

Assume the symmetry is broken $\rightarrow$ Goldstone boson
\begin{align*}
\L = (\partial_t \phi)^2 + (\partial_i \phi)^2
\end{align*}
Correlation functions go as
\begin{align*}
\langle \phi(x,t) \phi(0,0) \rangle &= \int \frac{d\omega dk \, k^{d-1}}{\omega^2+ k^2}\\ 
&= \int \frac{dk}{k} k^{d-1},
\end{align*}
which diverges (in the IR) when $d\le 1$
\end{frame}

\begin{frame}{Interpretation}
When the correlation function does not diverge ($d>1$), there can be a phase distinguished by symmetry properties\\\

At the critical dimension ($d=1$), there can still be a phase distinguished by a non-vanishing compressibility

\setlength{\leftskip}{1cm}

Luttinger liquid

Quasi-long-range order
\end{frame}

\section{Multipole symmetries}

\begin{frame}{Multipole symmetries}
Recall that an ordinary symmetry acts as $\phi(x) \rightarrow \phi(x) + c$\\\

Gauge ``symmetries" are not physical symmetries\\\

Generalize to $\phi(x) \rightarrow \phi(x) + P(x)$

\setlength{\leftskip}{1cm}

$P(x) = a_ix^i + b$, $c(x^2+1)$, etc.\\\

\setlength{\leftskip}{0cm}

Gauging these symmetries can lead to fractons [Pretko 2018]
\end{frame}

\begin{frame}{Multipole groups}
Choose an internal group $G_{\rm int}$ and a set of polynomials\\\

The internal group has to be Abelian [Glorioso et. al. 2020], stick with $G_{\rm int}=U(1)$\\\

There is a generator corresponding to each polynomial [Gromov 2019]

\setlength{\leftskip}{1cm}

$\mathcal{P}_0, \P_1^i, \P_2^{ij}$, etc.\\\

$\exp(c \P_0) \phi(x) = \phi(x) +c$\\\

$\exp (a_{ij} \P_2^{ij}) \phi(x) = \phi(x) + a_{ij} x^i x^j$\\

\setlength{\leftskip}{0cm}

\small (restriction to monomials makes indices simpler)
\end{frame}

\begin{frame}{Multipole groups}
Generators might not commute with space symmetries [Gromov 2019]
\begin{align*}
[T_j, \P_a^{i_1\dots i_a}] &= \delta^{i_m}_j \P_{a-1} ^{i_1\dots i_{m-1}i_{m+1}\dots i_a}\\
[R_{jk}, \P^{i_1\dots i_a}_a] &= \delta_{[j}^{(i_1} \P^{k]i_2)\dots i_a}_a
\end{align*}

For example, $[T_1,\P_1^1] = \P_0$ and $[R_{12}, \P_1^1] = \P_1^2$\\\

Include all resulting polynomials, or exclude some translations or rotations
\end{frame}

\begin{frame}{Maximal multipole group}
$\M{a}$ contains all polynomials of degree $<a$

Note: this is NOT the definition in the literature, but it makes formulas nicer\\\

Any multipole group is therefore a subgroup of $\M{a}$ for some $a$\\\

Also contains all translations and rotations\\\

$\M{1}$ is ordinary $U(1)$, $\M{0}$ is the trivial group
\end{frame}

\section{Multipole field theories and Mermin-Wagner}

\begin{frame}{Multipole-invariant field theories}
Example: $\mathcal{L} = (\partial_t \phi)^2 + g (\partial_i \partial_j \phi)^2$ is invariant under $\M{2}$\\\

For a generic multipole group, it is not possible to find such a theory [Gromov 2019]\\\

Interesting examples do exist, and can lead to fracton phases after gauging [Pretko 2018; Gromov 2019]\\\

A special example leads to Haah's code [Bulmash, Barkeshli 2018]
\end{frame}

\begin{frame}{Multipole-invariant field theories}
Stick with rotationally invariant theory
\begin{align*}
\L &= (\partial_t \phi)^2 + g(\partial_{I_a} \phi)^2,
\end{align*}
where $I_a = i_1 \dots i_a$ is a multi-index and 
\begin{align*}
\partial_{I_a} = \partial_{i_1}\dots \partial_{i_a}
\end{align*}
This is invariant under $\M{a}$\\\

Interesting things to say about total derivatives and rotational invariance...
\end{frame}

\begin{frame}{Mermin-Wagner for multipole theories}
$\L = (\partial_t \phi)^2 + g (\partial_{I_a} \phi)^2$\\\

Assume the $\phi$ field exists. Is long-range order possible?
\begin{align*}
\left\langle \phi(x,t) \phi(0) \right\rangle &\sim \int \frac{d\omega dk\, k^{d-1}}{\omega^2 + k^{2a}}\\
&= \int \frac{dk}{k} k^{d-a}
\end{align*}

IR divergence for $d\le a$ [Griffin 2015]\\\

Critical dimension at $T=0$ is $d_{\rm c} = a$
\end{frame}

\begin{frame}{Mermin-Wagner for multipole theories}
What other possibilities are there?\\\

Non-maximal groups, eg. Haah group\\\

Partial breaking from $\M{a} \rightarrow \M{c}$\\\

Combine both?
\end{frame}

\section{Mermin-Wagner for partial breaking of multipole symmetries}

\begin{frame}{Mermin-Wagner for partial breaking, I}
Naive method to break from $\M{a}$ to $\M{c}$ with $c<a$\\\

Define field $\phi_{i_1\dots i_c} = \phi_\I{c}$

\setlength{\leftskip}{1cm}

$\exp(\beta_{I_b} \P_b^{I_b})\phi_{J_{c}}(x) = \phi_\J{c} + \beta_{\J{c}\I{b-c}} x^\I{b-c}$

Invariant under polynomials of degree $<c$

$\phi_\I{c}$ develops LRO: $\M{a} \rightarrow \M{c}$\\\

\setlength{\leftskip}{0cm}

Example: $\phi_i$ would break the dipole group $\M{2}$ to the monopole group $\M{1}$\\\

[CNS, Lake, Nandkishore]
\end{frame}

\begin{frame}{Mermin-Wagner for partial breaking, I}
Naive result: can $\phi_\I{c}$ develop LRO?\\\

Consider $\L = (\partial_t \phi_\I{c})^2 + (\partial_\J{a-c} \phi_\I{c})^2$. Then
\begin{align*}
\left \langle \phi_\I{c} (x,t) \phi_\I{c} (0) \right \rangle &\sim \int \frac{d\omega dk\, k^{d-1}}{\omega^2 + k^{2a-2c}}\\
&= \int \frac{dk}{k} k^{d-a+c}
\end{align*}
It looks like the critical dimension should be $d_{\rm c} = a-c$\dots
\end{frame}

\begin{frame}{Mermin-Wagner for partial breaking, II}
The operator $\partial_\I{c-b} \phi_\J{b}$ is breaks $\M{a} \rightarrow \M{c}$\\\

Let's look at the theory $\L = (\partial_t \phi_\J{b})^2 + (\partial_\I{a-b} \phi_\J{b})^2$

\setlength{\leftskip}{1cm}

First $b$ compressibilities are all 0 $\rightarrow$ specifies theory

$\kappa_e \equiv \frac{d \langle n_e^\I{e} \rangle}{d \mu_e^\I{e}}=0$ for $e\le b$\\\

\setlength{\leftskip}{0cm}

Now the correlation function becomes
\begin{align*}
\langle \partial_\I{c-b} \phi_\J{b} (x,t) \partial_\I{c-b} \phi_\J{b} (0,0) \rangle &= \int \frac{d\omega dk\,k^{d-1}\, k^{2(c-b)}}{ \omega^2 + k^{2(a-b)}}\\
&= \int \frac{dk}{k}k^{d-a-b+2c}
\end{align*}
and $\partial_\I{c-b} \phi_\J{b}$ can only order if $d \ge a+b-2c$
\end{frame}

\begin{frame}{Interpreting the critical dimension}
$a$ and $b$ define the theory, $c$ defines the preserved subgroup\\\

At fixed $a,b,c$

\setlength{\leftskip}{1cm}

In this theory, $\M{c+1}$ can only be broken if $d > a+b-2c$

Doesn't tell you what the preserved subgroup actually is\\\

\setlength{\leftskip}{0cm}

At fixed $a,b,d$: What is the preserved subgroup?

\setlength{\leftskip}{1cm}

$\M{c}$ is broken for $c > (a+b-d)/2$

Preserved subgroup is $\M{\lfloor (a+b-d)/2 \rfloor} $\\\

\setlength{\leftskip}{0cm}

At fixed $a,c$: When is this SSB possible?

\setlength{\leftskip}{1cm}

Critical dimension is $d = \min_b a+b-2c = a-2c$

Symmetry breaking is possible at $d=1$!
\end{frame}

\begin{frame}{Example: dipolar Bose-Hubbard model}
Bose-Hubbard model with dipole constraint imposed
\begin{align*}
H = -t \sum_{i,\mu,\nu} b^\dagger_i b_{i+\mu} b^\dagger_{i+\mu+\nu} b_{i+\nu} -\mu \sum_i n_i + \frac{U}{2} n_i(n_i-1)
\end{align*}
\begin{figure}
	\centering
	\includegraphics[scale=.8]{bhm}
	\includegraphics[scale=.8]{dbhm}
\end{figure}
[Lake, Hermele, Senthil]
\end{frame}

\begin{frame}[t]{Example: dipolar Bose-Hubbard model}
\begin{figure}
	\centering
	\includegraphics[scale=.8]{bhm}
	\includegraphics[scale=.8]{dbhm}
\end{figure}
Dipole Condensate: \hspace{.2cm} $(\partial_t \phi_i)^2 + (\partial_j \phi_i)^2 \quad (b=1)$\\\

Bose-Einstein Insulator: $(\partial_t \phi)^2 + (\partial_j \phi)^2 \quad (b=0)$
\end{frame}

\begin{frame}[t]{Example: dipolar Bose-Hubbard model}
\begin{figure}
	\centering
	\includegraphics[scale=.8]{bhm}
	\includegraphics[scale=.8]{dbhm}
\end{figure}
Normal Bose-Hubard model

\setlength{\leftskip}{1cm}

$d\ge 2$: SF has SSB $\M{1} \rightarrow \M{0}$

$d=1$: SF has no SSB, instead has QLRO
\end{frame}

\begin{frame}[t]{Example: dipolar Bose-Hubbard model}
\begin{figure}
	\centering
	\includegraphics[scale=.8]{bhm}
	\includegraphics[scale=.8]{dbhm}
\end{figure}
Dipolar Bose-Hubard model

\setlength{\leftskip}{1cm}

$d\ge 3$: BEI:$\M{2} \rightarrow \M{0}$, DC: $\M{2} \rightarrow \M{1}$

$d=2$: BEI:$\M{2} \rightarrow \M{1}$, DC: $\M{2} \rightarrow \M{1}$

$d=1$: BEI:$\M{2} \rightarrow \M{1}$, DC: $\M{2} \rightarrow \M{2}$
\end{frame}

\end{document}

### 2022-03-08 multipoleSSB/manuscript.tex
starting referee response

@@
-\documentclass[pra,aps,twocolumn, amsfonts,amsmath,amssymb,nofootinbib,superscriptaddress]{revtex4-2}
+\documentclass[prb,aps,twocolumn, amsfonts,amsmath,amssymb,nofootinbib,superscriptaddress]{revtex4-2}
 
@@
 
+The differences from the ordinary Mermin-Wagner theorem all come from the soft dispersions of the relevant Goldstone modes.
+Although soft Goldstone modes exist in other models, they generically only emerge at fine-tuned points.
+As we will see, the multipole symmetries make the soft modes natural, in the sense that the theories do not have to be fine-tuned.
+Rather, the soft modes must exist in the presence of various multipolar symmetries.
+
 This paper is organized as follows.
@@
 This group also includes all translations and rotations.
+We will use $\mmax{-1}$ to denote the rotations and translations without any internal symmetry.
 
@@
 \end{align}
-where $\mu^{\beta}$ are a basis for the rank-2 traceless symmetric $d\times d$ matrices.
+where $\mu^{\beta}$ are a basis for the traceless symmetric $d\times d$ matrices.
 Let us call this group $\mathcal{M}^2_{\text{sym}}$.
@@
 \begin{align}
-K_1[\phi(x)] = g_1 (\partial^2\phi)^2 + g_2 \sum_{ij} (\partial_i \partial_j \phi)^2.
+K_1[\phi(x)] = g_1 \big(\sum_{i} \partial_i \partial_i \phi\big)^2 + g_2 \sum_{i \ne j} (\partial_i \partial_j \phi)^2.
 \end{align}
-For the special cases of $g_2=0$ or $g_2=-g_1$, the symmetry group expands to $\mathcal{M}^2_\text{sym}$ or $\mathcal{M}^2_\text{tr}$, respectively.
+For the special cases of $g_2=0$ or $g_2=0$, the symmetry group expands to $\mathcal{M}^2_\text{sym}$ or $\mathcal{M}^2_\text{tr}$, respectively.
 However, for generic $g_1,g_2$ the symmetry group is $\mathcal{M}^1_\max$.
 Similar statements can be made for larger $a$.
+
+Although we have been discussing field theories in the continuum, we will often care about theories that emerge from the low-energy degrees of freedom of a lattice model.
+In ordinary theories with monopole and lattice rotation symmetries, the lowest-order derivative term that can emerge from the lattice theory is $(\partial_i \phi)^2$, which is symmetric under the larger group of continuous rotations.
+Any terms that are not invariant under the continuous rotation group are irrelevant, so the continuous symmetry emerges.
+For multipole symmetries, leading-order terms may contain lattice anisotropies, so that the continuous rotation group does not emerge.
+For this reason we will only require that $\mmax{a}$ contain discrete rotation symmetries.
 
@@
 Here, we will discuss both thermal and $T=0$ systems.
+We will start with a heuristic argument in terms of domain wall energies before discussing a more careful diagnosis of spontaneous symmetry breaking in terms of correlation functions.
 
@@
 \end{align} 
-Now consider nucleating domains of linear size $L$ in which the expectation value of the order parameter --- namely $\langle e^{i\phi}\rangle$ --- varies by an amount of order 1.
-Since the order parameter is continuous there will be a thick domain ``wall" over which the order parameter changes gradually from its value in the background phase to its value in the domain.
-The thickness will be of order $L$.
-Equation~\ref{eqn:adisp} tells us the  energy cost of the field changing its value will be minimized if it varies as a polynomial with degree $a+1$.
-In that case the energy density will scale as $K_k \sim k^{2(a+1)} \sim L^{-2(a+1)}$.
-Integrating that energy density over a region of size $L$ gives a total energy cost to nucleating domains of $L^{d-2(a+1)}$.
+Symmetry breaking field configurations will be of the form $\phi(x) = c + c_ix^i + \dots+ c_{I_a}x^{I_a}$, so there is no kinetic energy.
+
+Now consider nucleating domains of linear size $L'$ in which the field takes on a new configuration $\phi(x) = c' + c'_ix^i + \dots + c'_{I_a}x^{I_a}$.
+In the thick domain ``wall" surrounding the domain, the coefficients will vary from their background unprimed values to the primed values in the domain.
+The equations of motion from~\eqref{eqn:adisp} say that $\phi$ should vary as a degree-$(2a+1)$ polynomial in this region.
+
+Let us align the domain wall in the $x^1$ direction with thickness $L$, so that $\phi$ takes its unprimed values at $x^1=0$ and its primed values at $x^1=L$.
+We can build a degree-$(2a+1)$ polynomial by including factors of $(x^1)^{2a+1}/L^{2a+1}$.
+Terms with fewer powers of $L$ in the denominator will not be seen by the kinetic energy.
+For large $L$, terms with more powers of $L$ in the denominator will be subleading.
+Thus the energy density in the domain wall scales as $K\sim L^{-2(a+1)}$.
+Integrating over a spherical shell gives a total energy cost of
+\begin{align}
+E &\sim \left( (L'+L)^d -(L')^d \right) L^{-2(a+1)} \nonumber\\
+&\sim L^{d-2(a+1)},
+\end{align}
+where the last line comes from setting $L\sim L'$, which minimizes the energy cost.
 
@@
 
-We can also reproduce this argument by considering correlation functions.
+We can reproduce this argument by considering correlation functions.
 As a warm-up, consider the standard case where a monopole U(1) symmetry is spontaneously broken, leading to a Goldstone boson $\phi(x)$.
@@
 
-Consider a situation in which a dipolar symmetry $\mcm^1_{\rm max}$ is spontaneously broken to its monopolar subgroup $\mcm^0$, and specialize to the case of $d=3$ and $T>0$.
+Consider a situation in which a dipolar symmetry $\mcm^1_{\rm max}$ is spontaneously broken to its monopolar subgroup $\mcm^0_{\rm max}$, and specialize to the case of $d=3$ and $T>0$.
 In the spirit of Goldstone's theorem, the most naive thing to do when analyzing the symmetry broken phase would be to write down an action involving a set of $d=3$ Goldstone fields $\theta_j$, with $\t_j$ transforming linearly under the dipolar part of the symmetry group, but remaining invariant under the unbroken monopole subgroup, so that $\theta_j \mapsto \theta_j + \beta_j$ under a transformation parameterized by $P(x) = \alpha + \beta_j x^j$.
@@
 As we are working in $d=3$ and at $T>0$, the theory \eqref{vector} indeed has SSB for the dipole symmetry, with $e^{i\t_j}$ developing long-range order.
+
+Of course, such a Goldstone field would break continuous rotation symmetries, but recall that we are only requiring that each maximal multipole group contain lattice rotation symmetries.
+We can then say that the theory~\eqref{vector} spontaneously breaks $\mmax{1}$ to $\mmax{0}$, in $d=3$ and at $T>0$.
 
@@
 A general formal framework for understanding the counting of Goldstone modes in the present context can be developed by following the analysis of \cite{watanabe2013redundancies}, but here we will simply content ourselves with a heuristic physical argument.
-For simplicity, we will focus on systems in which the total symmetry group $\mcm_{\rm max}^a$ is spontaneously broken down to some nontrivial subgroup.
+For simplicity, we will focus on systems in which the total symmetry group $\mcm_{\rm max}^a$ is spontaneously broken down to some nontrivial subgroup $\mcm_{\rm max}^b$.
+As mentioned previously, $\mmax{b}$ might only contain discrete rotation symmetries, so we will only concern ourselves with the spontaneous breaking of the polynomial generators.
+We will indicate that a generator corresponding to a polynomial of degree $c$ is spontaneously broken by saying that $\mmax{c}$ is broken.
 
-Let us define the compressibilities $\kappa_c$ by 
+Determining the counting of the Goldstone modes requires introducing the compressibilities $\kappa_c$, defined by 
 \be \label{kappadef} \kappa_c^{I_c} = \frac{d\langle n_c^{I_c}\rangle}{d\mu_c^{I_c}},\ee 
@@
 
-Let $b\leq a$ denote the smallest degree of multipole charge with nonzero compressibility, i.e.
-consider a situation in which $\kappa_c>0$ for all $b\leq c\leq a$, while $\kappa_c=0$ for all $c<b$.
-Then the minimal IR Lagrangian for the symmetry broken phase looks schematically like
-\be \label{generall} \mathcal{L} = \kappa_b (\partial_\tau \phi_{I_b})^2 + g_{J_{a-b+1}}(\partial_{J_{a-b+1}}\phi_{I_b})^2,\ee 
-where 
-$\p_{J_{a-b+1}} \equiv \p_{j_1} \cdots \p_{j_{a-b+1}}$, and where $\phi_{I_b}$ is a field that shifts by a polynomial of degree $a-b$ under the action of $\mmax a$.
-The compressibility appears as the coefficient of the time derivative term since $\p_\tau \phi_{I_b}$ is the momentum conjugate to $\phi_{I_b}$ in the quantum theory, and hence correlation functions of fluctuations in $n_b$ are determined by those of $\p_\tau \phi_{I_b}$.
+A necessary condition for $\mmax c$ to be spontaneously broken is that $\kappa_c \neq 0$.
+Indeed, if $\mmax c$ is spontaneously broken then $n_c$ creates Goldstones when acting on the ground state, thereby ensuring that the density-density correlator appearing in the calculation of \eqref{kappadef} must be nonzero at long wavelengths.
+Since we are assuming that $\mmax c$ is spontaneously broken for all $c>b$, we must accordingly have $\kappa_{c>b} \neq 0$.
 
-A necessary condition for $\mmax b$ to be spontaneously broken is that $\kappa_b \neq 0$.
-Indeed, if $\mmax b$ is spontaneously broken then $n_b$ creates Goldstones when acting on the ground state, thereby ensuring that the density-density correlator appearing in the calculation of \eqref{kappadef} must be nonzero at long wavelengths.
+However, $\kappa_c \neq 0$ does not {\it necessarily} imply that $\mmax c$ is spontaneously broken: $\kappa_c\neq 0$ means that the system possesses gapless modes created by $n_c$, but it is possible for fluctuations to be strong enough so that these modes cannot be associated with the Goldstone modes of a spontaneously broken $\mmax c$ (this is of course what happens in the QLRO phase of the 1+1D XY model).
+Therefore, we are allowed to have $\kappa^c \neq 0$ even if $c \leq b$, without contradicting our assumption about the symmetry breaking pattern.
 
-However, a nonzero $\kappa_b$ does not {\it necessarily} imply that $\mmax b$ is spontaneously broken.
-$\kappa_b\neq 0$ means that the system possesses gapless modes created by $n_b$, but it is possible for fluctuations to be strong enough so that these modes cannot be associated with the Goldstone modes of a spontaneously broken $\mmax b$.
-Thus the theory \eqref{generall} does not necessarily describe a phase where $\mmax b$ is spontaneously broken, and the exact pattern of symmetry breaking depends on the spatial dimension and whether or not $T$ is nonzero.
-For example, we can again consider the theory in \eqref{singlescalar} at $d=3, T>0$: even though the charge compressibility $\kappa_0$ (the coefficient of $(\p_\tau\phi)^2$ in \eqref{singlescalar}) is nonzero, $\phi$ fluctuates strongly enough that the monopole symmetry is not spontaneously broken.
+Let then $s\leq a$ denote the smallest degree of multipole charge with nonzero compressibility, i.e.
+consider a situation in which $\kappa_c>0$ for all $s\leq c\leq a$, while $\kappa_c=0$ for all $c<s$.
+Given $s$, the minimal IR Lagrangian looks schematically like 
+\be \label{generall} \mathcal{L} = \kappa_s (\partial_\tau \phi_{I_s})^2 + g_{J_{a+1}}(\partial_{J_{a-s+1}}\phi_{I_s})^2,\ee where 
+$\p_{J_{a-s+1}} \equiv \p_{j_1} \cdots \p_{j_{a-s+1}}$, and where $\phi_{I_s}$ is a field that shifts by a polynomial of degree $a-s$ under the action of $\mmax a$ (the compressibility appears as the coefficient of the time derivative term since $\p_\tau \phi_{I_s}$ is the momentum conjugate to $\phi_{I_s}$ in the quantum theory, and hence correlation functions of fluctuations in $n_s$ are determined by those of $\p_\tau \phi_{I_s}$).
+If $s=b+1$, then $\phi_{I_s}$ may develop LRO while leaving $\mmax{b}$ invariant.
+
+Note that the number of modes in the IR theory is controlled by $s$, and {\it not} by $b$.
+In order for $s< b+1$ to be consistent with our assumed pattern of symmetry breaking, it must be the case that the fluctuations of the fields in \eqref{generall} are strong enough so as to prevent $\mmax {s}$ from spontanoeusly breaking.
+The novelty provided by multipolar symmetries is that they offer an easy way of allowing for $s$ to be smaller than $b+1$, and thus provide an easy way of realizing multiple IR theories that have the same symmetry breaking pattern, but contain different numbers of modes.
+This is illustrated by the examples of the theories \eqref{vector} ($s=1$) and \eqref{singlescalar} ($s=0$), which have the same symmetry breaking pattern ($b=0$) in $d=3$ and at $T>0$.
 
@@
 
-In the previous subsection, we saw that the number of Goldstone modes in the symmetry-breaking theory is determined by the smallest degree of multipole charge $b$ such that $\kappa_b\neq0$.
-We now turn to determining the symmetry breaking pattern that occurs for a given choice of $b$, spatial dimension $d,$ and temperature $T$, thereby allowing us to formulate a generalized version of the Mermin-Wagner theorem for the case of partial multipole breaking.
+In the previous subsection, we saw that the number of Goldstone modes in the symmetry-breaking theory is determined by the smallest degree of multipole charge $s$ such that $\kappa_s\neq0$.
+We now turn to determining the symmetry breaking pattern that occurs for a given choice of $s$, spatial dimension $d,$ and temperature $T$, thereby allowing us to formulate a generalized version of the Mermin-Wagner theorem for the case of partial multipole breaking.
 
-Since $\kappa_c=0$ if $c<b$, none of the subgroups $\mmax {c<b}$ can be spontaneously broken.
-Suppose then that the symmetry is spontaneously broken down to $\mmax {c}$, where $c\geq b$.
-To find the allowed values of $c$ consistent with this assumption, we need to calculate correlation functions of operators of the form $\exp(i\p_{J_{c-b}}\phi_{I_b})$, these being the operators transforming nontrivially under $\mmax c$ with the longest-range correlation functions.
+Since $\kappa_c=0$ if $c<s$, none of the subgroups $\mmax {c<s}$ can be spontaneously broken.
+Suppose then that the subgroup $\mmax{c}$ is spontaneously broken, where $c\geq s$.
+To find the allowed values of $c$ consistent with this assumption, we need to calculate correlation functions of operators of the form $\exp(i\p_{J_{c-s}}\phi_{I_s})$, these being the operators transforming nontrivially under $\mmax c$ with the longest-range correlation functions.
 We have
-\be\ba \langle \p_{J_{c-b}}\phi_{I_b}(x) \p_{J_{c-b}}\phi_{I_b}(0)\rangle & \sim \int \frac{d\omega \, dk\, k^{d-1+2(c-b)}}{\omega^2 + k^{2(a-b+1)}} \\ 
-& \sim \int dk\, k^{d-1+2(c-b) - \varsigma_T(a-b+1)}, \ea\ee 
+\be\ba \langle \p_{J_{c-s}}\phi_{I_s}(x) \p_{J_{c-s}}\phi_{I_s}(0)\rangle & \sim \int \frac{d\omega \, dk\, k^{d-1+2(c-s)}}{\omega^2 + k^{2(a-s+1)}} \\ 
+& \sim \int dk\, k^{d-1+2(c-s) - \varsigma_T(a-s+1)}, \ea\ee 
 where we have defined the function $\varsigma_T$ such that $\varsigma_T = 1$ if $T=0$ and $\varsigma_T = 2$ if $T>0$.
 Consequently, we find that $\mmax c$ can be spontaneously broken only in dimensions $d$ such that
-\be \label{pbreak_mw} d>2(b-c) + \varsigma_T(a-b+1).\ee 
-For a given choice of symmetry breaking pattern (captured by $a$ and $c$) and a given choice of which compressibilities are nonzero (captured by $b$), the quantity on the RHS determines the lower critical dimension for symmetry breaking, thereby generalizing the Mermin-Wagner theorem to the partial breaking case.
-As a sanity check on \eqref{pbreak_mw}, note that when we set $c=b=0$ (full multipole breaking) we recover the lower critical dimension of $\varsigma_T(a+1)$ obtained in the previous section.
-Also note that as expected, the lower critical dimensions in the quantum $(\varsigma_T = 1)$ and classical $(\varsigma_T = 2)$ cases are related by $d_{\rm cl} = d_{\rm q} + z$, where $z=a-b+1$ is the dynamical exponent of the Goldstone theory \eqref{generall}.
+\be d>2(s-c) + \varsigma_T(a-s+1).\ee 
+The preserved subgroup is then the largest multipole group $\mmax{b}$ that is not spontaneously broken.
 
-Finally, as an interesting example, consider the case where the dipolar group $(a=1)$ is broken down to the monopole group $(b=0,c=1)$.
-At $T=0$, we see from \eqref{pbreak_mw} that this is possible provided that $d>0$.
+For a given choice of symmetry breaking pattern (captured by $a$ and $b$) and a given choice of which compressibilities are nonzero (captured by $s$), the critical dimension for symmetry breaking is
+\begin{align}
+d_\text{lc} = 2(s - b-1) + \varsigma_T (a-s+1) \label{pbreak_mw} 
+\end{align}
+generalizing the Mermin-Wagner theorem to the partial breaking case. 
+As expected, the lower critical dimensions in the quantum $(\varsigma_T = 1)$ and classical $(\varsigma_T = 2)$ cases are related by $d_{\rm cl} = d_{\rm q} + z$, where $z=a-s+1$ is the dynamical exponent of the Goldstone theory \eqref{generall}.
+
+If we only care about the pattern of symmetry breaking, $\mmax{a}\rightarrow \mmax{b}$, the critical dimension can be found by setting $s=0$, so that
+\begin{align}
+d_\text{lc} = \varsigma_T (a+1) -2 (b+1). \label{pbreak_mw_nos}
+\end{align}
+As a sanity check on \eqref{pbreak_mw_nos}, note that for full multipole breaking, $b=-1$ (recall that $\mmax{-1}$ contains only translations and rotations), we recover the lower critical dimension of $\varsigma_T(a+1)$ obtained in the previous section.
+In any higher dimension, such a symmetry breaking pattern will be possible.
+However, $s$ must always be chosen so as to give the correct preserved subgroup.
+As emphasized earlier, the number of Goldstone modes depends on $s$, not $b$.
+
+As an interesting example, consider the case where the dipolar group $(a=1)$ is broken down to the monopole group $(b=0)$.
+At $T=0$, we see from \eqref{pbreak_mw_nos} that this is possible provided that $d>0$.
 Thus the partial breaking of dipole symmetry provides us with an example where a continuous symmetry can be spontaneously broken in $d=1$ at zero temperature.
@@
 where $\langle h_{-k} h_{k} \rangle e^{ik\cdot x}$ does not affect the divergence at small $k$, assuming the disorder is short range correlated.
-We can compare to Eqn.~\ref{eqn:correl} to see that in the presence of disorder that couples linearly to an order parameter fully breaking multipole symmetry, the critical dimension for a disordered system is twice the critical dimension for having direct full multipole symmetry breaking in a clean classical system.
+We can compare to~\eqref{eqn:correl} to see that in the presence of disorder that couples linearly to an order parameter fully breaking multipole symmetry, the critical dimension for a disordered system is twice the critical dimension for having direct full multipole symmetry breaking in a clean classical system.
 
@@
 We now present a transparent quantum lattice example to illustrate the spontaneous breaking of multipolar symmetry.
-To this end, we take seriously the idea that ``dipole charges are monopole charges of a dipole symmetry," and include explicit secondary degrees of freedom that couple to dipole moments of the original degrees of freedom.
-This allows us to access phases corresponding to partial or complete symmetry breaking.
+We will not derive further results from this model.
+Rather, we apply our previous results to this model to understand how various phases are available in different dimensions.
+Reference~\cite{dbhm} gives detailed descriptions of the various phases and their transitions.
 
@@
 so there is a robust insulating phase at weak hopping $t_b, t_d$~\cite{Fisheretal}.
-A detailed discussion of the phase diagram and phenomenology of this model will be given in~\cite{dbhm}; in the following we will simply discuss the symmetry breaking patterns realized in various different limits.
+A detailed discussion of the phase diagram and phenomenology of this model is given in~\cite{dbhm}; in the following we will simply discuss the symmetry breaking patterns realized in various different limits.
 

### 2022-03-09 multipoleSSB/manuscript.tex
added referee response

@@
 \end{align}
-For the special cases of $g_2=0$ or $g_2=0$, the symmetry group expands to $\mathcal{M}^2_\text{sym}$ or $\mathcal{M}^2_\text{tr}$, respectively.
+For the special cases of $g_2=0$ or $g_1=0$, the symmetry group expands to $\mathcal{M}^2_\text{sym}$ or $\mathcal{M}^2_\text{tr}$, respectively.
 However, for generic $g_1,g_2$ the symmetry group is $\mathcal{M}^1_\max$.

### 2022-03-16 multipoleSSB/manuscript.tex
ack for Leo

@@
 
-The differences from the ordinary Mermin-Wagner theorem all come from the soft dispersions of the relevant Goldstone modes.
+Many of the differences from the ordinary Mermin-Wagner theorem come from the soft dispersions of the relevant Goldstone modes.
 Although soft Goldstone modes exist in other models, they generically only emerge at fine-tuned points.
 As we will see, the multipole symmetries make the soft modes natural, in the sense that the theories do not have to be fine-tuned.
-Rather, the soft modes must exist in the presence of various multipolar symmetries.
+In addition, in Sec.~\ref{sub:partial} we will see that there are exotic partial symmetry-breaking options that are not available in the presence of non-multipolar soft modes.
 
@@
 This already appears in Ref.~\cite{Griffin2015}, so we are simply reviewing it here in preparation for the more generic cases to follow.
-Here, we will discuss both thermal and $T=0$ systems.
-We will start with a heuristic argument in terms of domain wall energies before discussing a more careful diagnosis of spontaneous symmetry breaking in terms of correlation functions.
+Here, we will discuss both thermal and $T=0$ systems, and diagnose spontaneous symmetry breaking using the divergence of correlation functions.
 
-Consider first a clean system at $T>0$ with a spontaneously broken maximal multipole symmetry of degree $a$.
-Then the kinetic term is proportional to 
-\begin{align}
-K[\phi(x)] = (\partial_{I_{a+1}} \phi)^2. \label{eqn:adisp}
-\end{align} 
-Symmetry breaking field configurations will be of the form $\phi(x) = c + c_ix^i + \dots+ c_{I_a}x^{I_a}$, so there is no kinetic energy.
-
-Now consider nucleating domains of linear size $L'$ in which the field takes on a new configuration $\phi(x) = c' + c'_ix^i + \dots + c'_{I_a}x^{I_a}$.
-In the thick domain ``wall" surrounding the domain, the coefficients will vary from their background unprimed values to the primed values in the domain.
-The equations of motion from~\eqref{eqn:adisp} say that $\phi$ should vary as a degree-$(2a+1)$ polynomial in this region.
-
-Let us align the domain wall in the $x^1$ direction with thickness $L$, so that $\phi$ takes its unprimed values at $x^1=0$ and its primed values at $x^1=L$.
-We can build a degree-$(2a+1)$ polynomial by including factors of $(x^1)^{2a+1}/L^{2a+1}$.
-Terms with fewer powers of $L$ in the denominator will not be seen by the kinetic energy.
-For large $L$, terms with more powers of $L$ in the denominator will be subleading.
-Thus the energy density in the domain wall scales as $K\sim L^{-2(a+1)}$.
-Integrating over a spherical shell gives a total energy cost of
-\begin{align}
-E &\sim \left( (L'+L)^d -(L')^d \right) L^{-2(a+1)} \nonumber\\
-&\sim L^{d-2(a+1)},
-\end{align}
-where the last line comes from setting $L\sim L'$, which minimizes the energy cost.
-
-In clean systems there is no energy gain from domain nucleation.
-When $d\le2(a+1)$, the energy cost is bounded, so the entropy gain favors domain creation.
-Thus, ordered phases are unstable and SSB cannot occur for $d\le2(a+1)$.
-
-We can reproduce this argument by considering correlation functions.
-As a warm-up, consider the standard case where a monopole U(1) symmetry is spontaneously broken, leading to a Goldstone boson $\phi(x)$.
-The correlation function
+As a warm-up, consider the standard case where a monopole U(1) symmetry is spontaneously broken, leading to a Goldstone boson $\phi(x)$. The correlation function
 \begin{align}
@@
 
-If we look at the zero temperature setting, the energy scaling no longer tells us where the critical dimension is.
-This is because even when there is no energy cost to forming domains, there is no entropy gain.
-Instead, quantum fluctuations must be the motivation for domain nucleation.
-To find the quantum critical dimension, we calculate the correlation function by way of the (imaginary-time) IR Lagrangian
+In the zero temperature setting, we calculate the correlation function by way of the (imaginary-time) IR Lagrangian 
 \begin{align}
@@
 If $d_1=2$ the symmetry can be broken for any $d_2>0$, and if $d_1=1$ then $d_2=2$  is critical. Of course, if $d_1=0$ we have an isotropic quartic dispersion and $d=d_2=4$ is critical.
-
-We can recover this result in  the energy-scaling argument.
-Since the system is anisotropic, the domains will have different sizes in different directions.
-Consider forming a domain of linear size $L_p$ in the  direction with quadratic dispersion and $L_k$ in the quartic direction.
-When the gradient of the order parameter is in the quadratic direction, the order parameter has to change over a length $L_p$ so the energy density is $L_p^{-2}$.
-Similarly, when the order parameter is changing in the quartic direction its energy density is $L_k^{-4}$.
-
-Both of these energy densities will be integrated over domain ``walls" with volume $L_k^{d_2} L_p^{d_1}$.
-As a result, the total energy cost is
-\begin{align}
-E &\sim L_k^{d_2} L_p^{d_1-2} + L_k^{d_2-4} L_p^{d_1},
-\end{align}
-so that $L_p \sim L_k^2$ to make the terms match, and $E\sim L_p^{d_2/2+d_1-2}$.
-Recall that when $E$ does not increase for larger domains, the entropy gain will cause them to nucleate, destroying the ordered phase.
-We see that the critical dimension $d= d_2 + d_1$ can be 2, 3, or 4, depending on how many directions have quadratic or quartic dispersions.
 
@@
 
-\subsection{Goldstone counting} \label{sub:subgroup}
+\subsection{Counting massless modes} \label{sub:subgroup}
 
@@
 \be \label{generall} \mathcal{L} = \kappa_s (\partial_\tau \phi_{I_s})^2 + g_{J_{a+1}}(\partial_{J_{a-s+1}}\phi_{I_s})^2,\ee where 
-$\p_{J_{a-s+1}} \equiv \p_{j_1} \cdots \p_{j_{a-s+1}}$, and where $\phi_{I_s}$ is a field that shifts by a polynomial of degree $a-s$ under the action of $\mmax a$ (the compressibility appears as the coefficient of the time derivative term since $\p_\tau \phi_{I_s}$ is the momentum conjugate to $\phi_{I_s}$ in the quantum theory, and hence correlation functions of fluctuations in $n_s$ are determined by those of $\p_\tau \phi_{I_s}$).
-If $s=b+1$, then $\phi_{I_s}$ may develop LRO while leaving $\mmax{b}$ invariant.
+$\p_{J_{a-s+1}} \equiv \p_{j_1} \cdots \p_{j_{a-s+1}}$, and where $\phi_{I_s}$ is a field that shifts by a polynomial of degree $a-s$ under the action of $\mmax a$.\footnote{The compressibility appears as the coefficient of the time derivative term since $\p_\tau \phi_{I_s}$ is the momentum conjugate to $\phi_{I_s}$ in the quantum theory, and hence correlation functions of fluctuations in $n_s$ are determined by those of $\p_\tau \phi_{I_s}$.}  If $s=b+1$, then $\phi_{I_s}$ may develop LRO while leaving $\mmax{b}$ invariant.
 
-Note that the number of modes in the IR theory is controlled by $s$, and {\it not} by $b$.
 In order for $s< b+1$ to be consistent with our assumed pattern of symmetry breaking, it must be the case that the fluctuations of the fields in \eqref{generall} are strong enough so as to prevent $\mmax {s}$ from spontanoeusly breaking.
@@
 
-\subsection{Generalized Mermin-Wagner: partial multipole breaking}
+Our analysis here would naively imply that there are $(d+s-1 \text{ choose } s)$ massless modes, which is the number of degrees of freedom in a fully-symmetric rank-$s$ tensor.
+For the reasons discussed above, these modes may or may not be Goldstone modes, and other choices of symmetry-breaking fields may be possible.
+A detailed count of independent modes will require further analysis.
+
+\subsection{Generalized Mermin-Wagner: partial multipole breaking} \label{sub:partial}
 
@@
 
-In a disordered classical system, the energy scaling argument gives a nice explanation for why the critical dimension changes.
-Consider the formation of a domain of linear size $L$ in an otherwise ordered phase fully breaking the maximal multipole group $\mathcal{M}^a_\max$.
-There are $\sim L^d$ disorder samples in the new domain.
-Since each sample is taken independently, the typical energy gain from forming the domain will be $\sim L^{d/2}$, by the central limit theorem.
-Comparing this to the cost of domain formation we calculated previously, the ordered phase will be unstable to domain nucleation when $d\le 4(a+1)$.
-We emphasize that we are here considering disorder that couples linearly to the order parameter.
-
-In clean systems, the transition from classical to quantum brought down the critical dimension because the argument depended on entropy.
-The Imry-Ma argument only appeals to energy considerations, so the quantum critical dimension in the presence of disorder remains the same as the classical critical dimension~\cite{Vojta2013}.
-In disordered quantum systems, full SSB is impossible for $d\le 4(a+1)$.
-
-To reproduce this argument using a correlation function calculation~\cite{ImryMa} we need to calculate the response to disorder.
+In order to calculate the
+correlation function~\cite{ImryMa} we need to calculate the response to disorder.
 Call the disorder field $h(x)$ and redefine $\phi(x)$ to be the variation away from the average value $\phi_0$.
@@
 
-{\bf Acknowledgements} This work was supported by
-the U.S. Department of Energy, Office of Science, Basic
-Energy Sciences, under Award \# DE-SC0021346. E.L. was supported by a Hertz Fellowship.
+{\bf Acknowledgements} We thank Leo Radzihovsky for pointing out that Mermin-Wagner instability could mean flow to ordered but non-trivial fixed points. This work was supported by
+the U.S. Department of Energy, Office of Science, Basic Energy Sciences, under Award \# DE-SC0021346. E.L. was supported by a Hertz Fellowship.
 

### 2022-03-23 multipoleSSB/manuscript_DW.tex
domain walls vs no domain walls

--- 2022-03-16 multipoleSSB/manuscript.tex
@@
 This already appears in Ref.~\cite{Griffin2015}, so we are simply reviewing it here in preparation for the more generic cases to follow.
-Here, we will discuss both thermal and $T=0$ systems, and diagnose spontaneous symmetry breaking using the divergence of correlation functions.
+Here, we will discuss both thermal and $T=0$ systems.
+We will start with a heuristic argument in terms of domain wall energies before discussing a more careful diagnosis of spontaneous symmetry breaking in terms of correlation functions.
 
+Consider first a clean system at $T>0$ with a spontaneously broken maximal multipole symmetry of degree $a$.
+Then the kinetic term is proportional to 
+\begin{align}
+K[\phi(x)] = (\partial_{I_{a+1}} \phi)^2. \label{eqn:adisp}
+\end{align} 
+Symmetry breaking field configurations will be of the form $\phi(x) = a^0 + a^1_ix^i + \dots+ a^a_{I_a}x^{I_a}$, so there is no kinetic energy.
+
+Now consider nucleating domains in which the field takes on a new configuration $\phi(x) = c^0 + c^1_ix^i + \dots + c^a_{I_a}x^{I_a}$.
+There is a $d$-dimensional region surrounding the domain, where the coefficients in $\phi(x)$ smoothly change their values over a length $L$.
+We will call this the thickened domain wall.
+
+Let us specialize to $d=1$ to simplify the analysis of the thickened domain wall.
+In higher dimensions, for large enough thickened domain walls, the coefficients of $\phi(x)$ will locally appear to only vary in one direction, so that the analysis reduces to the $d=1$ problem.
+In the thickened domain wall, we have
+\begin{align}
+\phi(x) = b^0(y) + b^1(y) x + \dots + b^a(y) x^a,
+\end{align}
+where the $b^i(y)$ are polynomials of the dimensionless parameter $y=x/L$, whose degrees are constrained by the equations of motion coming from~\eqref{eqn:adisp}.
+We want to find thickened domain walls with the smallest possible energies.
+The total energy of a domain wall is minimized by taking $L$ large.
+In that limit, the smallest contribution to the energy comes from $b^0(y)$.
+Dimensional analysis then tells us the energy density will scale as $L^{-2(a+1)}$.
+
+When we go back to $d$ dimensions, the energy density remains unchanged. We can integrate the density over the entire thickened domain wall to see that the total energy is $E\sim L^{d-2(a+1)}$.
+In clean systems there is no energy gain from domain nucleation.
+When $d\le2(a+1)$, the energy cost is bounded, so the entropy gain favors domain creation.
+Thus, ordered phases are unstable and SSB cannot occur for $d\le2(a+1)$.
+
+We can reproduce this argument by considering correlation functions.
 As a warm-up, consider the standard case where a monopole U(1) symmetry is spontaneously broken, leading to a Goldstone boson $\phi(x)$. The correlation function
@@
 
-In the zero temperature setting, we calculate the correlation function by way of the (imaginary-time) IR Lagrangian 
+If we look at the zero temperature setting, the energy scaling no longer tells us where the critical dimension is.
+This is because even when there is no energy cost to forming domains, there is no entropy gain.
+Instead, quantum fluctuations must be the motivation for domain nucleation.
+To find the quantum critical dimension, we calculate the correlation function by way of the (imaginary-time) IR Lagrangian
 \begin{align}
@@
 If $d_1=2$ the symmetry can be broken for any $d_2>0$, and if $d_1=1$ then $d_2=2$  is critical. Of course, if $d_1=0$ we have an isotropic quartic dispersion and $d=d_2=4$ is critical.
+
+We can recover this result in  the energy-scaling argument.
+Since the system is anisotropic, the domains will have different sizes in different directions.
+Consider forming a domain of linear size $L_p$ in the  direction with quadratic dispersion and $L_k$ in the quartic direction.
+When the gradient of the order parameter coefficients is in the quadratic direction, the coefficients have to change over a length $L_p$ so the energy density is $L_p^{-2}$.
+Similarly, when the coefficients are changing in the quartic direction its energy density is $L_k^{-4}$.
+
+Both of these energy densities will be integrated over domain ``walls" with volume $L_k^{d_2} L_p^{d_1}$.
+As a result, the total energy cost is
+\begin{align}
+E &\sim L_k^{d_2} L_p^{d_1-2} + L_k^{d_2-4} L_p^{d_1},
+\end{align}
+so that $L_p \sim L_k^2$ to make the terms match, and $E\sim L_p^{d_2/2+d_1-2}$.
+Recall that when $E$ does not increase for larger domains, the entropy gain will cause them to nucleate, destroying the ordered phase.
+We see that the critical dimension $d= d_2 + d_1$ can be 2, 3, or 4, depending on how many directions have quadratic or quartic dispersions.
 
@@
 
-Our analysis here would naively imply that there are $(d+s-1 \text{ choose } s)$ massless modes, which is the number of degrees of freedom in a fully-symmetric rank-$s$ tensor.
-For the reasons discussed above, these modes may or may not be Goldstone modes, and other choices of symmetry-breaking fields may be possible.
-A detailed count of independent modes will require further analysis.
+The theories we consider here have $(d+s-1 \text{ choose } s)$ massless modes, which is the number of degrees of freedom in a fully-symmetric rank-$s$ tensor.
+For the reasons discussed above, these modes may or may not be Goldstone modes.
+In different effective field theories, there may be other choices of symmetry-breaking fields and thus different numbers of independent modes.
+Still, we expect that in general the number of independent modes will depend on the compressibilities, not necessarily on the pattern of spontaneous symmetry breaking.
 
@@
 
-In order to calculate the
-correlation function~\cite{ImryMa} we need to calculate the response to disorder.
+In a disordered classical system, the energy scaling argument gives a nice explanation for why the critical dimension changes.
+Consider the formation of a domain of linear size $L$ in an otherwise ordered phase fully breaking the maximal multipole group $\mathcal{M}^a_\max$.
+There are $\sim L^d$ disorder samples in the new domain.
+Since each sample is taken independently, the typical energy gain from forming the domain will be $\sim L^{d/2}$, by the central limit theorem.
+Comparing this to the cost of domain formation we calculated previously, the ordered phase will be unstable to domain nucleation when $d\le 4(a+1)$.
+We emphasize that we are here considering disorder that couples linearly to the order parameter.
+
+In clean systems, the transition from classical to quantum brought down the critical dimension because the argument depended on entropy.
+The Imry-Ma argument instead only appeals to energy considerations, so the quantum critical dimension in the presence of disorder remains the same as the classical critical dimension~\cite{Vojta2013}.
+In disordered quantum systems, full SSB is impossible for $d\le 4(a+1)$.
+
+To reproduce this argument using a correlation function calculation~\cite{ImryMa} we need to calculate the response to disorder.
 Call the disorder field $h(x)$ and redefine $\phi(x)$ to be the variation away from the average value $\phi_0$.

### 2022-03-23 multipoleSSB/manuscript.tex
accepted changes from Rahul and Ethan

@@
 \end{align}
-where $\gamma$ is an index that runs over polynomials of degree $c$ and $c\le a$.
+where $\gamma$ is an index that runs over polynomials of degree $c$, and $c\le a$.
 The composite index $I_c = \{i_1,\dots,i_c\}$  expands as $x^{I_c}=x^{i_1}\cdots x^{i_c}$, and similarly for $\partial_{I_c}$.
@@
 This already appears in Ref.~\cite{Griffin2015}, so we are simply reviewing it here in preparation for the more generic cases to follow.
-Here, we will discuss both thermal and $T=0$ systems, and diagnose spontaneous symmetry breaking using the divergence of correlation functions.
+Here, we will discuss both thermal and $T=0$ systems.
+We will start with a heuristic argument in terms of domain wall energies before discussing a more careful diagnosis of spontaneous symmetry breaking in terms of correlation functions.
 
+Consider first a clean system at $T>0$ with a spontaneously broken maximal multipole symmetry of degree $a$.
+Then the kinetic term is proportional to 
+\begin{align}
+K[\phi(x)] = (\partial_{I_{a+1}} \phi)^2. \label{eqn:adisp}
+\end{align} 
+Symmetry breaking field configurations will be of the form $\phi(x) = a^0 + a^1_ix^i + \dots+ a^a_{I_a}x^{I_a}$, where $a^b$ are symmetric rank-$b$ tensors. This ensures there is no kinetic energy.
+
+Now consider nucleating domains of linear size $L$, in which the field takes on a different configuration $\phi(x) = c^0 + c^1_ix^i + \dots + c^a_{I_a}x^{I_a}$.
+The difference in the coefficients $c^b_{I_b}-a^b_{I_b}$ will be of order $L^{-b}$, as that is the natural scale introduced by the presence of the domain.
+There is a $d$-dimensional region surrounding the domain, where the coefficients in $\phi(x)$ smoothly change their values over a length $L$. We will call this the thickened domain wall.
+
+Let us specialize to $d=1$ to simplify the analysis of the thickened domain wall.
+In higher dimensions, for large enough thickened domain walls, the coefficients of $\phi(x)$ will locally appear to only vary in one direction, so that the analysis reduces to the $d=1$ problem.
+In the thickened domain wall, we have
+\begin{align}
+\phi(x) = b^0(y) + b^1(y) y + \dots + b^a(y) y^a,
+\end{align}
+where the $b^b(y)$ are polynomials of the dimensionless parameter $y=x/L$, with degrees constrained by the equations of motion coming from~\eqref{eqn:adisp}.
+Each derivative in~\eqref{eqn:adisp} pulls down a factor of $L^{-1}$, so the energy density will scale as $L^{-2(a+1)}$.
+
+When we go back to $d$ dimensions, the energy density remains unchanged. We can integrate the density over the entire thickened domain wall to see that the total energy is $E\sim L^{d-2(a+1)}$.
+In clean systems there is no energy gain from domain nucleation.
+When $d\le2(a+1)$, the energy cost is bounded, so the entropy gain favors domain creation.
+Thus, ordered phases are unstable and SSB cannot occur for $d\le2(a+1)$.
+
+We can recover the result of this heuristic argument in a more formal way by considering correlation functions.
 As a warm-up, consider the standard case where a monopole U(1) symmetry is spontaneously broken, leading to a Goldstone boson $\phi(x)$. The correlation function
@@
 
-In the zero temperature setting, we calculate the correlation function by way of the (imaginary-time) IR Lagrangian 
+In the zero temperature setting, the energy scaling no longer tells us where the critical dimension is.
+This is because even when there is no energy cost to forming domains, there is no entropy gain.
+Instead, quantum fluctuations must be the motivation for domain nucleation.
+To find the quantum critical dimension, we calculate the correlation function by way of the (imaginary-time) IR Lagrangian
 \begin{align}
